---
title: "Modeling Aboveground Biomass of Individual Trees in Pinyon-Juniper Woodlands"
author: "Jake Howell"
date: "`r Sys.Date()`"
purpose: 
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
---


```{r}
# Set output formats
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Setup Libraries
```{r, include=FALSE}
# Check required packages are installed
pckgs <- c("sf",
           "dplyr",
           "sp",
           "raster",
           "terra",
           "lidR",
           "mapview",
           "units",
           "ggplot2",
           "broom",
           "purrr",
           "stringr",
           "tidyr",
           "readr",
           "leaflet")

# Point to packages
newPckgs <- pckgs[!(pckgs %in% installed.packages()[,"Package"])]

# If length is greater than 0, install the packages
if(length(newPckgs) > 0) install.packages(newPckgs)

# Load libraries
library(sf)
library(dplyr)
library(sp)
library(raster)
library(terra)
library(lidR)
library(mapview)
library(units)
library(ggplot2)
library(broom)
library(purrr)
library(stringr)
library(tidyr)
library(readr)
library(leaflet)
 

```

## Initialize Data Paths
```{r}
# Set root data location
loc <- r'(S:\ursa\howell\Research\PJ_forests)'

# Set data sub-roots
dataRaw <- file.path(loc, "data", "raw")
dataInt <- file.path(loc, "data", "intermediate")
dataOut  <- file.path(loc, "data", "output")

# Input paths
pathNAIPsites     <- file.path(dataRaw, "sites",  "01_NAIP_CHMs")
pathNAIPplots     <- file.path(dataRaw, "plots",  "naipCHMs")
pathLidarPlots    <- file.path(dataRaw, "plots",  "lidarCHMs")
pathCampbellPlots <- file.path(dataRaw, "plots", "campbellCHMs")
pathPlotData      <- file.path(dataRaw, "plots",  "field", "plot_level_data_uofu.csv")
pathTreeData      <- file.path(dataRaw, "trees",  "tree_level_data.csv")
pathAGBData       <- file.path(dataRaw, "trees",  "all_tree_species_in_pj_fia_plots_new.csv")
pathDEM           <- file.path(dataRaw, "dem_prism", "PRISM_us_dem_800m_bil.bil")
pathPrecip        <- file.path(dataRaw, "ppt_prism", "prism_ppt_us_30s_2020_avg_30y.tif")
pathPJBounds      <- file.path(dataRaw, "polygons",  "pj_polygon.shp")
pathSitePolys     <- file.path(dataRaw, "sites",     "00_SiteBounds", "site_polygons.shp")


```

## Read Data
```{r}
# List lidar-derived plot CHMs and read as a named list
lidarPlotCHM_names <- list.files(pathLidarPlots, pattern = "\\.tif$", full.names = TRUE)
lidarPlotRasters <- setNames(
  lapply(lidarPlotCHM_names, rast),
  tools::file_path_sans_ext(basename(lidarPlotCHM_names))
)

# List NAIP-derived plot CHMs and read as a named list
NAIPplotCHM_names <- list.files(pathNAIPplots, pattern = "\\.tif$", full.names = TRUE)
NAIPplotRasters <- setNames(
  lapply(NAIPplotCHM_names, rast),
  tools::file_path_sans_ext(basename(NAIPplotCHM_names))
)

# List Campbell-derived plot CHMs and read as a named list
CampbellPlotCHM_names <- list.files(pathCampbellPlots, pattern = "\\.tif$", full.names = TRUE)
CampbellPlotRasters <- setNames(
  lapply(CampbellPlotCHM_names, rast),
  tools::file_path_sans_ext(basename(CampbellPlotCHM_names))
)

# Read plot field data
plotData <- read.csv(pathPlotData)

# Read tree data
treeData <- read.csv(pathTreeData)

# Read biomass equations
biomassEquations <- read.csv(pathAGBData)

# Read data
dem <- rast(pathDEM)
ppt <- rast(pathPrecip)
pj <- st_read(pathPJBounds)
siteBounds <- st_read(pathSitePolys)

# Reproject
dem <- project(dem, "EPSG:32612")
ppt <- project(ppt, "EPSG:32612")
pj  <- st_transform(pj, "EPSG:32612")
siteBounds <- st_transform(siteBounds, "EPSG:32612")

```

## Data Pre-Processing 
```{r}
### Apply Biomass Equations to Field Data--------------------------------------
# In the tree data table, alter the species column
treeData <- treeData |>
  rename(Species = tree_sp)

# In the biomass equations table, alter the species column so it matches with the tree table
biomassEquations <- biomassEquations |>
  rename(Species = plants.code)

# Join the tree data to the biomass equations by Species column
treeData <- left_join(treeData, biomassEquations, by = "Species")

# Now estimate biomass for the field data trees by applying the biomass equations
treeData <- treeData |>
  mutate(tree_biomass_kg = exp(beta.0 + beta.1 * log(tree_diam)))

### Create Field Plot Extents---------------------------------------------------
# Create function to make one polygon the four sides of each field plot extent
createFieldPlotExtents <- function(row) {
  coords <- matrix(c(
    row$plot_ne_x, row$plot_ne_y,
    row$plot_se_x, row$plot_se_y,
    row$plot_sw_x, row$plot_sw_y,
    row$plot_nw_x, row$plot_nw_y,
    row$plot_ne_x, row$plot_ne_y
  ), ncol = 2, byrow = TRUE)
  st_polygon(list(coords))
}

# Go through each row of the plot data and create the field plot extents
fieldPlots_extents <- plotData |>
  rowwise() |>
  mutate(geometry = st_sfc(createFieldPlotExtents(pick(everything())))) |>
  ungroup()

# Set the current CRS and reproject
fieldPlots_extents <- st_as_sf(fieldPlots_extents, crs = 4326) |>
  st_transform(32612)

# Return just the plot ID's from the field plot data
plot_ids <- unique(fieldPlots_extents$plot_id)

### Sum the Plot-Level Biomass--------------------------------------------------
# Subset the tree data to just the unique plot names
treeDataInPlots_raw <- treeData[treeData$plot_id %in% plot_ids, ]

# FIX THE ERROR IN LAT/LONG ASSIGNMENT:
treeDataInPlots_fixed <- treeDataInPlots_raw %>%
  mutate(
    # start by assuming they are correct
    lon = tree_x,
    lat = tree_y,

    # identify likely swapped rows:
    # lon should be negative ~ -180..0 and lat should be positive ~ 0..90
    swapped = (tree_x > 0 & tree_x < 90) & (tree_y < 0 & tree_y > -180),

    # for swapped rows, flip them
    lon = if_else(swapped, tree_y, lon),
    lat = if_else(swapped, tree_x, lat)
  )

# Convert the treeDataInPlots to a spatial object
treeDataInPlots_fixed <- st_as_sf(treeDataInPlots_fixed,
                            coords = c("lon", "lat"), 
                            crs = 4326,
                            remove = FALSE)

# Now transform to UTM Zone 12N
treeDataInPlots <- st_transform(treeDataInPlots_fixed, 32612)

# Estimate Biomass Summary for each plot
plotLevelBiomass <- treeDataInPlots |>
  group_by(plot_id) |>
  summarise(
    total_biomass = sum(tree_biomass_kg, na.rm = TRUE),
    mean_biomass = mean(tree_biomass_kg, na.rm = TRUE),
    n_trees_GPS = n(),
    .groups = 'drop'
  )


```

## Input Parameters for Tree Detection and Modeling:
```{r}
# 1) Treetop Detection parameters-----------------------------------------------
# Set window size parameter
h <- function(x) {
  cw <- 1.0827 * x - 0.3993
  ws <- pmin(10, pmax(3, cw/2))
  return(ws)
}

# Set minimum tree size parameter
min_tree_ht_m <- 1

## 2) NAIP-derived CHM parameters-----------------------------------------------
# Segmentation parameters
th_tree_NAIP  = 0.5
th_seed_NAIP  = 0.3
th_cr_NAIP    = 0.6
max_cr_NAIP   = 15

# Smoothing parameters
smoothNAIP = TRUE

# Smoothing buffer distance
negBuffNAIP = -0.75
posBuffNAIP = 0.75

# Model form switches: below are the potential model forms
# 1 = Power Law          : a * H^b * A^c
# 2 = Additive           : a + b*H + c*A
# 3 = Additive non-linear: a * H^b + c * A^d  (fits 4 parameters)
# 4 = Height only        : a * H^b
# 5 = Area only          : a * A^b
naip_model_form = 1
 
# 3) Lidar-derived CHM parameters-----------------------------------------------
# Segmentation parameters
th_tree_lidar  = 0.5
th_seed_lidar  = 0.25
th_cr_lidar    = 0.3
max_cr_lidar   = 15

# Smoothing parameters
smoothLidar = TRUE

# Smoothing buffer distance
negBuff = -0.25
posBuff = 0.25

# 4) Campbell-derived CHM parameters--------------------------------------------
# Segmentation parameters
th_tree_Campbell = 0.5
th_seed_Campbell = 0.3
th_cr_Campbell   = 0.6
max_cr_Campbell  = 15

# Smoothing parameters
smoothCampbell = TRUE

# Smoothing buffer distance
negBuff = -0.75
posBuff = 0.75

```


### RUN NAIP-CHM TREE DETECTION AND CREATE MODELS
```{r}
NAIP_treePoints_list <- list()
NAIP_treeCrowns_list <- list()

for (plot_id in plot_ids) {

  chm <- NAIPplotRasters[[plot_id]]
  if (is.null(chm)) next

  chm <- terra::project(chm, "EPSG:32612")

  # Detect tree tops
  ttops_sf <- locate_trees(chm, lmf(ws = h, hmin = min_tree_ht_m)) |> st_as_sf()
  if (nrow(ttops_sf) == 0) next

  # Segment crowns
  crowns_poly <- dalponte2016(
    chm      = chm,
    treetops = ttops_sf,
    th_tree  = th_tree_NAIP,
    th_seed  = th_seed_NAIP,
    th_cr    = th_cr_NAIP,
    max_cr   = max_cr_NAIP
  )() |>
    terra::as.polygons(dissolve = TRUE, na.rm = TRUE)

  if (terra::nrow(crowns_poly) == 0) next

  # Extract crown metrics
  names(crowns_poly)        <- "treeID"
  crowns_poly$area_m2       <- terra::expanse(crowns_poly, unit = "m")
  crowns_poly$tree_height_m <- terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE)[, 2]

  # Convert to sf and tag with plot ID
  crowns_sf        <- sf::st_as_sf(crowns_poly)
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID  <- plot_id

  NAIP_treePoints_list[[plot_id]] <- ttops_sf
  NAIP_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Combine, reproject, and clip to field plot extents
allTreePoints_NAIP <- dplyr::bind_rows(NAIP_treePoints_list) |>
  st_transform(st_crs(fieldPlots_extents))

allTreeCrowns_NAIP <- dplyr::bind_rows(NAIP_treeCrowns_list) |>
  st_transform(st_crs(fieldPlots_extents))

# Keep only tree tops and crowns within field plot extents
keys <- allTreePoints_NAIP[
  lengths(st_within(allTreePoints_NAIP, fieldPlots_extents)) > 0, ] |>
  st_drop_geometry() |>
  distinct(plotID, treeID)

allTreeCrowns_NAIP_final <- allTreeCrowns_NAIP |>
  inner_join(keys, by = c("plotID", "treeID"))

allTreePoints_NAIP_final <- allTreePoints_NAIP |>
  inner_join(keys, by = c("plotID", "treeID"))
```

### NAIP: Compare Model Forms Performances
```{r}
# Preserve the model form selected in the input parameters
selected_naip_model_form <- naip_model_form

all_forms <- 1:5

model_results <- lapply(all_forms, function(form) {

  # Set model form used by objective() and predict_plot_biomass()
  naip_model_form <<- form

  sv <- switch(
    as.character(form),
    `1` = c(a = 0.1, b = 1, c = 1),
    `2` = c(a = 0.1, b = 1, c = 1),
    `3` = c(a = 0.1, b = 1, c = 0.1, d = 1),
    `4` = c(a = 0.1, b = 1),
    `5` = c(a = 0.1, b = 1)
  )

  lb <- switch(
    as.character(form),
    `1` = c(a = 1e-6, b = -5, c = -5),
    `2` = c(a = -Inf, b = -Inf, c = -Inf),
    `3` = c(a = 1e-6, b = -5, c = 1e-6, d = -5),
    `4` = c(a = 1e-6, b = -5),
    `5` = c(a = 1e-6, b = -5)
  )

  ub <- switch(
    as.character(form),
    `1` = c(a = 100, b = 5, c = 5),
    `2` = c(a = Inf, b = Inf, c = Inf),
    `3` = c(a = 100, b = 5, c = 100, d = 5),
    `4` = c(a = 100, b = 5),
    `5` = c(a = 100, b = 5)
  )

  fit_f <- tryCatch(
    optim(
      par    = sv,
      fn     = objective,
      method = "L-BFGS-B",
      lower  = lb,
      upper  = ub
    ),
    error = function(e) {
      warning(sprintf("Model form %s failed: %s", form, e$message))
      NULL
    }
  )

  if (is.null(fit_f)) {
    return(NULL)
  }

  pred_f <- predict_plot_biomass(fit_f$par)
  obs_f  <- plot_biomass[names(pred_f)]

  valid <- is.finite(obs_f) & is.finite(pred_f)
  obs_f <- obs_f[valid]
  pred_f <- pred_f[valid]

  SSE  <- sum((obs_f - pred_f)^2)
  SST  <- sum((obs_f - mean(obs_f))^2)
  RMSE <- sqrt(mean((obs_f - pred_f)^2))

  list(
    form = form,
    fit = fit_f,
    coefficients = fit_f$par,
    metrics = data.frame(
      model_form = form,
      equation = c(
        "a*H^b*A^c",
        "a+b*H+c*A",
        "a*H^b+c*A^d",
        "a*H^b",
        "a*A^b"
      )[form],
      n_params  = length(fit_f$par),
      R2        = 1 - SSE / SST,
      RMSE      = RMSE,
      rRMSE_pct = RMSE / mean(obs_f) * 100,
      MAPE_pct  = mean(abs((obs_f - pred_f) / obs_f)) * 100,
      convergence = fit_f$convergence
    )
  )
})

# Remove failed model fits
model_results <- Filter(Negate(is.null), model_results)

# Combine metrics
model_comparison <- dplyr::bind_rows(
  lapply(model_results, `[[`, "metrics")
) |>
  mutate(
    R2        = round(R2, 3),
    RMSE      = round(RMSE, 1),
    rRMSE_pct = round(rRMSE_pct, 1),
    MAPE_pct  = round(MAPE_pct, 1)
  )

model_comparison

# Restore the model form originally selected in your input parameters
naip_model_form <- selected_naip_model_form

selected_result <- model_results[
  vapply(
    model_results,
    function(x) x$form == selected_naip_model_form,
    logical(1)
  )
][[1]]

NAIPFit <- selected_result$fit
NAIPCoefficients <- selected_result$coefficients

print(naip_model_form)
print(NAIPCoefficients)
```

### Apply Results to the Tree List
```{r}
detectedTrees_NAIP <- allTreeCrowns_NAIP_smoothed |>
  mutate(
    detected_AGB_kg = switch(
      as.character(naip_model_form),

      `1` = NAIPCoefficients[["a"]] *
        tree_height_m^NAIPCoefficients[["b"]] *
        area_m2^NAIPCoefficients[["c"]],

      `2` = NAIPCoefficients[["a"]] +
        NAIPCoefficients[["b"]] * tree_height_m +
        NAIPCoefficients[["c"]] * area_m2,

      `3` = NAIPCoefficients[["a"]] *
        tree_height_m^NAIPCoefficients[["b"]] +
        NAIPCoefficients[["c"]] *
        area_m2^NAIPCoefficients[["d"]],

      `4` = NAIPCoefficients[["a"]] *
        tree_height_m^NAIPCoefficients[["b"]],

      `5` = NAIPCoefficients[["a"]] *
        area_m2^NAIPCoefficients[["b"]],

      stop("Unsupported NAIP model form.")
    )
  )
```

### Review results of NAIP-based tree detection
```{r}
# NAIP-derived tree detection result
print(NAIP_TreeDetection_ModResults)

# What are my individual tree level coefficients?
print(NAIPCoefficients)
NAIPCoefficients_df <- as.data.frame(NAIPCoefficients)

# What's the name of the CHM you want to check?
chm_name <- "UOFU_UTCF_P04"

# Get the single NAIP CHM
chm_NAIP <- NAIPplotRasters[[chm_name]]

# Subset the NAIP tree points to selected plot
treePoints_NAIP <- allTreePoints_NAIP_final |> 
  filter(plotID == chm_name)

# Convert to simple feature
treePoints_NAIP <- st_as_sf(treePoints_NAIP)

# Drop 0 values
treePoints_NAIP <- st_zm(treePoints_NAIP, drop = TRUE, what = "ZM")

# Subset the NAIP tree crowns — swap source to compare raw vs smoothed
treeCrowns_NAIP <- allTreeCrowns_NAIP_final |>   
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Subset the NAIP tree crowns to smoothed
treeCrowns_NAIP_smoothed <- allTreeCrowns_NAIP_smoothed |>  
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Get the GPS tree points for the plot
fieldTrees <- treeDataInPlots %>%
  filter(plot_id == chm_name)

# Get the field plot extent 
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

# Join detected AGB to crowns and tree tops for hover info
crowns_agb <- treeCrowns_NAIP_smoothed |>
  left_join(
    detectedTrees_NAIP |> st_drop_geometry() |> dplyr::select(treeID, plotID, detected_AGB_kg),
    by = c("treeID", "plotID")
  )

ttops_agb <- treePoints_NAIP |>
  st_zm(drop = TRUE, what = "ZM") |>
  left_join(
    detectedTrees_NAIP |> st_drop_geometry() |> dplyr::select(treeID, plotID, detected_AGB_kg),
    by = c("treeID", "plotID")
  )

# Observed GPS trees with biomass — filter zero AGB
obs_agb <- fieldTrees |>
  filter(tree_biomass_kg > 0) |>
  dplyr::select(plot_id, tree_id, tree_biomass_kg) |>
  st_zm(drop = TRUE, what = "ZM")

# Map: CHM + smoothed crowns + tree tops + GPS trees, all coloured and labelled by AGB
mapview(chm_NAIP,
        layer.name  = "NAIP CHM",
        col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(crowns_agb,
          zcol       = "detected_AGB_kg",
          layer.name = "Detected Crowns (AGB)") +
  mapview(ttops_agb,
          zcol       = "detected_AGB_kg",
          layer.name = "Tree Tops (Detected AGB)",
          cex        = 4) +
  mapview(obs_agb,
          zcol       = "tree_biomass_kg",
          layer.name = "Observed Trees (Field AGB)",
          cex        = 5) +
  mapview(plot_extent,
          color      = "black",
          lwd        = 3,
          alpha.regions = 0,
          layer.name = "Field Plot Extent")

```

### Summarize Plot-Level Counts and AGB
```{r}
# Detected tree counts and total AGB per plot
detected_plot_summary <- detectedTrees_NAIP |>
  st_drop_geometry() |>
  group_by(plotID) |>
  summarise(
    n_detected         = n(),
    detected_total_AGB = sum(detected_AGB_kg, na.rm = TRUE),
    .groups = "drop"
  )

# Observed tree counts and total AGB per plot
observed_plot_summary <- plotLevelBiomass |>
  st_drop_geometry() |>
  dplyr::select(plot_id, n_observed = n_trees_GPS, observed_total_AGB = total_biomass)

# Join summaries and compute count ratio
plot_comparison <- detected_plot_summary |>
  inner_join(observed_plot_summary, by = c("plotID" = "plot_id")) |>
  mutate(count_ratio = n_detected / n_observed)
```

### Filter to Low-AGB Plots with Similar Tree Counts
```{r}
# Thresholds - adjust as needed
low_AGB_thresh     <- quantile(plot_comparison$observed_total_AGB, 0.25)
count_ratio_margin <- 0.25  # detected within 25% of observed count

low_AGB_plots <- plot_comparison |>
  filter(
    observed_total_AGB <= low_AGB_thresh,
    count_ratio        >= (1 - count_ratio_margin),
    count_ratio        <= (1 + count_ratio_margin)
  )

message("Plots meeting criteria: ", nrow(low_AGB_plots))
low_AGB_plots
```

### Join Detected Crowns to Observed Trees
```{r}
# AGB relative error tolerance for accepting a match
AGB_margin <- 150

# Subset to target plots
detected_subset <- detectedTrees_NAIP |>
  filter(plotID %in% low_AGB_plots$plotID)

observed_subset <- treeDataInPlots |>
  filter(plot_id %in% low_AGB_plots$plotID) |>
  st_transform(st_crs(detected_subset))

# Spatial join: observed tree point falls within detected crown polygon
joined_trees <- st_join(observed_subset, detected_subset, join = st_within, left = FALSE) |>
  st_drop_geometry() |>
  mutate(AGB_rel_err = abs(detected_AGB_kg - tree_biomass_kg) / tree_biomass_kg) |>
  filter(AGB_rel_err <= AGB_margin)

message("Matched trees: ", nrow(joined_trees), " of ", nrow(observed_subset), " observed")
```

### Plot Observed vs. Detected AGB
```{r}
ggplot(joined_trees, aes(x = tree_biomass_kg, y = detected_AGB_kg)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick") +
  labs(
    x        = "Observed Tree AGB (kg)",
    y        = "Detected Tree AGB (kg)",
    title    = "Observed vs. Detected Tree AGB",
    subtitle = "Low-AGB plots where n_detected is close to n_observed"
  )
```


### RUN LIDAR-CHM TREE DETECTION
```{r}
# Save the names of the Lidar plots
Lidar_PlotNames <- names(lidarPlotRasters)

# Store tree tops and crowns as a list
Lidar_treePoints_list <- list()
Lidar_treeCrowns_list <- list()

# Loop through each plot CHM and run tree detection
for (plot_id in Lidar_PlotNames) {

  message("Processing plot: ", plot_id)
  
   # Select a single plot CHM
    chm <- lidarPlotRasters[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    # Ensure proper projection
    chm <- terra::project(chm, "EPSG:32612")
    
    # Identify Tree Tops
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree_lidar,
        th_seed  = th_seed_lidar,
        th_cr    = th_cr_lidar,
        max_cr   = max_cr_lidar
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    # Store names as treeID
    names(crowns_poly) <- "treeID"  

  # Drop empty / invalid geometries
  crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

  if (nrow(crowns_poly) == 0) next

  # Calculate Crown Area in m2
  crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

  # Extract the maximum height for each crown polygon
  zmax <- tryCatch(
    terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
    error = function(e) {
      message("extract failed: ", e$message)
      NULL
    }
  )

  if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
    message("Height extraction row mismatch for plot: ", plot_id,
            " | crowns=", terra::nrow(crowns_poly),
            " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax)))
    next
  }

  # Assign the maximum height to the crown polygons
  crowns_poly$tree_height_m <- zmax[,2]

  # Convert to sf 
  crowns_sf <- sf::st_as_sf(crowns_poly)

  # Add plot ID
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID <- plot_id

  # Store
  Lidar_treePoints_list[[plot_id]] <- ttops_sf
  Lidar_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Bind all rows
allTreePoints_Lidar <- dplyr::bind_rows(Lidar_treePoints_list)
allTreeCrowns_Lidar <- dplyr::bind_rows(Lidar_treeCrowns_list)

# Check projection
allTreePoints_Lidar <- st_transform(allTreePoints_Lidar, st_crs(fieldPlots_extents))
allTreeCrowns_Lidar <- st_transform(allTreeCrowns_Lidar, st_crs(fieldPlots_extents))

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- allTreePoints_Lidar[
  lengths(st_within(allTreePoints_Lidar, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
allTreeCrowns_Lidar_final <- allTreeCrowns_Lidar %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Subset to only the tree tops we want to keep
allTreePoints_Lidar_final <- allTreePoints_Lidar %>%
  inner_join(keys, by = c("plotID", "treeID"))
```

### Apply Crown Smoothing (Lidar)
```{r}
# Get CHM resolution (consistent across all Lidar plots)
lidar_res <- res(lidarPlotRasters[[1]])[1]

# Set small crown threshold (3 pixels) — small crowns are left unsmoothed
small_thresh <- 3 * lidar_res^2

# Split into small and large crowns by pre-smoothing area
small_crowns_sf <- allTreeCrowns_Lidar_final[allTreeCrowns_Lidar_final$area_m2 <= small_thresh, ]
big_crowns_sf   <- allTreeCrowns_Lidar_final[allTreeCrowns_Lidar_final$area_m2 > small_thresh, ]

# Apply buffer-in / buffer-out smoothing only to large crowns
big_crowns_sf <- big_crowns_sf |>
  st_buffer(-0.25 * lidar_res) |>
  st_buffer(0.25 * lidar_res) |>
  st_simplify(preserveTopology = TRUE, dTolerance = lidar_res / 4)

# Recombine and drop any geometries that collapsed during smoothing
allTreeCrowns_Lidar_smoothed <- bind_rows(big_crowns_sf, small_crowns_sf) |>
  filter(!st_is_empty(geometry))

# Cast to single-part polygons so split crowns become separate rows,
# then keep only the largest fragment per crown
allTreeCrowns_Lidar_smoothed <- allTreeCrowns_Lidar_smoothed |>
  st_cast("POLYGON") |>
  mutate(area_m2 = as.numeric(st_area(geometry))) |>
  group_by(plotID, treeID) |>
  slice_max(area_m2, n = 1, with_ties = FALSE) |>
  ungroup()

cat("Crowns before smoothing:", nrow(allTreeCrowns_Lidar_final), "\n")
cat("Crowns after smoothing: ", nrow(allTreeCrowns_Lidar_smoothed), "\n")

```

### Review results of Lidar-based tree detection
```{r}
# What's the name of the CHM you want to check?
chm_name <- "UOFU_NMQU_P09"

# Get the single Lidar CHM
chm_Lidar <- lidarPlotRasters[[chm_name]]

# Subset the Lidar tree points to selected plot
treePoints_Lidar <- allTreePoints_Lidar_final |> 
  filter(plotID == chm_name)

# Convert to simple feature
treePoints_Lidar <- st_as_sf(treePoints_Lidar)

# Drop 0 values
treePoints_Lidar <- st_zm(treePoints_Lidar, drop = TRUE, what = "ZM")

# Subset the Lidar tree crowns — swap source to compare raw vs smoothed
treeCrowns_Lidar <- allTreeCrowns_Lidar_final |>
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Subset the Lidar tree crowns to smoothed
treeCrowns_Lidar_smoothed <- allTreeCrowns_Lidar_smoothed |>
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Get the GPS tree points for the plot
fieldTrees <- treeDataInPlots %>%
  filter(plot_id == chm_name)

# Get the field plot extent 
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

# Plot it
mapview(
    chm_Lidar,
    layer.name = "Lidar CHM",
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(treePoints_Lidar, 
          color = "blue", 
          layer.name = "Lidar Tree Tops") +
  mapview(treeCrowns_Lidar, 
          color = "red", 
          lwd = 3,
          layer.name = "Lidar Crowns") +
  mapview(plot_extent, 
          color = "black", 
          lwd = 3, 
          fill = NA, 
          layer.name = "Field Plot Extents")

# Plot the smoothed crowns
mapview(
    chm_Lidar,
    layer.name = "Lidar CHM",
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(treePoints_Lidar,
          color = "blue",
          layer.name = "Lidar Tree Tops") +
  mapview(treeCrowns_Lidar_smoothed,
          color = "red",
          lwd = 3,
          layer.name = "Lidar Crowns (Smoothed)") +
  mapview(plot_extent,
          color = "black",
          lwd = 3,
          fill = NA,
          layer.name = "Field Plot Extents")

```

### RUN CAMPBELL NAIP-CHM TREE DETECTION
```{r}
# Save the names of the Campbell plots
Campbell_PlotNames <- names(CampbellPlotRasters)

# Store tree tops and crowns as a list
Campbell_treePoints_list <- list()
Campbell_treeCrowns_list <- list()

# Loop through each plot CHM and run tree detection
for (plot_id in plot_ids) {

  message("Processing plot: ", plot_id)

  # Select a single plot CHM
  chm <- CampbellPlotRasters[[plot_id]]

  if (is.null(chm)) {
    message("CHM is NULL for plot: ", plot_id)
    next
  }

  # Ensure proper projection
  chm <- terra::project(chm, "EPSG:32612")

  # Identify Tree Tops
  ttops <- tryCatch(
    locate_trees(chm, lmf(ws = h, hmin = 0.5)),
    error = function(e) {
      message("Treetops failed: ", e$message)
      NULL
    }
  )

  # Skip if no treetops
  if (is.null(ttops) ||
      (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
      (inherits(ttops, "sf") && nrow(ttops) == 0) ||
      (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
    message("No treetops detected for plot: ", plot_id)
    next
  }

  # Convert to sf object
  ttops_sf <- tryCatch(
    st_as_sf(ttops),
    error = function(e) {
      message("st_as_sf failed: ", e$message)
      NULL
    }
  )

  if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
    message("No valid treetop sf object for plot: ", plot_id)
    next
  }

  # Run Dalponte2016 tree segmentation algorithm
  crowns_id <- tryCatch(
    dalponte2016(
      chm      = chm,
      treetops = ttops_sf,
      th_tree  = th_tree_Campbell,
      th_seed  = th_seed_Campbell,
      th_cr    = th_cr_Campbell,
      max_cr   = max_cr_Campbell
    )(),
    error = function(e) {
      message("Dalponte CHM-only failed: ", e$message)
      NULL
    }
  )

  if (is.null(crowns_id)) {
    message("No crown raster returned for plot: ", plot_id)
    next
  }

  # Convert crown segmentation raster to polygons
  crowns_poly <- tryCatch(
    terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
    error = function(e) {
      message("as.polygons failed: ", e$message)
      NULL
    }
  )

  if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
    message("No crown polygons for plot: ", plot_id)
    next
  }

  # Store names as treeID
  names(crowns_poly) <- "treeID"

  # Calculate initial crown area for small/large classification
  crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

  # Extract the maximum height for each crown polygon
  zmax <- tryCatch(
    terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
    error = function(e) {
      message("extract failed: ", e$message)
      NULL
    }
  )

  if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
    message("Height extraction row mismatch for plot: ", plot_id,
            " | crowns=", terra::nrow(crowns_poly),
            " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax)))
    next
  }

  # Assign the maximum height to the crown polygons
  crowns_poly$tree_height_m <- zmax[, 2]

  # Convert to sf
  crowns_sf <- sf::st_as_sf(crowns_poly)

  # Add plot ID
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID  <- plot_id

  # Store
  Campbell_treePoints_list[[plot_id]] <- ttops_sf
  Campbell_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Bind all rows
allTreePoints_Campbell <- dplyr::bind_rows(Campbell_treePoints_list)
allTreeCrowns_Campbell <- dplyr::bind_rows(Campbell_treeCrowns_list)

# Check projection
allTreePoints_Campbell <- st_transform(allTreePoints_Campbell, st_crs(fieldPlots_extents))
allTreeCrowns_Campbell <- st_transform(allTreeCrowns_Campbell, st_crs(fieldPlots_extents))

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- allTreePoints_Campbell[
  lengths(st_within(allTreePoints_Campbell, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
allTreeCrowns_Campbell_final <- allTreeCrowns_Campbell %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Subset to only the tree tops we want to keep
allTreePoints_Campbell_final <- allTreePoints_Campbell %>%
  inner_join(keys, by = c("plotID", "treeID"))

```

### Review results of Campbell NAIP-CHM Tree Detection
```{r}
# What's the name of the CHM you want to check?
chm_name <- "UOFU_NMQU_P09"

# Get the single Campbell CHM
chm_Campbell <- CampbellPlotRasters[[chm_name]]

# Subset the Campbell tree points to selected plot
treePoints_Campbell <- allTreePoints_Campbell_final |>
  filter(plotID == chm_name)

# Convert to simple feature
treePoints_Campbell <- st_as_sf(treePoints_Campbell)

# Drop Z/M values
treePoints_Campbell <- st_zm(treePoints_Campbell, drop = TRUE, what = "ZM")

# Subset the Campbell tree crowns
treeCrowns_Campbell <- allTreeCrowns_Campbell_final |>
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Get the GPS tree points for the plot
fieldTrees <- treeDataInPlots %>%
  filter(plot_id == chm_name)

# Get the field plot extent
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

# Plot it
mapview(
    chm_Campbell,
    layer.name = "Campbell CHM",
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(treePoints_Campbell,
          color = "blue",
          layer.name = "Campbell Tree Tops") +
  mapview(treeCrowns_Campbell,
          color = "red",
          lwd = 3,
          layer.name = "Campbell Crowns") +
  mapview(plot_extent,
          color = "black",
          lwd = 3,
          fill = NA,
          layer.name = "Field Plot Extents")

```

### Create Models for the NAIP-derived tree list
```{r}
# Create tree and plot df
model_trees_NAIP <- allTreeCrowns_NAIP_smoothed
model_plots_NAIP <- plotLevelBiomass

# Select tree variables
model_trees_NAIP <- model_trees_NAIP[,c("plotID", "tree_height_m", "area_m2")]
model_trees_NAIP <- st_drop_geometry(model_trees_NAIP)

# Select plot variables
model_plots_NAIP <- model_plots_NAIP[,c("plot_id", "total_biomass")]
model_plots_NAIP <- st_drop_geometry(model_plots_NAIP)

# Subset data to these variables
colnames(model_trees_NAIP) <- c("plot_id", "tree_height", "tree_area")
colnames(model_plots_NAIP) <- c("plot_id", "plot_biomass")

# Merge the data
model_df <- merge(model_plots_NAIP, model_trees_NAIP)

# Split into list by plot
plot_ids <- unique(model_df$plot_id)
tree_list <- split(model_df, model_df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(model_df$plot_biomass, model_df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  # ensure ordering matches
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1, 
                b = 1, 
                c = 1
)

# Fit the parameters
fit <- optim(
  par = 
    start_vals,
  fn = objective,
  
  # Broyden, Fletcher, Goldfarb, and Shanno
  # Uses function values and gradients to build up a picture of the surface to be optimized
  method = "L-BFGS-B",
  lower = c(1e-6, 
            -5,
            -5),
  upper = c(100,  
            5,
            5))

# Whats here?
# The best set of parameters found
# The coefficients for individual tree level biomass! 
NAIPCoefficients <- fit$par 

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_NAIP_estimates <- left_join(predicted_df, observed_df, by = "PlotID")

# Observed and predicted vectors
obs <- plot_biomass
pred <- predicted[names(plot_biomass)]   

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2 <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
NAIP_TreeDetection_ModResults <- data.frame(
  RMSE = RMSE,
  MAE = MAE,
  Bias = Bias,
  MAPE_percent = MAPE,
  R2 = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result 
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)

# Final NAIP estimates dataframe
final_NAIP_estimates <- final_NAIP_estimates |>
  dplyr::select(PlotID,
                predicted,
                plot_biomass) |>
  rename("observed" = plot_biomass)

```

### Create models for lidar-derived tree list 
```{r}
# Create tree and plot df
model_trees_lidar <- allTreeCrowns_Lidar_smoothed
model_plots_lidar <- plotLevelBiomass

# Select tree variables
model_trees_lidar <- model_trees_lidar[,c("plotID", "tree_height_m", "area_m2")]
model_trees_lidar <- st_drop_geometry(model_trees_lidar)

# Select plot variables
model_plots_lidar <- model_plots_lidar[,c("plot_id", "total_biomass")]
model_plots_lidar <- st_drop_geometry(model_plots_lidar)

# Subset data to these variables
colnames(model_trees_lidar) <- c("plot_id", "tree_height", "tree_area")
colnames(model_plots_lidar) <- c("plot_id", "plot_biomass")

# Merge the data
df <- merge(model_plots_lidar, model_trees_lidar)

# Split into list by plot
plot_ids <- unique(df$plot_id)
tree_list <- split(df, df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(df$plot_biomass, df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  # ensure ordering matches
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1, 
                b = 1, 
                c = 1
)

# Fit the parameters
fit <- optim(
  par = 
    start_vals,
  fn = objective,
  
  # Broyden, Fletcher, Goldfarb, and Shanno
  # Uses function values and gradients to build up a picture of the surface to be optimized
  method = "L-BFGS-B",
  lower = c(1e-6, 
            -5,
            -5),
  upper = c(100,  
            5,
            5))

# Whats here?
# The best set of parameters found
# The coefficients for individual tree level biomass! 
lidarCoefficients <- fit$par 

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_Lidar_estimates <- left_join(predicted_df, observed_df, by = "PlotID")

# Observed and predicted vectors
obs <- plot_biomass
pred <- predicted[names(plot_biomass)]   

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2 <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
Lidar_TreeDetection_ModResults <- data.frame(
  RMSE = RMSE,
  MAE = MAE,
  Bias = Bias,
  MAPE_percent = MAPE,
  R2 = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result 
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)

# Edit final lidar estimates table
final_Lidar_estimates <- final_Lidar_estimates |>
  dplyr::select(PlotID,
         predicted,
         plot_biomass) |>
  rename("observed" = plot_biomass)

```

### Create models for Campbell NAIP tree list
```{r}
# Create tree and plot df
model_trees_Campbell <- allTreeCrowns_Campbell_final
model_plots_Campbell <- plotLevelBiomass

# Select tree variables
model_trees_Campbell <- model_trees_Campbell[,c("plotID", "tree_height_m", "area_m2")]
model_trees_Campbell <- st_drop_geometry(model_trees_Campbell)

# Select plot variables
model_plots_Campbell <- model_plots_Campbell[,c("plot_id", "total_biomass")]
model_plots_Campbell <- st_drop_geometry(model_plots_Campbell)

# Rename to common column names
colnames(model_trees_Campbell) <- c("plot_id", "tree_height", "tree_area")
colnames(model_plots_Campbell) <- c("plot_id", "plot_biomass")

# Merge the data
model_df <- merge(model_plots_Campbell, model_trees_Campbell)

# Split into list by plot
plot_ids <- unique(model_df$plot_id)
tree_list <- split(model_df, model_df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(model_df$plot_biomass, model_df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1,
                b = 1,
                c = 1)

# Fit the parameters
fit <- optim(
  par    = start_vals,
  fn     = objective,
  method = "L-BFGS-B",
  lower  = c(1e-6, -5, -5),
  upper  = c(100,   5,  5))

# The best set of parameters found
# The coefficients for individual tree level biomass!
CampbellCoefficients <- fit$par

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_Campbell_estimates <- left_join(predicted_df, observed_df, by = "PlotID")

# Observed and predicted vectors
obs  <- plot_biomass
pred <- predicted[names(plot_biomass)]

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2  <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
Campbell_TreeDetection_ModResults <- data.frame(
  RMSE          = RMSE,
  MAE           = MAE,
  Bias          = Bias,
  MAPE_percent  = MAPE,
  R2            = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass (Campbell NAIP)")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)

# Final Campbell estimates dataframe
final_Campbell_estimates <- final_Campbell_estimates |>
  dplyr::select(PlotID,
                predicted,
                plot_biomass) |>
  rename("observed" = plot_biomass)

```

### What are my modeling results? 
```{r}
# NAIP-derived tree detection result
print(NAIP_TreeDetection_ModResults)

# What are my individual tree level coefficients?
print(NAIPCoefficients)
NAIPCoefficients_df <- as.data.frame(NAIPCoefficients)

# Lidar-derived tree detection result
print(Lidar_TreeDetection_ModResults)

# What are my individual tree level coefficients
print(lidarCoefficients)
lidarCoefficients_df <- as.data.frame(lidarCoefficients)

```


### Apply model coefficients to the NAIP tree list
```{r}
# Apply model coefficients to the NAIP-derived smoothed tree list
NAIPTreeList_AGB <- allTreeCrowns_NAIP_smoothed
NAIPTreeList_AGB$beta.a <- 3.4553996 
NAIPTreeList_AGB$beta.b <- 0.8206139 
NAIPTreeList_AGB$beta.c <- 0.9328608 

# Calculate tree-level AGB using coefficients
NAIPTreeList_AGB <- NAIPTreeList_AGB |>
  mutate(tree_biomass = beta.a * (tree_height_m^beta.b) * (area_m2^beta.c))

# Calculate plot-level AGB by summing tree-level biomass
NAIPPlots_AGB <- NAIPTreeList_AGB |>
  group_by(plotID) |>
  summarise(predicted_AGB = sum(tree_biomass, na.rm = TRUE)) |>
  rename(PlotID = "plotID")


```

### Apply model coefficients to the Lidar tree list
```{r}
LidarTreeList_AGB <- allTreeCrowns_Lidar_final
LidarTreeList_AGB$beta.a <- 4.6192000 
LidarTreeList_AGB$beta.b <- 1.1009588 
LidarTreeList_AGB$beta.c <- 0.6555886 

# Calculate tree-level AGB using coefficients
LidarTreeList_AGB <- LidarTreeList_AGB |>
  mutate(tree_biomass = beta.a * (tree_height_m^beta.b) * (area_m2^beta.c))

```

### Create landscape stratification
```{r}
# Create a buffer around PJ ecosystem
pj_buffer <- st_buffer(pj, dist = 1000)  

# Merge into a single spatial layer
pj_buffer <- st_union(pj_buffer)

# Clip & mask my rasters to the PJ extent
dem_pj <- terra::mask(terra::crop(dem, pj_buffer), pj_buffer)
ppt_pj <- terra::mask(terra::crop(ppt, pj_buffer), pj_buffer)

# Create a landscape stratification
names(dem_pj) <- "elevation"
names(ppt_pj) <- "precipitation"

# Create a composite raster
compRast <- c(dem_pj, ppt_pj)

# Create a landscape stratification
stratRast <- sgsR::strat_quantiles(
  mraster = compRast[[1:2]],
  nStrat = list(3, 3),
  plot = FALSE,
  map = TRUE
)

# Select the stratified raster
pj_strat <- stratRast[[3]]

# Extract the strata class for the site extents
siteBounds_strata <- siteBounds
site_strata_vals <- terra::extract(pj_strat, siteBounds_strata, fun = modal, na.rm = TRUE)
siteBounds_strata$site_strata <- round(strata_vals[,2])  # Extract the mean strata value and round to nearest integer)

# Extract the strata class for the plot extents
fieldPlots_strata <- fieldPlots_extents
plot_strata_vals <- terra::extract(pj_strat, fieldPlots_strata, fun = modal, na.rm = TRUE)
fieldPlots_strata$plot_strata <- round(plot_strata_vals[,2])  

# Build a site-level strata lookup from siteBounds
site_strata_lookup <- siteBounds_strata |>
  sf::st_drop_geometry() |>
  dplyr::select(site_id, site_strata)

# Build a plot-level strata lookup from fieldPlots_strata
plot_strata_lookup <- fieldPlots_strata %>%
  st_drop_geometry() %>%
  dplyr::select(plot_id, plot_strata)

# Join site-level and plot-level strata to modeled estimates via site_id parsed from PlotID
final_Lidar_estimates_strata <- final_Lidar_estimates |>
  dplyr::select(-dplyr::any_of(c("strata", "site_id"))) |>
  dplyr::mutate(site_id = sub("_P\\d+$", "", PlotID)) |>
  dplyr::left_join(site_strata_lookup, by = "site_id") |>
  left_join(plot_strata_lookup, by = c("PlotID" = "plot_id"))


final_NAIP_estimates_strata <- final_NAIP_estimates |>
  dplyr::select(-dplyr::any_of(c("strata", "site_id"))) |>
  dplyr::mutate(site_id = sub("_P\\d+$", "", PlotID)) |>
  dplyr::left_join(site_strata_lookup, by = "site_id") |>
  dplyr::left_join(plot_strata_lookup, by = c("PlotID" = "plot_id"))

# Define descriptive labels for the 9-class stratification (elevation x precipitation, 3x3 grid)
# Classes increment across precipitation (low/mid/high) within each elevation band (low/mid/high)
strata_label_key <- c(
  "Low Elev / Low Precip",   # 1
  "Low Elev / Mid Precip",   # 2
  "Low Elev / High Precip",  # 3
  "Mid Elev / Low Precip",   # 4
  "Mid Elev / Mid Precip",   # 5
  "Mid Elev / High Precip",  # 6
  "High Elev / Low Precip",  # 7
  "High Elev / Mid Precip",  # 8
  "High Elev / High Precip"  # 9
)

# Separate lookups to avoid column name collisions on join
site_strata_labels <- data.frame(site_strata = 1:9, site_strata_label = strata_label_key)
plot_strata_labels <- data.frame(plot_strata  = 1:9, plot_strata_label  = strata_label_key)

# Join site- and plot-level strata labels to both estimate data frames
final_Lidar_estimates_strata <- final_Lidar_estimates_strata |>
  dplyr::left_join(site_strata_labels, by = "site_strata") |>
  dplyr::left_join(plot_strata_labels, by = "plot_strata")

final_NAIP_estimates_strata <- final_NAIP_estimates_strata |>
  dplyr::left_join(site_strata_labels, by = "site_strata") |>
  dplyr::left_join(plot_strata_labels, by = "plot_strata")

# Join n_trees_GPS from plotLevelBiomass to both estimate data frames
trees_lookup <- plotLevelBiomass |>
  sf::st_drop_geometry() |>
  dplyr::select(plot_id, n_trees_GPS)

final_Lidar_estimates_strata <- final_Lidar_estimates_strata |>
  dplyr::left_join(trees_lookup, by = c("PlotID" = "plot_id"))

final_NAIP_estimates_strata <- final_NAIP_estimates_strata |>
  dplyr::left_join(trees_lookup, by = c("PlotID" = "plot_id"))





```


### LiDAR Observed vs. Predicted AGB by Site Strata
```{r fig.width=6, fig.height=7}
full_labels <- c(
  "Low Elev / Low Precip",  "Low Elev / Mid Precip",  "Low Elev / High Precip",
  "Mid Elev / Low Precip",  "Mid Elev / Mid Precip",  "Mid Elev / High Precip",
  "High Elev / Low Precip", "High Elev / Mid Precip", "High Elev / High Precip"
)
strata_order_full <- paste0("Strata ", 1:9, ": ", full_labels)

keep_strata <- c(1, 2, 3, 5, 6, 8, 9)

lidar_plot_data <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(site_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", site_strata, ": ", full_labels[site_strata]),
    levels = strata_order_full
  ))

site_trees_mean <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(site_strata), !is.na(n_trees_GPS)) |>
  dplyr::group_by(strata = site_strata) |>
  dplyr::summarise(mean_trees = mean(n_trees_GPS), .groups = "drop")

lidar_annot_filt <- lidar_strata_stats |>
  dplyr::filter(strata %in% keep_strata) |>
  dplyr::left_join(site_trees_mean, by = "strata") |>
  dplyr::mutate(
    facet_label = factor(paste0("Strata ", strata, ": ", full_labels[strata]), levels = strata_order_full),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

lidar_plot_filt <- lidar_plot_data |> dplyr::filter(site_strata %in% keep_strata)

ax_lim <- range(c(lidar_plot_filt$observed, lidar_plot_filt$predicted), na.rm = TRUE)

ggplot(lidar_plot_filt, aes(x = observed, y = predicted, color = site_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = lidar_annot_filt,
    aes(x = ax_lim[1], y = ax_lim[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(
    colours = c("red3", "orange", "yellow2", "green3"),
    limits  = c(1, 9)
  ) +
  guides(color = "none") +
  coord_equal(xlim = ax_lim, ylim = ax_lim) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "LiDAR: Observed vs. Predicted Plot-Level AGB by Site Strata Class"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```

### NAIP Observed vs. Predicted AGB by Site Strata
```{r fig.width=6, fig.height=7}
naip_plot_data <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(site_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", site_strata, ": ", full_labels[site_strata]),
    levels = strata_order_full
  ))

site_trees_mean_naip <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(site_strata), !is.na(n_trees_GPS)) |>
  dplyr::group_by(strata = site_strata) |>
  dplyr::summarise(mean_trees = mean(n_trees_GPS), .groups = "drop")

naip_annot_filt <- naip_strata_stats |>
  dplyr::filter(strata %in% keep_strata) |>
  dplyr::left_join(site_trees_mean_naip, by = "strata") |>
  dplyr::mutate(
    facet_label = factor(paste0("Strata ", strata, ": ", full_labels[strata]), levels = strata_order_full),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

naip_plot_filt <- naip_plot_data |> dplyr::filter(site_strata %in% keep_strata)

ax_lim_naip <- range(c(naip_plot_filt$observed, naip_plot_filt$predicted), na.rm = TRUE)

ggplot(naip_plot_filt, aes(x = observed, y = predicted, color = site_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = naip_annot_filt,
    aes(x = ax_lim_naip[1], y = ax_lim_naip[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(colours = c("red3", "orange", "yellow2", "green3"), limits = c(1, 9)) +
  guides(color = "none") +
  coord_equal(xlim = ax_lim_naip, ylim = ax_lim_naip) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "NAIP: Observed vs. Predicted Plot-Level AGB by Site Strata"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```

### LiDAR Observed vs. Predicted AGB by Plot Strata
```{r fig.width=6, fig.height=7}
# Per-plot-strata RMSE and R2 for LiDAR
lidar_plot_strata_stats <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::group_by(plot_strata, plot_strata_label) |>
  dplyr::summarise(
    n          = dplyr::n(),
    RMSE       = sqrt(mean((predicted - observed)^2)),
    R2         = cor(predicted, observed)^2,
    mean_trees = mean(n_trees_GPS, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  dplyr::arrange(plot_strata)

plot_strata_order_full <- paste0("Strata ", 1:9, ": ", full_labels)

lidar_ps_data <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
    levels = plot_strata_order_full
  ))

lidar_ps_annot <- lidar_plot_strata_stats |>
  dplyr::mutate(
    facet_label = factor(
      paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
      levels = plot_strata_order_full
    ),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

ax_lidar_ps <- range(c(lidar_ps_data$observed, lidar_ps_data$predicted), na.rm = TRUE)

ggplot(lidar_ps_data, aes(x = observed, y = predicted, color = plot_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = lidar_ps_annot,
    aes(x = ax_lidar_ps[1], y = ax_lidar_ps[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(colours = c("red3", "orange", "yellow2", "green3"), limits = c(1, 9)) +
  guides(color = "none") +
  coord_equal(xlim = ax_lidar_ps, ylim = ax_lidar_ps) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "LiDAR: Observed vs. Predicted Plot-Level AGB by Plot Strata"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```

### NAIP Observed vs. Predicted AGB by Plot Strata
```{r fig.width=6, fig.height=7}
# Per-plot-strata RMSE and R2 for NAIP
naip_plot_strata_stats <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::group_by(plot_strata, plot_strata_label) |>
  dplyr::summarise(
    n          = dplyr::n(),
    RMSE       = sqrt(mean((predicted - observed)^2)),
    R2         = cor(predicted, observed)^2,
    mean_trees = mean(n_trees_GPS, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  dplyr::arrange(plot_strata)

naip_ps_data <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
    levels = plot_strata_order_full
  ))

naip_ps_annot <- naip_plot_strata_stats |>
  dplyr::mutate(
    facet_label = factor(
      paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
      levels = plot_strata_order_full
    ),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

ax_naip_ps <- range(c(naip_ps_data$observed, naip_ps_data$predicted), na.rm = TRUE)

ggplot(naip_ps_data, aes(x = observed, y = predicted, color = plot_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = naip_ps_annot,
    aes(x = ax_naip_ps[1], y = ax_naip_ps[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(colours = c("red3", "orange", "yellow2", "green3"), limits = c(1, 9)) +
  guides(color = "none") +
  coord_equal(xlim = ax_naip_ps, ylim = ax_naip_ps) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "NAIP: Observed vs. Predicted Plot-Level AGB by Plot Strata"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```

### Strata Results Summary
```{r}
# --- Site strata summary ---
site_strata_results <- lidar_strata_stats |>
  dplyr::filter(strata %in% keep_strata) |>
  dplyr::left_join(site_trees_mean, by = "strata") |>
  dplyr::select(strata, strata_label, n,
                lidar_RMSE = RMSE, lidar_R2 = R2) |>
  dplyr::left_join(
    naip_strata_stats |>
      dplyr::filter(strata %in% keep_strata) |>
      dplyr::select(strata, naip_RMSE = RMSE, naip_R2 = R2),
    by = "strata"
  ) |>
  dplyr::left_join(site_trees_mean, by = "strata") |>
  dplyr::select(
    `Strata`         = strata,
    `Class`          = strata_label,
    `n`              = n,
    `LiDAR RMSE`     = lidar_RMSE,
    `LiDAR R2`       = lidar_R2,
    `NAIP RMSE`      = naip_RMSE,
    `NAIP R2`        = naip_R2,
    `Avg Trees/Plot` = mean_trees
  ) |>
  dplyr::mutate(dplyr::across(c(`LiDAR RMSE`, `NAIP RMSE`), \(x) round(x, 0)),
                dplyr::across(c(`LiDAR R2`,   `NAIP R2`),   \(x) round(x, 3)),
                `Avg Trees/Plot` = round(`Avg Trees/Plot`, 1))

# --- Plot strata summary ---
plot_strata_results <- lidar_plot_strata_stats |>
  dplyr::select(strata = plot_strata, strata_label = plot_strata_label, n,
                lidar_RMSE = RMSE, lidar_R2 = R2, mean_trees) |>
  dplyr::left_join(
    naip_plot_strata_stats |>
      dplyr::select(strata = plot_strata, naip_RMSE = RMSE, naip_R2 = R2),
    by = "strata"
  ) |>
  dplyr::select(
    `Strata`         = strata,
    `Class`          = strata_label,
    `n`              = n,
    `LiDAR RMSE`     = lidar_RMSE,
    `LiDAR R2`       = lidar_R2,
    `NAIP RMSE`      = naip_RMSE,
    `NAIP R2`        = naip_R2,
    `Avg Trees/Plot` = mean_trees
  ) |>
  dplyr::mutate(dplyr::across(c(`LiDAR RMSE`, `NAIP RMSE`), \(x) round(x, 0)),
                dplyr::across(c(`LiDAR R2`,   `NAIP R2`),   \(x) round(x, 3)),
                `Avg Trees/Plot` = round(`Avg Trees/Plot`, 1))

cat("=== Site Strata ===\n")
site_strata_results

cat("\n=== Plot Strata ===\n")
plot_strata_results

outStratPath <- file.path(dataOut, "03_StrataPerformance")
write.csv(site_strata_results, file.path(outStratPath, "siteStrataResults.csv"))
write.csv(plot_strata_results, file.path(outStratPath, "plotStrataResults.csv"))

```


### Leave-One-Out Cross Validation
```{r loocv-helper}
# ── Shared LOOCV function ─────────────────────────────────────────────────────
# Holds out each site in turn, refits the allometric model (a * h^b * area^c)
# on the remaining sites, and predicts biomass for the held-out plots.
#
# plot_df : data.frame — PlotID, observed, predicted, Site
# tree_df : data.frame — plot_id, tree_height, tree_area, Site

run_loocv <- function(plot_df, tree_df) {

  # Model form defined once
  predict_biomass <- function(params, tree_list) {
    a <- params[1]; b <- params[2]; c <- params[3]
    vapply(tree_list, function(trees) {
      sum(a * trees$tree_height^b * trees$tree_area^c, na.rm = TRUE)
    }, numeric(1))
  }

  # SSE objective for optim — captures data via closure
  make_objective <- function(tree_list, plot_biomass) {
    function(params) {
      pred  <- predict_biomass(params, tree_list)
      resid <- plot_biomass - pred[names(plot_biomass)]
      sum(resid^2)
    }
  }

  sites   <- unique(plot_df$Site)
  results <- list()

  for (s in sites) {
    message("Holding out: ", s)

    # ── Training data (all sites except s) ──────────────────────────────────
    train_plots <- plot_df |> filter(Site != s)
    train_trees <- tree_df |> filter(Site != s)

    train_tree_list <- split(train_trees, train_trees$plot_id)

    # Drop training plots with no detected trees
    train_plots    <- train_plots |> filter(PlotID %in% names(train_tree_list))
    train_biomass  <- tapply(train_plots$observed, train_plots$PlotID,
                             function(x) x[1])

    # ── Fit model on training data ───────────────────────────────────────────
    fit <- optim(
      par    = c(a = 0.1, b = 1, c = 1),
      fn     = make_objective(train_tree_list, train_biomass),
      method = "L-BFGS-B",
      lower  = c(1e-6, -5, -5),
      upper  = c(100,   5,  5)
    )

    # ── Apply fitted coefficients to the held-out site ───────────────────────
    test_plots     <- plot_df |> filter(Site == s)
    test_trees     <- tree_df |> filter(Site == s)
    test_tree_list <- split(test_trees, test_trees$plot_id)

    # Drop test plots with no detected trees
    test_plots <- test_plots |> filter(PlotID %in% names(test_tree_list))

    cv_pred <- predict_biomass(fit$par, test_tree_list)
    obs     <- tapply(test_plots$observed, test_plots$PlotID,
                      function(x) x[1])

    pnames <- names(cv_pred)
    results[[s]] <- data.frame(
      PlotID        = pnames,
      Site          = s,
      cv_pred       = cv_pred[pnames],
      observed      = obs[pnames],
      original_pred = test_plots$predicted[match(pnames, test_plots$PlotID)]
    )
  }

  output <- bind_rows(results)

  metrics <- output |>
    filter(!is.na(observed)) |>
    group_by(Site) |>
    summarise(
      n_plots = n(),
      RMSE    = sqrt(mean((observed - cv_pred)^2)),
      R2      = 1 - sum((observed - cv_pred)^2) /
                    sum((observed - mean(observed))^2),
      .groups = "drop"
    ) |>
    arrange(R2)

  list(output = output, metrics = metrics)
}

# ── Add Site column to plot and tree tables ───────────────────────────────────
extract_site <- function(x) str_extract(x, "^[^_]+_[^_]+")

naip_plot_df  <- final_NAIP_estimates  |> mutate(Site = extract_site(PlotID))
lidar_plot_df <- final_Lidar_estimates |> mutate(Site = extract_site(PlotID))
naip_tree_df  <- model_trees_NAIP      |> mutate(Site = extract_site(plot_id))
lidar_tree_df <- model_trees_lidar     |> mutate(Site = extract_site(plot_id))
```

```{r loocv-naip}
# ── NAIP LOOCV ────────────────────────────────────────────────────────────────
naip_loocv      <- run_loocv(naip_plot_df, naip_tree_df)
naip_cv_output  <- naip_loocv$output
naip_cv_metrics <- naip_loocv$metrics

print(naip_cv_metrics)

naip_mean_r2 <- mean(naip_cv_metrics$R2)
naip_rmse    <- sqrt(mean((naip_cv_output$observed - naip_cv_output$cv_pred)^2, na.rm = TRUE))

ax_limits <- c(0, max(naip_cv_output$observed,  naip_cv_output$cv_pred,
                      na.rm = TRUE) * 1.05)

# Per-site labels: R² and RMSE
naip_labels <- naip_cv_metrics |>
  mutate(
    label = sprintf("R\u00B2 = %.2f\nRMSE = %.0f kg", R2, RMSE),
    x     = -Inf,
    y     = Inf
  )

ggplot(naip_cv_output |> filter(!is.na(observed)), aes(x = observed, y = cv_pred)) +
  geom_point(size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text(data = naip_labels, aes(x = x, y = y, label = label),
            hjust = -0.05, vjust = 1.2, size = 3.5, inherit.aes = FALSE) +
  scale_x_continuous(limits = ax_limits) +
  scale_y_continuous(limits = ax_limits) +
  facet_wrap(~ Site, ncol = 5) +
  labs(
    title    = "NAIP LOOCV: Observed vs. Predicted AGB",
    subtitle = sprintf("Mean R\u00B2 = %.2f  |  RMSE = %.0f kg", naip_mean_r2, naip_rmse),
    x        = "Observed AGB (kg)",
    y        = "LOOCV Predicted AGB (kg)"
  ) +
  theme_minimal(base_size = 12)

# R² bar chart
ggplot(naip_cv_metrics, aes(y = reorder(Site, R2), x = R2)) +
  geom_col() +
  geom_vline(xintercept = naip_mean_r2, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = naip_mean_r2 + 0.01, y = 0.5,
           label = sprintf("Mean R\u00B2 = %.2f", naip_mean_r2),
           hjust = 0, vjust = 0, color = "red", size = 3.5) +
  labs(title = "NAIP-derived CHM: LOOCV R\u00B2 by Site",
       x = expression(R^2), y = NULL) +
  theme_minimal()
```

```{r loocv-lidar}
# ── LiDAR LOOCV ───────────────────────────────────────────────────────────────
lidar_loocv      <- run_loocv(lidar_plot_df, lidar_tree_df)
lidar_cv_output  <- lidar_loocv$output
lidar_cv_metrics <- lidar_loocv$metrics

print(lidar_cv_metrics)

lidar_mean_r2 <- mean(lidar_cv_metrics$R2)
lidar_rmse    <- sqrt(mean((lidar_cv_output$observed - lidar_cv_output$cv_pred)^2, na.rm = TRUE))

ax_limits <- c(0, max(lidar_cv_output$observed, lidar_cv_output$cv_pred,
                      na.rm = TRUE) * 1.05)

# Per-site labels: R² and RMSE
lidar_labels <- lidar_cv_metrics |>
  mutate(
    label = sprintf("R\u00B2 = %.2f\nRMSE = %.0f kg", R2, RMSE),
    x     = -Inf,
    y     = Inf
  )

ggplot(lidar_cv_output |> filter(!is.na(observed)), aes(x = observed, y = cv_pred)) +
  geom_point(size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text(data = lidar_labels, aes(x = x, y = y, label = label),
            hjust = -0.05, vjust = 1.2, size = 3.5, inherit.aes = FALSE) +
  scale_x_continuous(limits = ax_limits) +
  scale_y_continuous(limits = ax_limits) +
  facet_wrap(~ Site, ncol = 5) +
  labs(
    title    = "LiDAR LOOCV: Observed vs. Predicted AGB",
    subtitle = sprintf("Mean R\u00B2 = %.2f  |  RMSE = %.0f kg", lidar_mean_r2, lidar_rmse),
    x        = "Observed AGB (kg/ha)",
    y        = "LOOCV Predicted AGB (kg/ha)"
  ) +
  theme_minimal(base_size = 12)

# R² bar chart
ggplot(lidar_cv_metrics, aes(y = reorder(Site, R2), x = R2)) +
  geom_col() +
  geom_vline(xintercept = lidar_mean_r2, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = lidar_mean_r2 + 0.01, y = 0.5,
           label = sprintf("Mean R\u00B2 = %.2f", lidar_mean_r2),
           hjust = 0, vjust = 0, color = "red", size = 3.5) +
  labs(title = "LiDAR-derived CHM: LOOCV R\u00B2 by Site",
       x = expression(R^2), y = NULL) +
  theme_minimal()
```


### Export Data
```{r}
outNAIP_estimates <- file.path(dataOut, "01_NAIP")
write.csv(final_NAIP_estimates, file = file.path(outNAIP_estimates, "/NAIP_estimates.csv"), append = TRUE)

outLidar_estimates <- file.path(dataOut, "02_Lidar")
write.csv(final_Lidar_estimates, file = file.path(outLidar_estimates, "/Lidar_estimates.csv"), append = TRUE)

outNAIPTreeCrowns <- file.path(dataOut, "treeCrowns_NAIP")
st_write(allTreeCrowns_NAIP_smoothed, file.path(outNAIPTreeCrowns, "TreeCrowns_NAIP.shp"), overwrite = TRUE)


```



