---
title: "Modeling Aboveground Biomass of Individual Trees in Pinyon-Juniper Woodlands"
author: "Jake Howell"
date: "`r Sys.Date()`"
purpose: 
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
---


```{r}
# Set output formats
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Setup Libraries
```{r, include=FALSE}
# Check required packages are installed
pckgs <- c("sf",
           "dplyr",
           "sp",
           "raster",
           "terra",
           "lidR",
           "mapview",
           "units",
           "ggplot2",
           "broom",
           "purrr",
           "stringr",
           "tidyr",
           "readr",
           "leaflet")

# Point to packages
newPckgs <- pckgs[!(pckgs %in% installed.packages()[,"Package"])]

# If length is greater than 0, install the packages
if(length(newPckgs) > 0) install.packages(newPckgs)

# Load libraries
library(sf)
library(dplyr)
library(sp)
library(raster)
library(terra)
library(lidR)
library(mapview)
library(units)
library(ggplot2)
library(broom)
library(purrr)
library(stringr)
library(tidyr)
library(readr)
library(leaflet)
 

```

## Initialize Data Paths
```{r}
# Set root data location
loc <- r'(S:\ursa\howell\Research\PJ_forests)'

# Set the name of the site data folder 
foldSite <- "01_Sites"

# Set the name of the plot level folder
foldPlot <- "02_Plots"

# Set the folder of NAIP-derived site-level CHMs
foldNAIP_CHM <- "01_NAIP_CHMs"

# Set the folder of the NAIP-derived plot-level CHM's
foldNAIP_plots <- "01_NAIPDerivedCHMs"

# Set the folder of the lidar-derived plot-level CHMs
foldLidar_CHM <- "03_LidarDerivedCHMs/Original"

# Set the name of the plot level .csv
namePlotData <- "plot_level_data_uofu.csv"

## TREE-LEVEL DATA: 
foldTree <- "03_TreeData"

# Set the name of the tree data
nameTreeData <- "tree_level_data.csv"

# Set name to AGB equations
nameAGB <- "all_tree_species_in_pj_fia_plots_new.csv"

# Initialize paths
pathNAIPsites <- file.path(loc, foldSite, foldNAIP_CHM)
pathNAIPplots <- file.path(loc, foldPlot, foldNAIP_plots)
pathLidarPlots <- file.path(loc, foldPlot, foldLidar_CHM)
pathPlotData <- file.path(loc, foldPlot, namePlotData)
pathTreeData <- file.path(loc, foldTree, nameTreeData)
pathAGBData <- file.path(loc, foldTree, nameAGB)

```

### Read in the raster data
```{r}
demPath <- r'(S:\ursa\howell\Research\PJ_forests\07_Ancillary\PRISM_us_dem_800m_bil\PRISM_us_dem_800m_bil.bil)'
precipPath <- r'(S:\ursa\howell\Research\PJ_forests\07_Ancillary\prism_ppt_us_30s_2020_avg_30y\prism_ppt_us_30s_2020_avg_30y.tif)'
pjPath <- r'(S:\ursa\howell\Research\PJ_forests\06_PJ_Extents\polygons\pj_polygon.shp)'
sitePolyPath <- r'(S:\ursa\howell\Research\PJ_forests\01_Sites\00_SiteBounds\site_polygons.shp)'

# Read data
dem <- rast(demPath)
ppt <- rast(precipPath)
pj <- st_read(pjPath)
siteBounds <- st_read(sitePolyPath)

# Reproject
dem <- project(dem, "EPSG:32612")
ppt <- project(ppt, "EPSG:32612")
pj  <- st_transform(pj, "EPSG:32612")
siteBounds <- st_transform(siteBounds, "EPSG:32612")


```

## Read Data
```{r}
### Lidar Plot-Level Rasters----------------------------------------------------
# List the lidar-derived plots within the folder
lidarPlotCHM_names <- list.files(pathLidarPlots, pattern = "\\.tif$", full.names = TRUE)

# Now read the rasters as a named list (strip path and extension)
lidarPlotRasters <- setNames(
  lapply(lidarPlotCHM_names, rast),
  tools::file_path_sans_ext(basename(lidarPlotCHM_names))
)

### NAIP Plot-Level Rasters-----------------------------------------------------
# List the NAIP-derived plots within the folder
NAIPplotCHM_names <- list.files(pathNAIPplots, pattern = "\\.tif$", full.names = TRUE)

# Now read the NAIP rasters as a named list (strip path, extension, and "buffered_" prefix)
NAIPplotRasters <- setNames(
  lapply(NAIPplotCHM_names, rast),
  sub("^buffered_", "", tools::file_path_sans_ext(basename(NAIPplotCHM_names)))
)

### Field Data tables-----------------------------------------------------------
# Read plot field data
plotData <- read.csv(pathPlotData)

# Read tree data
treeData <- read.csv(pathTreeData)

# Read biomass equations
biomassEquations <- read.csv(pathAGBData)

```

### Apply Biomass Equatiosn to Field Data
```{r}
# In the tree data table, alter the species column
treeData <- treeData |>
  rename(Species = tree_sp)

# In the biomass equations table, alter the species column so it matches with the tree table
biomassEquations <- biomassEquations |>
  rename(Species = plants.code)

# Join the tree data to the biomass equations by Species column
treeData <- left_join(treeData, biomassEquations, by = "Species")

# Now estimate biomass for the field data trees by applying the biomass equations
treeData <- treeData |>
  mutate(tree_biomass_kg = exp(beta.0 + beta.1 * log(tree_diam)))

# Remove outliers for plotting purposes
treeData_filtered <- treeData |>
  filter(tree_diam < 100)

# Now plot the biomass of the field data trees and color by species
ggplot2::ggplot(treeData_filtered, aes(x = tree_diam, y = tree_biomass_kg, color = common.name)) +
  geom_point() +
  labs(
    title = "Species-Level AGB from Stem Diameter (DBH)",
    x = "Diameter at Breast Height (cm)", 
    y = "Aboveground Biomass (kg)",
    color = "Tree Species"
  ) +
  theme_minimal()

```

### Create Field Plot Extents
```{r}
# Create function to make one polygon the four sides of each field plot extent
createFieldPlotExtents <- function(row) {
  coords <- matrix(c(
    row$plot_ne_x, row$plot_ne_y,
    row$plot_se_x, row$plot_se_y,
    row$plot_sw_x, row$plot_sw_y,
    row$plot_nw_x, row$plot_nw_y,
    row$plot_ne_x, row$plot_ne_y
  ), ncol = 2, byrow = TRUE)
  st_polygon(list(coords))
}

# Go through each row of the plot data spreadsheet and create the field plot extents
fieldPlots_extents <- plotData |>
  rowwise() |>
  mutate(geometry = st_sfc(createFieldPlotExtents(pick(everything())))) |>
  ungroup()

# Set the current CRS and reproject
fieldPlots_extents <- st_as_sf(fieldPlots_extents, crs = 4326) |>
  st_transform(32612)
```

### Sum the Plot-Level Biomass
```{r}
# Return just the plot ID's from the field plot data
plot_ids <- unique(fieldPlots_extents$plot_id)

# Subset the tree data to just the unique plot names
treeDataInPlots_raw <- treeData[treeData$plot_id %in% plot_ids, ]

# FIX THE ERROR IN LAT/LONG ASSIGNMENT:
treeDataInPlots_fixed <- treeDataInPlots_raw %>%
  mutate(
    # start by assuming they are correct
    lon = tree_x,
    lat = tree_y,

    # identify likely swapped rows:
    # lon should be negative ~ -180..0 and lat should be positive ~ 0..90
    swapped = (tree_x > 0 & tree_x < 90) & (tree_y < 0 & tree_y > -180),

    # for swapped rows, flip them
    lon = if_else(swapped, tree_y, lon),
    lat = if_else(swapped, tree_x, lat)
  )

# Convert the treeDataInPlots to a spatial object
treeDataInPlots_fixed <- st_as_sf(treeDataInPlots_fixed,
                            coords = c("lon", "lat"), 
                            crs = 4326,
                            remove = FALSE)

# Now transform to UTM Zone 12N
treeDataInPlots <- st_transform(treeDataInPlots_fixed, 32612)

# Estimate Biomass Summary for each plot
plotLevelBiomass <- treeDataInPlots |>
  group_by(plot_id) |>
  summarise(
    total_biomass = sum(tree_biomass_kg, na.rm = TRUE),
    mean_biomass = mean(tree_biomass_kg, na.rm = TRUE),
    n_trees_GPS = n(),
    .groups = 'drop'
  )


```

------------------------------ END DATA PREPARATION ----------------------------

## Input Parameters for tree detection
```{r}
# 1) Set window size parameter:
h <- function(x) {
  cw <- 1.0827 * x - 0.3993
  ws <- pmin(10, pmax(3, cw/2))
  return(ws)
}

# 2) NAIP-derived segmentation parameters
th_tree_NAIP  = 0.5
th_seed_NAIP  = 0.3
th_cr_NAIP    = 0.6
max_cr_NAIP   = 15

# 3) Lidar-derived segmentation parameters
th_tree_lidar  = 0.5
th_seed_lidar  = 0.25
th_cr_lidar    = 0.3
max_cr_lidar   = 15

```


### RUN NAIP-CHM TREE DETECTION: NO SMOOTHING
```{r}
# Save the names of the NAIP plots
NAIP_PlotNames <- names(NAIPplotRasters)

# Store tree tops and crowns as a list
NAIP_treePoints_list <- list()
NAIP_treeCrowns_list <- list()

# Loop through each plot CHM and run tree detection
for (plot_id in plot_ids) {

  message("Processing plot: ", plot_id)

   # Select a single plot CHM
    chm <- NAIPplotRasters[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    # Ensure proper projection
    chm <- terra::project(chm, "EPSG:32612")
    
    # Identify Tree Tops
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree_NAIP,
        th_seed  = th_seed_NAIP,
        th_cr    = th_cr_NAIP,
        max_cr   = max_cr_NAIP
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    # Store names as treeID
    names(crowns_poly) <- "treeID"

  # Calculate initial crown area for small/large classification
  crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

  # Extract the maximum height for each crown polygon
  zmax <- tryCatch(
    terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
    error = function(e) {
      message("extract failed: ", e$message)
      NULL
    }
  )

  if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
    message("Height extraction row mismatch for plot: ", plot_id,
            " | crowns=", terra::nrow(crowns_poly),
            " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax)))
    next
  }

  # Assign the maximum height to the crown polygons
  crowns_poly$tree_height_m <- zmax[,2]

  # Convert to sf
  crowns_sf <- sf::st_as_sf(crowns_poly)

  # Add plot ID
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID <- plot_id

  # Store
  NAIP_treePoints_list[[plot_id]] <- ttops_sf
  NAIP_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Bind all rows
allTreePoints_NAIP <- dplyr::bind_rows(NAIP_treePoints_list)
allTreeCrowns_NAIP <- dplyr::bind_rows(NAIP_treeCrowns_list)

# Check projection
allTreePoints_NAIP <- st_transform(allTreePoints_NAIP, st_crs(fieldPlots_extents))
allTreeCrowns_NAIP <- st_transform(allTreeCrowns_NAIP, st_crs(fieldPlots_extents))

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- allTreePoints_NAIP[
  lengths(st_within(allTreePoints_NAIP, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
allTreeCrowns_NAIP_final <- allTreeCrowns_NAIP %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Subset to only the tree tops we want to keep
allTreePoints_NAIP_final <- allTreePoints_NAIP %>%
  inner_join(keys, by = c("plotID", "treeID"))

```

### Apply Crown Smoothing (NAIP)
```{r}
# Get CHM resolution (consistent across all NAIP plots)
naip_res <- res(NAIPplotRasters[[1]])[1]

# Set small crown threshold (3 pixels) — small crowns are left unsmoothed
small_thresh <- 3 * naip_res^2

# Split into small and large crowns by pre-smoothing area
small_crowns_sf <- allTreeCrowns_NAIP_final[allTreeCrowns_NAIP_final$area_m2 <= small_thresh, ]
big_crowns_sf   <- allTreeCrowns_NAIP_final[allTreeCrowns_NAIP_final$area_m2 > small_thresh, ]

# Apply buffer-in / buffer-out smoothing only to large crowns
big_crowns_sf <- big_crowns_sf |>
  st_buffer(-0.75 * naip_res) |>
  st_buffer(0.75 * naip_res) |>
  st_simplify(preserveTopology = TRUE, dTolerance = naip_res / 4)

# Recombine and drop any geometries that collapsed during smoothing
allTreeCrowns_NAIP_smoothed <- bind_rows(big_crowns_sf, small_crowns_sf) |>
  filter(!st_is_empty(geometry))

# Cast to single-part polygons so split crowns become separate rows,
# then keep only the largest fragment per crown
allTreeCrowns_NAIP_smoothed <- allTreeCrowns_NAIP_smoothed |>
  st_cast("POLYGON") |>
  mutate(area_m2 = as.numeric(st_area(geometry))) |>
  group_by(plotID, treeID) |>
  slice_max(area_m2, n = 1, with_ties = FALSE) |>
  ungroup()

cat("Crowns before smoothing:", nrow(allTreeCrowns_NAIP_final), "\n")
cat("Crowns after smoothing: ", nrow(allTreeCrowns_NAIP_smoothed), "\n")
```

### Review results of NAIP-based tree detection
```{r}
# What's the name of the CHM you want to check?
chm_name <- "UOFU_NMQU_P09"

# Get the single NAIP CHM
chm_NAIP <- NAIPplotRasters[[chm_name]]

# Subset the NAIP tree points to selected plot
treePoints_NAIP <- allTreePoints_NAIP_final |> 
  filter(plotID == chm_name)

# Convert to simple feature
treePoints_NAIP <- st_as_sf(treePoints_NAIP)

# Drop 0 values
treePoints_NAIP <- st_zm(treePoints_NAIP, drop = TRUE, what = "ZM")

# Subset the NAIP tree crowns — swap source to compare raw vs smoothed
treeCrowns_NAIP <- allTreeCrowns_NAIP_final |>   
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Subset the NAIP tree crowns to smoothed
treeCrowns_NAIP_smoothed <- allTreeCrowns_NAIP_smoothed |>  
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Get the GPS tree points for the plot
fieldTrees <- treeDataInPlots %>%
  filter(plot_id == chm_name)

# Get the field plot extent 
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

# Plot it
mapview(
    chm_NAIP,
    layer.name = "NAIP CHM",
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(treePoints_NAIP, 
          color = "blue", 
          layer.name = "NAIP Tree Tops") +
  mapview(treeCrowns_NAIP, 
          color = "red", 
          lwd = 3,
          layer.name = "NAIP Crowns") +
  mapview(plot_extent, 
          color = "black", 
          lwd = 3, 
          fill = NA, 
          layer.name = "Field Plot Extents")

# Plot the smoothed trees
mapview(
    chm_NAIP,
    layer.name = "NAIP CHM",
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(treePoints_NAIP, 
          color = "blue", 
          layer.name = "NAIP Tree Tops") +
  mapview(treeCrowns_NAIP_smoothed, 
          color = "red", 
          lwd = 3,
          layer.name = "NAIP Crowns") +
  mapview(plot_extent, 
          color = "black", 
          lwd = 3, 
          fill = NA, 
          layer.name = "Field Plot Extents")

```


### RUN LIDAR-CHM TREE DETECTION
```{r}
# Save the names of the Lidar plots
Lidar_PlotNames <- names(lidarPlotRasters)

# Store tree tops and crowns as a list
Lidar_treePoints_list <- list()
Lidar_treeCrowns_list <- list()

# Loop through each plot CHM and run tree detection
for (plot_id in Lidar_PlotNames) {

  message("Processing plot: ", plot_id)
  
   # Select a single plot CHM
    chm <- lidarPlotRasters[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    # Ensure proper projection
    chm <- terra::project(chm, "EPSG:32612")
    
    # Identify Tree Tops
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree_lidar,
        th_seed  = th_seed_lidar,
        th_cr    = th_cr_lidar,
        max_cr   = max_cr_lidar
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    # Store names as treeID
    names(crowns_poly) <- "treeID"  

  # Drop empty / invalid geometries
  crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

  if (nrow(crowns_poly) == 0) next

  # Calculate Crown Area in m2
  crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

  # Extract the maximum height for each crown polygon
  zmax <- tryCatch(
    terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
    error = function(e) {
      message("extract failed: ", e$message)
      NULL
    }
  )

  if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
    message("Height extraction row mismatch for plot: ", plot_id,
            " | crowns=", terra::nrow(crowns_poly),
            " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax)))
    next
  }

  # Assign the maximum height to the crown polygons
  crowns_poly$tree_height_m <- zmax[,2]

  # Convert to sf 
  crowns_sf <- sf::st_as_sf(crowns_poly)

  # Add plot ID
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID <- plot_id

  # Store
  Lidar_treePoints_list[[plot_id]] <- ttops_sf
  Lidar_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Bind all rows
allTreePoints_Lidar <- dplyr::bind_rows(Lidar_treePoints_list)
allTreeCrowns_Lidar <- dplyr::bind_rows(Lidar_treeCrowns_list)

# Check projection
allTreePoints_Lidar <- st_transform(allTreePoints_Lidar, st_crs(fieldPlots_extents))
allTreeCrowns_Lidar <- st_transform(allTreeCrowns_Lidar, st_crs(fieldPlots_extents))

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- allTreePoints_Lidar[
  lengths(st_within(allTreePoints_Lidar, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
allTreeCrowns_Lidar_final <- allTreeCrowns_Lidar %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Subset to only the tree tops we want to keep
allTreePoints_Lidar_final <- allTreePoints_Lidar %>%
  inner_join(keys, by = c("plotID", "treeID"))
```

### Review results of Lidar-based tree detection
```{r}
# What's the name of the CHM you want to check?
chm_name <- "UOFU_AZTU_P03"

# Get the single Lidar CHM
chm_Lidar <- lidarPlotRasters[[chm_name]]

# Subset the Lidar tree points to selected plot
treePoints_Lidar <- allTreePoints_Lidar_final |> 
  filter(plotID == chm_name)

# Convert to simple feature
treePoints_Lidar <- st_as_sf(treePoints_Lidar)

# Drop 0 values
treePoints_Lidar <- st_zm(treePoints_Lidar, drop = TRUE, what = "ZM")

# Subset the Lidar tree crowns — swap source to compare raw vs smoothed
treeCrowns_Lidar <- allTreeCrowns_Lidar_final |>   
  filter(plotID == chm_name) |>
  st_as_sf() |>
  st_zm(drop = TRUE, what = "ZM")

# Get the GPS tree points for the plot
fieldTrees <- treeDataInPlots %>%
  filter(plot_id == chm_name)

# Get the field plot extent 
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

class(treePoints_Lidar)
# Plot it
mapview(
    chm_Lidar,
    layer.name = "Lidar CHM",
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))) +
  mapview(treePoints_Lidar, 
          color = "blue", 
          layer.name = "Lidar Tree Tops") +
  mapview(treeCrowns_Lidar, 
          color = "red", 
          lwd = 3,
          layer.name = "Lidar Crowns") +
  mapview(plot_extent, 
          color = "black", 
          lwd = 3, 
          fill = NA, 
          layer.name = "Field Plot Extents")

```

### Create Models for the NAIP-derived tree list
```{r}
# Create tree and plot df
model_trees_NAIP <- allTreeCrowns_NAIP_smoothed
model_plots_NAIP <- plotLevelBiomass

# Select tree variables
model_trees_NAIP <- model_trees_NAIP[,c("plotID", "tree_height_m", "area_m2")]
model_trees_NAIP <- st_drop_geometry(model_trees_NAIP)

# Select plot variables
model_plots_NAIP <- model_plots_NAIP[,c("plot_id", "total_biomass")]
model_plots_NAIP <- st_drop_geometry(model_plots_NAIP)

# Subset data to these variables
colnames(model_trees_NAIP) <- c("plot_id", "tree_height", "tree_area")
colnames(model_plots_NAIP) <- c("plot_id", "plot_biomass")

# Merge the data
model_df <- merge(model_plots_NAIP, model_trees_NAIP)

# Split into list by plot
plot_ids <- unique(model_df$plot_id)
tree_list <- split(model_df, model_df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(model_df$plot_biomass, model_df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  # ensure ordering matches
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1, 
                b = 1, 
                c = 1
)

# Fit the parameters
fit <- optim(
  par = 
    start_vals,
  fn = objective,
  
  # Broyden, Fletcher, Goldfarb, and Shanno
  # Uses function values and gradients to build up a picture of the surface to be optimized
  method = "L-BFGS-B",
  lower = c(1e-6, 
            -5,
            -5),
  upper = c(100,  
            5,
            5))

# Whats here?
# The best set of parameters found
# The coefficients for individual tree level biomass! 
NAIPCoefficients <- fit$par 

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_NAIP_estimates <- left_join(predicted_df, observed_df, by = "PlotID")

# Observed and predicted vectors
obs <- plot_biomass
pred <- predicted[names(plot_biomass)]   

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2 <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
NAIP_TreeDetection_ModResults <- data.frame(
  RMSE = RMSE,
  MAE = MAE,
  Bias = Bias,
  MAPE_percent = MAPE,
  R2 = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result 
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)

# Final NAIP estimates dataframe
final_NAIP_estimates <- final_NAIP_estimates |>
  dplyr::select(PlotID,
                predicted,
                plot_biomass) |>
  rename("observed" = plot_biomass)

```

### Create models for lidar-derived tree list 
```{r}
# Create tree and plot df
model_trees_lidar <- allTreeCrowns_Lidar_final
model_plots_lidar <- plotLevelBiomass

# Select tree variables
model_trees_lidar <- model_trees_lidar[,c("plotID", "tree_height_m", "area_m2")]
model_trees_lidar <- st_drop_geometry(model_trees_lidar)

# Select plot variables
model_plots_lidar <- model_plots_lidar[,c("plot_id", "total_biomass")]
model_plots_lidar <- st_drop_geometry(model_plots_lidar)

# Subset data to these variables
colnames(model_trees_lidar) <- c("plot_id", "tree_height", "tree_area")
colnames(model_plots_lidar) <- c("plot_id", "plot_biomass")

# Merge the data
df <- merge(model_plots_lidar, model_trees_lidar)

# Split into list by plot
plot_ids <- unique(df$plot_id)
tree_list <- split(df, df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(df$plot_biomass, df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  # ensure ordering matches
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1, 
                b = 1, 
                c = 1
)

# Fit the parameters
fit <- optim(
  par = 
    start_vals,
  fn = objective,
  
  # Broyden, Fletcher, Goldfarb, and Shanno
  # Uses function values and gradients to build up a picture of the surface to be optimized
  method = "L-BFGS-B",
  lower = c(1e-6, 
            -5,
            -5),
  upper = c(100,  
            5,
            5))

# Whats here?
# The best set of parameters found
# The coefficients for individual tree level biomass! 
lidarCoefficients <- fit$par 

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_Lidar_estimates <- left_join(predicted_df, observed_df, by = "PlotID")

# Observed and predicted vectors
obs <- plot_biomass
pred <- predicted[names(plot_biomass)]   

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2 <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
Lidar_TreeDetection_ModResults <- data.frame(
  RMSE = RMSE,
  MAE = MAE,
  Bias = Bias,
  MAPE_percent = MAPE,
  R2 = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result 
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)

# Edit final lidar estimates table
final_Lidar_estimates <- final_Lidar_estimates |>
  dplyr::select(PlotID,
         predicted,
         plot_biomass) |>
  rename("observed" = plot_biomass)

```

### What are my modeling results? 
```{r}
# NAIP-derived tree detection result
print(NAIP_TreeDetection_ModResults)

# What are my individual tree level coefficients?
print(NAIPCoefficients)
NAIPCoefficients_df <- as.data.frame(NAIPCoefficients)

# Lidar-derived tree detection result
print(Lidar_TreeDetection_ModResults)

# What are my individual tree level coefficients
print(lidarCoefficients)
lidarCoefficients_df <- as.data.frame(lidarCoefficients)

```


### Apply model coefficients to the NAIP tree list
```{r}
# Apply model coefficients to the NAIP-derived smoothed tree list
NAIPTreeList_AGB <- allTreeCrowns_NAIP_smoothed
NAIPTreeList_AGB$beta.a <- 3.4553996 
NAIPTreeList_AGB$beta.b <- 0.8206139 
NAIPTreeList_AGB$beta.c <- 0.9328608 

# Calculate tree-level AGB using coefficients
NAIPTreeList_AGB <- NAIPTreeList_AGB |>
  mutate(tree_biomass = beta.a * (tree_height_m^beta.b) * (area_m2^beta.c))

# Calculate plot-level AGB by summing tree-level biomass
NAIPPlots_AGB <- NAIPTreeList_AGB |>
  group_by(plotID) |>
  summarise(predicted_AGB = sum(tree_biomass, na.rm = TRUE)) |>
  rename(PlotID = "plotID")


```

### Apply model coefficients to the Lidar tree list
```{r}
LidarTreeList_AGB <- allTreeCrowns_Lidar_final
LidarTreeList_AGB$beta.a <- 4.6192000 
LidarTreeList_AGB$beta.b <- 1.1009588 
LidarTreeList_AGB$beta.c <- 0.6555886 

# Calculate tree-level AGB using coefficients
LidarTreeList_AGB <- LidarTreeList_AGB |>
  mutate(tree_biomass = beta.a * (tree_height_m^beta.b) * (area_m2^beta.c))

```

### Create landscape stratification
```{r}
# Create a buffer around PJ ecosystem
pj_buffer <- st_buffer(pj, dist = 1000)  

# Merge into a single spatial layer
pj_buffer <- st_union(pj_buffer)

# Clip & mask my rasters to the PJ extent
dem_pj <- terra::mask(terra::crop(dem, pj_buffer), pj_buffer)
ppt_pj <- terra::mask(terra::crop(ppt, pj_buffer), pj_buffer)

# Create a landscape stratification
names(dem_pj) <- "elevation"
names(ppt_pj) <- "precipitation"

# Create a composite raster
compRast <- c(dem_pj, ppt_pj)

# Create a landscape stratification
stratRast <- sgsR::strat_quantiles(
  mraster = compRast[[1:2]],
  nStrat = list(3, 3),
  plot = FALSE,
  map = TRUE
)

# Select the stratified raster
pj_strat <- stratRast[[3]]

# # Export:
# outStratPath <- r'(S:\ursa\howell\Research\PJ_forests\07_Ancillary\Stratification\pj_strat.tif)'
# writeRaster(pj_strat, file.path(outStratPath))

# Extract the strata class for the site extents
siteBounds_strata <- siteBounds
site_strata_vals <- terra::extract(pj_strat, siteBounds_strata, fun = modal, na.rm = TRUE)
siteBounds_strata$site_strata <- round(strata_vals[,2])  # Extract the mean strata value and round to nearest integer)

# Extract the strata class for the plot extents
fieldPlots_strata <- fieldPlots_extents
plot_strata_vals <- terra::extract(pj_strat, fieldPlots_strata, fun = modal, na.rm = TRUE)
fieldPlots_strata$plot_strata <- round(plot_strata_vals[,2])  

# Build a site-level strata lookup from siteBounds
site_strata_lookup <- siteBounds_strata |>
  sf::st_drop_geometry() |>
  dplyr::select(site_id, site_strata)

# Build a plot-level strata lookup from fieldPlots_strata
plot_strata_lookup <- fieldPlots_strata %>%
  st_drop_geometry() %>%
  dplyr::select(plot_id, plot_strata)

# Join site-level and plot-level strata to modeled estimates via site_id parsed from PlotID
final_Lidar_estimates_strata <- final_Lidar_estimates |>
  dplyr::select(-dplyr::any_of(c("strata", "site_id"))) |>
  dplyr::mutate(site_id = sub("_P\\d+$", "", PlotID)) |>
  dplyr::left_join(site_strata_lookup, by = "site_id") |>
  left_join(plot_strata_lookup, by = c("PlotID" = "plot_id"))


final_NAIP_estimates_strata <- final_NAIP_estimates |>
  dplyr::select(-dplyr::any_of(c("strata", "site_id"))) |>
  dplyr::mutate(site_id = sub("_P\\d+$", "", PlotID)) |>
  dplyr::left_join(site_strata_lookup, by = "site_id") |>
  dplyr::left_join(plot_strata_lookup, by = c("PlotID" = "plot_id"))

# Define descriptive labels for the 9-class stratification (elevation x precipitation, 3x3 grid)
# Classes increment across precipitation (low/mid/high) within each elevation band (low/mid/high)
strata_label_key <- c(
  "Low Elev / Low Precip",   # 1
  "Low Elev / Mid Precip",   # 2
  "Low Elev / High Precip",  # 3
  "Mid Elev / Low Precip",   # 4
  "Mid Elev / Mid Precip",   # 5
  "Mid Elev / High Precip",  # 6
  "High Elev / Low Precip",  # 7
  "High Elev / Mid Precip",  # 8
  "High Elev / High Precip"  # 9
)

# Separate lookups to avoid column name collisions on join
site_strata_labels <- data.frame(site_strata = 1:9, site_strata_label = strata_label_key)
plot_strata_labels <- data.frame(plot_strata  = 1:9, plot_strata_label  = strata_label_key)

# Join site- and plot-level strata labels to both estimate data frames
final_Lidar_estimates_strata <- final_Lidar_estimates_strata |>
  dplyr::left_join(site_strata_labels, by = "site_strata") |>
  dplyr::left_join(plot_strata_labels, by = "plot_strata")

final_NAIP_estimates_strata <- final_NAIP_estimates_strata |>
  dplyr::left_join(site_strata_labels, by = "site_strata") |>
  dplyr::left_join(plot_strata_labels, by = "plot_strata")

# Join n_trees_GPS from plotLevelBiomass to both estimate data frames
trees_lookup <- plotLevelBiomass |>
  sf::st_drop_geometry() |>
  dplyr::select(plot_id, n_trees_GPS)

final_Lidar_estimates_strata <- final_Lidar_estimates_strata |>
  dplyr::left_join(trees_lookup, by = c("PlotID" = "plot_id"))

final_NAIP_estimates_strata <- final_NAIP_estimates_strata |>
  dplyr::left_join(trees_lookup, by = c("PlotID" = "plot_id"))





```
### LiDAR Observed vs. Predicted AGB by Site Strata
```{r fig.width=6, fig.height=7}
full_labels <- c(
  "Low Elev / Low Precip",  "Low Elev / Mid Precip",  "Low Elev / High Precip",
  "Mid Elev / Low Precip",  "Mid Elev / Mid Precip",  "Mid Elev / High Precip",
  "High Elev / Low Precip", "High Elev / Mid Precip", "High Elev / High Precip"
)
strata_order_full <- paste0("Strata ", 1:9, ": ", full_labels)

keep_strata <- c(1, 2, 3, 5, 6, 8, 9)

lidar_plot_data <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(site_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", site_strata, ": ", full_labels[site_strata]),
    levels = strata_order_full
  ))

site_trees_mean <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(site_strata), !is.na(n_trees_GPS)) |>
  dplyr::group_by(strata = site_strata) |>
  dplyr::summarise(mean_trees = mean(n_trees_GPS), .groups = "drop")

lidar_annot_filt <- lidar_strata_stats |>
  dplyr::filter(strata %in% keep_strata) |>
  dplyr::left_join(site_trees_mean, by = "strata") |>
  dplyr::mutate(
    facet_label = factor(paste0("Strata ", strata, ": ", full_labels[strata]), levels = strata_order_full),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

lidar_plot_filt <- lidar_plot_data |> dplyr::filter(site_strata %in% keep_strata)

ax_lim <- range(c(lidar_plot_filt$observed, lidar_plot_filt$predicted), na.rm = TRUE)

ggplot(lidar_plot_filt, aes(x = observed, y = predicted, color = site_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = lidar_annot_filt,
    aes(x = ax_lim[1], y = ax_lim[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(
    colours = c("red3", "orange", "yellow2", "green3"),
    limits  = c(1, 9)
  ) +
  guides(color = "none") +
  coord_equal(xlim = ax_lim, ylim = ax_lim) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "LiDAR: Observed vs. Predicted Plot-Level AGB by Site Strata Class"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```
### NAIP Observed vs. Predicted AGB by Site Strata
```{r fig.width=6, fig.height=7}
naip_plot_data <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(site_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", site_strata, ": ", full_labels[site_strata]),
    levels = strata_order_full
  ))

site_trees_mean_naip <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(site_strata), !is.na(n_trees_GPS)) |>
  dplyr::group_by(strata = site_strata) |>
  dplyr::summarise(mean_trees = mean(n_trees_GPS), .groups = "drop")

naip_annot_filt <- naip_strata_stats |>
  dplyr::filter(strata %in% keep_strata) |>
  dplyr::left_join(site_trees_mean_naip, by = "strata") |>
  dplyr::mutate(
    facet_label = factor(paste0("Strata ", strata, ": ", full_labels[strata]), levels = strata_order_full),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

naip_plot_filt <- naip_plot_data |> dplyr::filter(site_strata %in% keep_strata)

ax_lim_naip <- range(c(naip_plot_filt$observed, naip_plot_filt$predicted), na.rm = TRUE)

ggplot(naip_plot_filt, aes(x = observed, y = predicted, color = site_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = naip_annot_filt,
    aes(x = ax_lim_naip[1], y = ax_lim_naip[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(colours = c("red3", "orange", "yellow2", "green3"), limits = c(1, 9)) +
  guides(color = "none") +
  coord_equal(xlim = ax_lim_naip, ylim = ax_lim_naip) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "NAIP: Observed vs. Predicted Plot-Level AGB by Site Strata"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```


### LiDAR Observed vs. Predicted AGB by Plot Strata
```{r fig.width=6, fig.height=7}
# Per-plot-strata RMSE and R2 for LiDAR
lidar_plot_strata_stats <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::group_by(plot_strata, plot_strata_label) |>
  dplyr::summarise(
    n          = dplyr::n(),
    RMSE       = sqrt(mean((predicted - observed)^2)),
    R2         = cor(predicted, observed)^2,
    mean_trees = mean(n_trees_GPS, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  dplyr::arrange(plot_strata)

plot_strata_order_full <- paste0("Strata ", 1:9, ": ", full_labels)

lidar_ps_data <- final_Lidar_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
    levels = plot_strata_order_full
  ))

lidar_ps_annot <- lidar_plot_strata_stats |>
  dplyr::mutate(
    facet_label = factor(
      paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
      levels = plot_strata_order_full
    ),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

ax_lidar_ps <- range(c(lidar_ps_data$observed, lidar_ps_data$predicted), na.rm = TRUE)

ggplot(lidar_ps_data, aes(x = observed, y = predicted, color = plot_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = lidar_ps_annot,
    aes(x = ax_lidar_ps[1], y = ax_lidar_ps[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(colours = c("red3", "orange", "yellow2", "green3"), limits = c(1, 9)) +
  guides(color = "none") +
  coord_equal(xlim = ax_lidar_ps, ylim = ax_lidar_ps) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "LiDAR: Observed vs. Predicted Plot-Level AGB by Plot Strata"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```


### NAIP Observed vs. Predicted AGB by Plot Strata
```{r fig.width=6, fig.height=7}
# Per-plot-strata RMSE and R2 for NAIP
naip_plot_strata_stats <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::group_by(plot_strata, plot_strata_label) |>
  dplyr::summarise(
    n          = dplyr::n(),
    RMSE       = sqrt(mean((predicted - observed)^2)),
    R2         = cor(predicted, observed)^2,
    mean_trees = mean(n_trees_GPS, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  dplyr::arrange(plot_strata)

naip_ps_data <- final_NAIP_estimates_strata |>
  dplyr::filter(!is.na(plot_strata)) |>
  dplyr::mutate(facet_label = factor(
    paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
    levels = plot_strata_order_full
  ))

naip_ps_annot <- naip_plot_strata_stats |>
  dplyr::mutate(
    facet_label = factor(
      paste0("Strata ", plot_strata, ": ", full_labels[plot_strata]),
      levels = plot_strata_order_full
    ),
    label = sprintf("RMSE = %.0f\nR\u00B2 = %.3f\nAvg Trees/Plot = %.0f", RMSE, R2, mean_trees)
  )

ax_naip_ps <- range(c(naip_ps_data$observed, naip_ps_data$predicted), na.rm = TRUE)

ggplot(naip_ps_data, aes(x = observed, y = predicted, color = plot_strata)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(alpha = 0.75, size = 2) +
  geom_text(
    data = naip_ps_annot,
    aes(x = ax_naip_ps[1], y = ax_naip_ps[2], label = label),
    hjust = 0, vjust = 1, size = 2.5, fontface = "bold", color = "grey20",
    inherit.aes = FALSE
  ) +
  facet_wrap(~facet_label, ncol = 3) +
  scale_color_gradientn(colours = c("red3", "orange", "yellow2", "green3"), limits = c(1, 9)) +
  guides(color = "none") +
  coord_equal(xlim = ax_naip_ps, ylim = ax_naip_ps) +
  labs(
    x     = "Observed AGB (kg)",
    y     = "Predicted AGB (kg)",
    title = "NAIP: Observed vs. Predicted Plot-Level AGB by Plot Strata"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    strip.text       = element_text(face = "bold", size = 7),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 10)
  )
```




### Strata Results Summary
```{r}
# --- Site strata summary ---
site_strata_results <- lidar_strata_stats |>
  dplyr::filter(strata %in% keep_strata) |>
  dplyr::left_join(site_trees_mean, by = "strata") |>
  dplyr::select(strata, strata_label, n,
                lidar_RMSE = RMSE, lidar_R2 = R2) |>
  dplyr::left_join(
    naip_strata_stats |>
      dplyr::filter(strata %in% keep_strata) |>
      dplyr::select(strata, naip_RMSE = RMSE, naip_R2 = R2),
    by = "strata"
  ) |>
  dplyr::left_join(site_trees_mean, by = "strata") |>
  dplyr::select(
    `Strata`         = strata,
    `Class`          = strata_label,
    `n`              = n,
    `LiDAR RMSE`     = lidar_RMSE,
    `LiDAR R2`       = lidar_R2,
    `NAIP RMSE`      = naip_RMSE,
    `NAIP R2`        = naip_R2,
    `Avg Trees/Plot` = mean_trees
  ) |>
  dplyr::mutate(dplyr::across(c(`LiDAR RMSE`, `NAIP RMSE`), \(x) round(x, 0)),
                dplyr::across(c(`LiDAR R2`,   `NAIP R2`),   \(x) round(x, 3)),
                `Avg Trees/Plot` = round(`Avg Trees/Plot`, 1))

# --- Plot strata summary ---
plot_strata_results <- lidar_plot_strata_stats |>
  dplyr::select(strata = plot_strata, strata_label = plot_strata_label, n,
                lidar_RMSE = RMSE, lidar_R2 = R2, mean_trees) |>
  dplyr::left_join(
    naip_plot_strata_stats |>
      dplyr::select(strata = plot_strata, naip_RMSE = RMSE, naip_R2 = R2),
    by = "strata"
  ) |>
  dplyr::select(
    `Strata`         = strata,
    `Class`          = strata_label,
    `n`              = n,
    `LiDAR RMSE`     = lidar_RMSE,
    `LiDAR R2`       = lidar_R2,
    `NAIP RMSE`      = naip_RMSE,
    `NAIP R2`        = naip_R2,
    `Avg Trees/Plot` = mean_trees
  ) |>
  dplyr::mutate(dplyr::across(c(`LiDAR RMSE`, `NAIP RMSE`), \(x) round(x, 0)),
                dplyr::across(c(`LiDAR R2`,   `NAIP R2`),   \(x) round(x, 3)),
                `Avg Trees/Plot` = round(`Avg Trees/Plot`, 1))

cat("=== Site Strata ===\n")
site_strata_results

cat("\n=== Plot Strata ===\n")
plot_strata_results

outStratPath <- r'(S:\ursa\howell\Research\PJ_forests\08_ModelResults\03_StrataPerformance)'
write.csv(site_strata_results, file.path(outStratPath, "siteStrataResults.csv"))
write.csv(plot_strata_results, file.path(outStratPath, "plotStrataResults.csv"))

```


### Leave-One-Out Cross Validation
```{r loocv-helper}
# ── Shared LOOCV function ─────────────────────────────────────────────────────
# Holds out each site in turn, refits the allometric model (a * h^b * area^c)
# on the remaining sites, and predicts biomass for the held-out plots.
#
# plot_df : data.frame — PlotID, observed, predicted, Site
# tree_df : data.frame — plot_id, tree_height, tree_area, Site

run_loocv <- function(plot_df, tree_df) {

  # Model form defined once
  predict_biomass <- function(params, tree_list) {
    a <- params[1]; b <- params[2]; c <- params[3]
    vapply(tree_list, function(trees) {
      sum(a * trees$tree_height^b * trees$tree_area^c, na.rm = TRUE)
    }, numeric(1))
  }

  # SSE objective for optim — captures data via closure
  make_objective <- function(tree_list, plot_biomass) {
    function(params) {
      pred  <- predict_biomass(params, tree_list)
      resid <- plot_biomass - pred[names(plot_biomass)]
      sum(resid^2)
    }
  }

  sites   <- unique(plot_df$Site)
  results <- list()

  for (s in sites) {
    message("Holding out: ", s)

    # ── Training data (all sites except s) ──────────────────────────────────
    train_plots <- plot_df |> filter(Site != s)
    train_trees <- tree_df |> filter(Site != s)

    train_tree_list <- split(train_trees, train_trees$plot_id)

    # Drop training plots with no detected trees
    train_plots    <- train_plots |> filter(PlotID %in% names(train_tree_list))
    train_biomass  <- tapply(train_plots$observed, train_plots$PlotID,
                             function(x) x[1])

    # ── Fit model on training data ───────────────────────────────────────────
    fit <- optim(
      par    = c(a = 0.1, b = 1, c = 1),
      fn     = make_objective(train_tree_list, train_biomass),
      method = "L-BFGS-B",
      lower  = c(1e-6, -5, -5),
      upper  = c(100,   5,  5)
    )

    # ── Apply fitted coefficients to the held-out site ───────────────────────
    test_plots     <- plot_df |> filter(Site == s)
    test_trees     <- tree_df |> filter(Site == s)
    test_tree_list <- split(test_trees, test_trees$plot_id)

    # Drop test plots with no detected trees
    test_plots <- test_plots |> filter(PlotID %in% names(test_tree_list))

    cv_pred <- predict_biomass(fit$par, test_tree_list)
    obs     <- tapply(test_plots$observed, test_plots$PlotID,
                      function(x) x[1])

    pnames <- names(cv_pred)
    results[[s]] <- data.frame(
      PlotID        = pnames,
      Site          = s,
      cv_pred       = cv_pred[pnames],
      observed      = obs[pnames],
      original_pred = test_plots$predicted[match(pnames, test_plots$PlotID)]
    )
  }

  output <- bind_rows(results)

  metrics <- output |>
    filter(!is.na(observed)) |>
    group_by(Site) |>
    summarise(
      n_plots = n(),
      RMSE    = sqrt(mean((observed - cv_pred)^2)),
      R2      = 1 - sum((observed - cv_pred)^2) /
                    sum((observed - mean(observed))^2),
      .groups = "drop"
    ) |>
    arrange(R2)

  list(output = output, metrics = metrics)
}

# ── Add Site column to plot and tree tables ───────────────────────────────────
extract_site <- function(x) str_extract(x, "^[^_]+_[^_]+")

naip_plot_df  <- final_NAIP_estimates  |> mutate(Site = extract_site(PlotID))
lidar_plot_df <- final_Lidar_estimates |> mutate(Site = extract_site(PlotID))
naip_tree_df  <- model_trees_NAIP      |> mutate(Site = extract_site(plot_id))
lidar_tree_df <- model_trees_lidar     |> mutate(Site = extract_site(plot_id))
```

```{r loocv-naip}
# ── NAIP LOOCV ────────────────────────────────────────────────────────────────
naip_loocv      <- run_loocv(naip_plot_df, naip_tree_df)
naip_cv_output  <- naip_loocv$output
naip_cv_metrics <- naip_loocv$metrics

print(naip_cv_metrics)

naip_mean_r2 <- mean(naip_cv_metrics$R2)
naip_rmse    <- sqrt(mean((naip_cv_output$observed - naip_cv_output$cv_pred)^2, na.rm = TRUE))

ax_limits <- c(0, max(naip_cv_output$observed,  naip_cv_output$cv_pred,
                      na.rm = TRUE) * 1.05)

# Per-site labels: R² and RMSE
naip_labels <- naip_cv_metrics |>
  mutate(
    label = sprintf("R\u00B2 = %.2f\nRMSE = %.0f kg", R2, RMSE),
    x     = -Inf,
    y     = Inf
  )

ggplot(naip_cv_output |> filter(!is.na(observed)), aes(x = observed, y = cv_pred)) +
  geom_point(size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text(data = naip_labels, aes(x = x, y = y, label = label),
            hjust = -0.05, vjust = 1.2, size = 3.5, inherit.aes = FALSE) +
  scale_x_continuous(limits = ax_limits) +
  scale_y_continuous(limits = ax_limits) +
  facet_wrap(~ Site, ncol = 5) +
  labs(
    title    = "NAIP LOOCV: Observed vs. Predicted AGB",
    subtitle = sprintf("Mean R\u00B2 = %.2f  |  RMSE = %.0f kg", naip_mean_r2, naip_rmse),
    x        = "Observed AGB (kg)",
    y        = "LOOCV Predicted AGB (kg)"
  ) +
  theme_minimal(base_size = 12)

# R² bar chart
ggplot(naip_cv_metrics, aes(y = reorder(Site, R2), x = R2)) +
  geom_col() +
  geom_vline(xintercept = naip_mean_r2, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = naip_mean_r2 + 0.01, y = 0.5,
           label = sprintf("Mean R\u00B2 = %.2f", naip_mean_r2),
           hjust = 0, vjust = 0, color = "red", size = 3.5) +
  labs(title = "NAIP-derived CHM: LOOCV R\u00B2 by Site",
       x = expression(R^2), y = NULL) +
  theme_minimal()
```

```{r loocv-lidar}
# ── LiDAR LOOCV ───────────────────────────────────────────────────────────────
lidar_loocv      <- run_loocv(lidar_plot_df, lidar_tree_df)
lidar_cv_output  <- lidar_loocv$output
lidar_cv_metrics <- lidar_loocv$metrics

print(lidar_cv_metrics)

lidar_mean_r2 <- mean(lidar_cv_metrics$R2)
lidar_rmse    <- sqrt(mean((lidar_cv_output$observed - lidar_cv_output$cv_pred)^2, na.rm = TRUE))

ax_limits <- c(0, max(lidar_cv_output$observed, lidar_cv_output$cv_pred,
                      na.rm = TRUE) * 1.05)

# Per-site labels: R² and RMSE
lidar_labels <- lidar_cv_metrics |>
  mutate(
    label = sprintf("R\u00B2 = %.2f\nRMSE = %.0f kg", R2, RMSE),
    x     = -Inf,
    y     = Inf
  )

ggplot(lidar_cv_output |> filter(!is.na(observed)), aes(x = observed, y = cv_pred)) +
  geom_point(size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_text(data = lidar_labels, aes(x = x, y = y, label = label),
            hjust = -0.05, vjust = 1.2, size = 3.5, inherit.aes = FALSE) +
  scale_x_continuous(limits = ax_limits) +
  scale_y_continuous(limits = ax_limits) +
  facet_wrap(~ Site, ncol = 5) +
  labs(
    title    = "LiDAR LOOCV: Observed vs. Predicted AGB",
    subtitle = sprintf("Mean R\u00B2 = %.2f  |  RMSE = %.0f kg", lidar_mean_r2, lidar_rmse),
    x        = "Observed AGB (kg/ha)",
    y        = "LOOCV Predicted AGB (kg/ha)"
  ) +
  theme_minimal(base_size = 12)

# R² bar chart
ggplot(lidar_cv_metrics, aes(y = reorder(Site, R2), x = R2)) +
  geom_col() +
  geom_vline(xintercept = lidar_mean_r2, color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = lidar_mean_r2 + 0.01, y = 0.5,
           label = sprintf("Mean R\u00B2 = %.2f", lidar_mean_r2),
           hjust = 0, vjust = 0, color = "red", size = 3.5) +
  labs(title = "LiDAR-derived CHM: LOOCV R\u00B2 by Site",
       x = expression(R^2), y = NULL) +
  theme_minimal()
```




### Export Data
```{r}
outNAIP_estimates <- r'(S:\ursa\howell\Research\PJ_forests\08_ModelResults\01_NAIP)'
write.csv(final_NAIP_estimates, file = file.path(outNAIP_estimates, "/NAIP_estimates.csv"), append = TRUE)

outLidar_estimates <- r'(S:\ursa\howell\Research\PJ_forests\08_ModelResults\02_Lidar)'
write.csv(final_Lidar_estimates, file = file.path(outLidar_estimates, "/Lidar_estimates.csv"), append = TRUE)

outNAIPTreeCrowns <- r'(S:\ursa\howell\Research\PJ_forests\03_TreeData\01_TreeDetection\01_NAIP_Derived\Smoothed1)'
st_write(allTreeCrowns_NAIP_smoothed, file.path(outNAIPTreeCrowns, "TreeCrowns_NAIP.shp"), overwrite = TRUE)


```



