---
title: "Modeling Aboveground Biomass of Individual Trees in Pinyon-Juniper Woodlands"
author: "Jake Howell"
date: "`r Sys.Date()`"
purpose: 
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
---


```{r}
# Set output formats
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Script Purpose:
 Here we present our methods for estimating the aboveground biomass of individual trees within Pinyon-Juniper Woodlands using Using field data collected from 180 field plots, paired with two Canopy Height Model (CHM) sources.

Let's begin by getting things set up.

Be sure the required packages are installed and libraries are loaded.

```{r setup, include = FALSE}
# Check required packages are installed
pckgs <- c("sf",
           "dplyr",
           "sp",
           "raster",
           "terra",
           "lidR",
           "mapview",
           "units",
           "ggplot2",
           "broom",
           "purrr",
           "stringr",
           "tidyr",
           "readr")

# Point to packages
newPckgs <- pckgs[!(pckgs %in% installed.packages()[,"Package"])]

# If length is greater than 0, install the packages
if(length(newPckgs) > 0) install.packages(newPckgs)

# Load libraries
library(sf)
library(dplyr)
library(sp)
library(raster)
library(terra)
library(lidR)
library(mapview)
library(units)
library(ggplot2)
library(broom)
library(purrr)
library(stringr)
library(tidyr)
library(readr)

```

## Set up

We'll begin by setting the paths to our data. The required data are as follows 

- Lidar-Derived Canopy Height Models (.tif)
- NAIP-derived Canopy Height Models (.tif)
- Spreadsheet of field plot data (.csv)
- Spreadsheet of tree level data (.csv)
- Spreadsheet of AGB equations (.csv)

```{r}
# Set root data location
loc <- r'(S:\ursa\howell\Research\PJ_forests)'

# Set the name of the site data folder 
foldSite <- "01_Sites"

# Set the name of the plot level folder
foldPlot <- "02_Plots"

# Set the folder of NAIP-derived site-level CHMs
foldNAIP_CHM <- "01_NAIP_CHMs"

# Set the folder of the lidar-derived plot-level CHMs
foldLidar_CHM <- "03_LidarDerivedCHMs/Original"

# Set the name of the plot level .csv
namePlotData <- "plot_level_data_uofu.csv"

## TREE-LEVEL DATA: 
foldTree <- "03_TreeData"

# Set the name of the tree data
nameTreeData <- "tree_level_data.csv"

# Set name to AGB equations
nameAGB <- "all_tree_species_in_pj_fia_plots_new.csv"

# Initialize paths
pathNAIPsites <- file.path(loc, foldSite, foldNAIP_CHM)
pathLidarPlots <- file.path(loc, foldPlot, foldLidar_CHM)
pathPlotData <- file.path(loc, foldPlot, namePlotData)
pathTreeData <- file.path(loc, foldTree, nameTreeData)
pathAGBData <- file.path(loc, foldTree, nameAGB)

```

Now let's set our output paths
```{r}
# Path to export our clipped NAIP-derived CHMs
outClippedPlots_NAIP <- r'(S:\ursa\howell\Research\PJ_forests\02_Plots\01_NAIPDerivedCHMs)'

# Path to export our clipped lidar-derived CHMs
outClippedPlots_Lidar <- r'(S:\ursa\howell\Research\PJ_forests\02_Plots\03_LidarDerivedCHMs)'

# Path to export our lidar-derived tree crowns
# outTreeCrowns_Lidar <- 
# 
# # Path to export our lidar-derived tree points
# outTreePoints_Lidar <- 
#   
# # Path to export our NAIP-derived tree crowns
# outTreeCrowns_NAIP <-   
#   
# # Path to export our NAIP-derived tree points
# outTreePoints_NAIP <- 

```


## Read Data
With our input and output data paths set, let's now load in our data. Let's start with the plot-level, lidar-derived CHM's: 
```{r}
# List the lidar-derived plots within the folder
lidarPlotCHM_names <- list.files(pathLidarPlots, pattern = "\\.tif$", full.names = TRUE)

# Now read the rasters as a list
lidarPlotRasters <- lapply(lidarPlotCHM_names, rast)

```

Plot one to make sure it loaded correctly
```{r}
plot(lidarPlotRasters[[1]])

```

Now, read the site-level NAIP-derived CHMs:
```{r}
# List the NAIP-derived CHM sites
NAIPSitesCHM_names <- list.files(pathNAIPsites, pattern = "\\.tif$", full.names = TRUE)

# Now read the rasters as a list
NAIPSiteRasters <- lapply(NAIPSitesCHM_names, rast)

```

We won't plot the full NAIP CHM raster, but be sure there are 18 that loaded in

Finally, let's load in the rest of our data that we need:

```{r}
# Read plot field data
plotData <- read.csv(pathPlotData)

# Read tree data
treeData <- read.csv(pathTreeData)

# Read biomass equations
biomassEquations <- read.csv(pathAGBData)
```

The plotData table contains the unique plot identification and GPS boundaries of the 180 field plots that were collected during the field campaign
```{r}
head(plotData)
```

The treeData table contains all the unique trees that were collected throughout the 180 field plots. 
```{r}
head(treeData)
```

Finally, the biomassEquations table hosts species specific biomass equations for 158 tree species. We'll apply these equations to the individual trees that were collected throughout the field campaign, which we can then summarize to get accurate plot-level AGB estimates. 

## Apply Biomass Equations to Field Data
Let's apply the biomass equations. What we're doing is using AGB-allometry to apply the species specific biomass equations to every tree that was collected within the field campaign. 

```{r}
# In the tree data table, alter the species column
treeData <- treeData |>
  rename(Species = tree_sp)

# In the biomass equations table, alter the species column so it matches with the tree table
biomassEquations <- biomassEquations |>
  rename(Species = plants.code)

# Join the tree data to the biomass equations by Species column
treeData <- left_join(treeData, biomassEquations, by = "Species")

# Now estimate biomass for the field data trees by applying the biomass equations
treeData <- treeData |>
  mutate(tree_biomass_kg = exp(beta.0 + beta.1 * log(tree_diam)))

# Remove outliers for plotting purposes
treeData_filtered <- treeData |>
  filter(tree_diam < 100)

# Now plot the biomass of the field data trees and color by species
ggplot2::ggplot(treeData_filtered, aes(x = tree_diam, y = tree_biomass_kg, color = common.name)) +
  geom_point() +
  labs(
    title = "Species-Level AGB from Stem Diameter (DBH)",
    x = "Diameter at Breast Height (cm)", 
    y = "Aboveground Biomass (kg)",
    color = "Tree Species"
  ) +
  theme_minimal()

```

## Create Field Plot Extents
Now let's create our field plot extents. Our plotData table has the coordinates of the GPS points for our plot boundaries. We'll coerce that into a geometric figure using those points.

```{r}
# Create function to make one polygon the four sides of each field plot extent
createFieldPlotExtents <- function(row) {
  coords <- matrix(c(
    row$plot_ne_x, row$plot_ne_y,
    row$plot_se_x, row$plot_se_y,
    row$plot_sw_x, row$plot_sw_y,
    row$plot_nw_x, row$plot_nw_y,
    row$plot_ne_x, row$plot_ne_y
  ), ncol = 2, byrow = TRUE)
  st_polygon(list(coords))
}

# Go through each row of the plot data spreadsheet and create the field plot extents
fieldPlots_extents <- plotData |>
  rowwise() |>
  mutate(geometry = st_sfc(createFieldPlotExtents(pick(everything())))) |>
  ungroup()

# Set the current CRS and reproject
fieldPlots_extents <- st_as_sf(fieldPlots_extents, crs = 4326) |>
  st_transform(32612)

```

We now have field plot extents for each one of our 180 field plots. These boundaries will serve as the spatial boundaries for performing individual tree detection. 

## Sum the Plot-Level Biomass
Now we can use the field extents and the individual tree AGB estimates to get plot-level AGB estimates for each of the 180 field plots.

```{r}
# Return just the plot ID's from the field plot data
plot_ids <- unique(fieldPlots_extents$plot_id)

# Subset the tree data to just the unique plot names
treeDataInPlots_raw <- treeData[treeData$plot_id %in% plot_ids, ]

# FIX THE ERROR IN LAT/LONG ASSIGNMENT:
treeDataInPlots_fixed <- treeDataInPlots_raw %>%
  mutate(
    # start by assuming they are correct
    lon = tree_x,
    lat = tree_y,

    # identify likely swapped rows:
    # lon should be negative ~ -180..0 and lat should be positive ~ 0..90
    swapped = (tree_x > 0 & tree_x < 90) & (tree_y < 0 & tree_y > -180),

    # for swapped rows, flip them
    lon = if_else(swapped, tree_y, lon),
    lat = if_else(swapped, tree_x, lat)
  )

# Convert the treeDataInPlots to a spatial object
treeDataInPlots_fixed <- st_as_sf(treeDataInPlots_fixed,
                            coords = c("lon", "lat"), 
                            crs = 4326,
                            remove = FALSE)

# Now transform to UTM Zone 12N
treeDataInPlots <- st_transform(treeDataInPlots_fixed, 32612)

# Estimate Biomass Summary for each plot
plotLevelBiomass <- treeDataInPlots |>
  group_by(plot_id) |>
  summarise(
    total_biomass = sum(tree_biomass_kg, na.rm = TRUE),
    mean_biomass = mean(tree_biomass_kg, na.rm = TRUE),
    n_trees_GPS = n(),
    .groups = 'drop'
  )

```

Thus far, we've applied AGB equations to our field data trees and summarized those values at the plot-level to get total AGB for each of the 180 field plots. 

## Review Plot Data
Let's plot just the field data we collected for a single plot
```{r}
# Print plot names
# print(fieldPlots_extents$plot_id)

# Select a single plot 
plot <- fieldPlots_extents[fieldPlots_extents$plot_id == "UOFU_CODC_P03",]

# And get the trees within this plot
plot_trees <- treeDataInPlots[treeDataInPlots$plot_id == "UOFU_CODC_P03", ]

# Continuous palette for biomass
# biomass_pal <- colorNumeric(
#   palette = colorRampPalette(c("red", "yellow", "lightgreen"))(100),
#   domain = plot_trees$tree_biomass_kg,
#   na.color = "transparent"
# )

# Plot using mapview, but suppress the default legend
m <- mapview(
  plot,
  alpha.regions = 0.1,
  color = "white",
  lwd = 2,
  map.types = "Esri.WorldImagery"
) +
  mapview(
    plot_trees,
    zcol = "tree_biomass_kg",
    cex = "tree_biomass_kg",
    col.regions = colorRampPalette(c("red", "yellow", "lightgreen"))(80),
    alpha.regions = 1,
    legend = FALSE
  )

m@map <- m@map |>
  clearTiles() |>
  addProviderTiles(
    providers$Esri.WorldImagery,
    options = providerTileOptions(
      maxNativeZoom = 18,
      maxZoom = 22
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = biomass_pal,
    values = plot_trees$tree_biomass_kg,
    title = "Per Tree Biomass (kg)",
    opacity = 1,
    labFormat = labelFormat(
      digits = 1,
      suffix = " kg"
    )
  )

m


```


## Crop Canopy Height Models to Field Plot Extents: 
We'll now crop both sources of CHM data to match the field plot extents, which we'll subsequently use to run tree detection.

We'll loop through each of the 18 site-level NAIP-derived CHMs and clip them to the extent of the 180 field plots. We'll also be sure to apply a 10-m buffer to each plot extent to ensure we properly capture every tree crown that was collected from the field data

```{r}
# Store the results as a list
allPlotsCHMs_NAIP <- list()

# Loop through the list of 18 sites
for (i in seq_along(NAIPSiteRasters)) {

  # Create message
  # message("Processing site: ", siteNames[i])
  
  # Select a single site-level raster
  chm <- NAIPSiteRasters[[i]]

  # Create one polygon from raster extent
  chm_poly <- st_as_sf(as.polygons(ext(chm), crs = crs(chm)))

  # Ensure proper projection
  fieldPlots_extents <- st_transform(fieldPlots_extents, crs(chm_poly))

  # This returns a true / false matrix of whether or not a CHM intersects with the field plot extent
  idx <- st_intersects(fieldPlots_extents, chm_poly, sparse = FALSE)
  
  # Extracts the plot which intersects the CHM
  plot_rows <- which(idx)
  
  # Subsets the field plot extents to just those which intersect the site CHM
  plots_in_tile <- fieldPlots_extents[plot_rows, ]

  # If no plots intersect, skip to the next:
  if (nrow(plots_in_tile) == 0) next

  # Loops through each row of the field plot extent that intersected the site CHM
  for (j in 1:nrow(plots_in_tile)) {
  
    # Selects a single plot 
    plot_sf <- plots_in_tile[j, ]

    # Create a 10m buffer
    plot_sf_buf <- st_buffer(plot_sf, dist = 10)

    # Convert sf to a spatvector
    plot_vect <- vect(plot_sf_buf)

    # Store plot name
    plot_name <- plot_sf$plot_id

    # Clip the CHM
    clipped <- crop(chm, plot_vect) |>
      mask(plot_vect)

    # Add to list with plot name
    allPlotsCHMs_NAIP[[plot_name]] <- clipped

    # # Build output filename
    # out_name <- file.path(outClippedPlots_NAIP, paste0(plot_sf$plot_id, ".tif"))
    # 
    # # Now write the raster
    # writeRaster(clipped, out_name, overwrite = TRUE)
  }
}

# Save the names of the final NAIP plots
NAIP_PlotNames <- names(allPlotsCHMs_NAIP)

```

Repeat the process for the plot-level lidar-derived CHMs

```{r}
# Store the results as a list
allPlotCHMs_Lidar <- list()

# Reproject field plots once if you want, or inside loop per raster
for (i in seq_along(lidarPlotRasters)) {
  
  # Select a single site-level raster
  chm <- lidarPlotRasters[[i]]
  
  if (is.null(chm)) next
  
  # Create one polygon from raster extent
  chm_poly <- st_as_sf(as.polygons(ext(chm), crs = crs(chm)))
  
  # Reproject field plots to match this raster
  fieldPlots_proj <- st_transform(fieldPlots_extents, crs(chm_poly))
  
  # Which plots intersect this CHM extent
  idx <- st_intersects(fieldPlots_proj, chm_poly, sparse = FALSE)
  plot_rows <- which(idx[, 1])
  
  # Subset to intersecting plots
  plots_in_tile <- fieldPlots_proj[plot_rows, ]
  
  # Skip if none intersect
  if (nrow(plots_in_tile) == 0) next
  
  # Loop through intersecting plots
  for (j in seq_len(nrow(plots_in_tile))) {
    
    plot_sf <- plots_in_tile[j, ]
    
    # Make valid just in case
    plot_sf <- st_make_valid(plot_sf)
    
    # Create a 10m buffer 
    plot_sf_buf <- st_buffer(plot_sf, dist = 10)
    
    # Convert sf to SpatVector
    plot_vect <- vect(plot_sf_buf)
    
    # Extra safety check before crop
    if (!relate(ext(chm), ext(plot_vect), relation = "intersects")) {
      message("No raster overlap for plot: ", plot_sf$plot_id)
      next
    }
    
    # Store plot name
    plot_name <- plot_sf$plot_id
    
    # Clip the CHM
    clipped <- tryCatch(
      crop(chm, plot_vect) |> mask(plot_vect),
      error = function(e) {
        message("Crop/mask failed for plot ", plot_name, ": ", e$message)
        NULL
      }
    )
    
    if (is.null(clipped)) next
    
    # Add to list with plot name
    allPlotCHMs_Lidar[[plot_name]] <- clipped
    
    # # Build output filename
    # out_name <- file.path(outClippedPlots_Lidar, paste0("lidar_", plot_name, ".tif"))
    # 
    # # Write the raster
    # writeRaster(clipped, out_name, overwrite = TRUE)
  }
}

# Save the names
Lidar_PlotNames <- names(allPlotCHMs_Lidar)

```

Now that we have CHM's, we'll perform individual tree detection for both the NAIP-derived and Lidar-derived CHMs. 

The chunk below is for tuning the tree detection and segmentation parameters for a single plot for the NAIP-derived CHM. 

```{r}
## PARAM 1: Variable window size 
h <- function(x) {
  cw <- 1.0827 * x - 0.3993
  ws <- pmin(10, pmax(3, cw/2))
  return(ws)
}

## PARAM 2: Max pixel height for crown segmentation
# The default is 2
# Threshold below which a pixel cannot be a tree
th_tree_NAIP  = 0.5

## PARAM 3: Growing Threshold 1: 
# A pixel is added to a region if its height is greater than the tree height multiplied by this value
# The default is = 0.45
# At 0.45, pixels must be at least 45% as tall as the tree top to be included in the crown
th_seed_NAIP  = 0.35

# PARAM 4: Growing Threshold 2: 
# A pixel is added to a region if its height is greater than the current mean height of the region multiplied by this value
# The default is = 0.55
# Low values = looser crowns, possible merging
# High values = stricter boundaries, possible over-segmentation
th_cr_NAIP    = 0.6

# PARAM 5: Maximum value of the crown diameter.
# DEFAULT = 10
max_cr_NAIP = 15

### END USER INPUT -------------------------------------------------------------
# Select the CHM name you want to test tree detection on
chm_name <- names(allPlotsCHMs_NAIP)[4]

# Run Tree Detection on a single NAIP CHM
chm_test <- allPlotsCHMs_NAIP[[chm_name]]

# Reproject
chm_test <- project(chm_test, "EPSG:32612")
    
# Ensure proper projection
chm_test <- terra::project(chm_test, "EPSG:32612")
    
# Identify Tree Tops
ttops <- tryCatch(
      locate_trees(chm_test, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm_test,
        treetops = ttops_sf,
        th_tree  = th_tree_NAIP,
        th_seed  = th_seed_NAIP,
        th_cr    = th_cr_NAIP,
        max_cr   = max_cr_NAIP,
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      nextppp
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
# Store names as treeID
names(crowns_poly) <- "treeID"

# Save an original crown polygon segmentation
crowns_poly_original <- crowns_poly

# Return the area of each crown polygon
crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")


# Set small crown threshold
small_thresh <- 3 * res(chm_test)[1]^2   # Returns 1.07 m2
small_thresh <- 4 * res(chm_test)[1]^2

# Coerce the crowns into small/large classes
small_crowns <- crowns_poly[crowns_poly$area_m2 <= small_thresh, ]
big_crowns   <- crowns_poly[crowns_poly$area_m2 > small_thresh, ]

# Only perform segmentation on large crowns
big_crowns <- buffer(big_crowns, -1 * res(chm_test)[1]) |>
  buffer(1 * res(chm_test)[1]) |>
  simplifyGeom(res(chm_test)[1] / 4)


crowns_poly <- rbind(big_crowns, small_crowns)
crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

# Drop empty / invalid geometries
crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

if (nrow(crowns_poly) == 0) return(NULL)

# Calculate Crown Area in m2
crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

# Extract the maximum height for each crown polygon
zmax <- terra::extract(chm_test, crowns_poly, fun = max, na.rm = TRUE)
  
# Assign the maximum height to the crown polygons
crowns_poly$tree_height_m <- zmax[,2]

# Convert to sf 
crowns_sf <- sf::st_as_sf(crowns_poly)

# Transform
ttops_sf <- st_transform(ttops_sf, st_crs("EPSG:32612"))
crowns_sf <- st_transform(crowns_sf, st_crs("EPSG:32612"))
fieldPlots_extents <- st_transform(fieldPlots_extents, st_crs("EPSG:32612"))

# Add plot ID
crowns_sf$plotID <- chm_name
ttops_sf$plotID <- chm_name

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- ttops_sf[
  lengths(st_within(ttops_sf, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
ttops_sf <- ttops_sf %>%
  inner_join(keys, by = c("plotID", "treeID"))

## ---------------- OPTIONAL METRICS ----------------

crowns_vect <- terra::vect(crowns_sf)

crowns_sf$area_m2 <- terra::expanse(crowns_vect, unit = "m")
zmax <- terra::extract(chm_test, crowns_vect, fun = max, na.rm = TRUE)
crowns_sf$tree_height_m <- zmax[, 2]

## ---------------- FILTER TO PLOT USING TREETOPS ----------------

ttops_sf$plotID  <- chm_name
crowns_sf$plotID <- chm_name

# Keep only treetops inside the chosen plot
ttops_in_plot <- ttops_sf[
  lengths(st_within(ttops_sf, plot_extent)) > 0,
]

if (nrow(ttops_in_plot) == 0) {
  stop("No treetops fall inside the selected plot boundary: ", chm_name)
}

keys <- ttops_in_plot |>
  st_drop_geometry() |>
  distinct(plotID, treeID)

ttops_sf <- ttops_sf |>
  semi_join(keys, by = c("plotID", "treeID"))

crowns_sf <- crowns_sf |>
  semi_join(keys, by = c("plotID", "treeID"))

if (nrow(crowns_sf) == 0) {
  stop("No crowns remained after filtering to plot boundary.")
}

## ---------------- MAPVIEW PREP ----------------

# Avoid FlatGeobuf errors
mapviewOptions(fgb = FALSE)

ttops_sf    <- drop_zm_safe(ttops_sf)
crowns_sf   <- drop_zm_safe(crowns_sf)
plot_extent <- drop_zm_safe(plot_extent)

## ---------------- MAPVIEW ----------------

m <-
  mapview(
    chm_test,
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))(100),
    alpha = 1,
    legend = FALSE
  ) +
  mapview(
    plot_extent,
    color = "blue",
    alpha.regions = 0,
    lwd = 3,
    legend = FALSE
  ) +
  mapview(
    crowns_sf,
    color = "black",
    alpha.regions = 0,
    lwd = 1,
    legend = FALSE
  ) +
  mapview(
    ttops_sf,
    col.regions = "white",
    cex = 4,
    legend = FALSE
  )

m



```

```{r}
library(terra)
library(sf)
library(dplyr)
library(lidR)
library(mapview)

## ---------------- USER INPUTS ----------------

# Variable window size function
h <- function(x) {
  cw <- 1.0827 * x - 0.3993
  ws <- pmin(10, pmax(3, cw / 2))
  return(ws)
}

# Dalponte parameters for NAIP CHM
th_tree_NAIP <- 0.5
th_seed_NAIP <- 0.35
th_cr_NAIP   <- 0.6
max_cr_NAIP  <- 15

# Plot to run
chm_name <- "UOFU_AZCV_P04"

# Smoothing settings
smooth_dist_mult  <- 0.5   # inward/outward smoothing distance in pixels
small_thresh_mult <- 6     # crowns <= this many pixels of area are not smoothed

## ---------------- HELPERS ----------------

drop_zm_safe <- function(x) {
  tryCatch(
    st_zm(x, drop = TRUE, what = "ZM"),
    error = function(e) x
  )
}

safe_smooth_geom <- function(g, dist_m, tol_m) {
  g2 <- tryCatch(
    g |>
      st_buffer(-dist_m) |>
      st_buffer(dist_m) |>
      st_simplify(dTolerance = tol_m, preserveTopology = TRUE) |>
      st_make_valid(),
    error = function(e) g
  )

  if (length(g2) == 0 || isTRUE(st_is_empty(g2))) {
    return(g)
  }

  g2
}

## ---------------- LOAD DATA ----------------

chm_test <- allPlotCHMs_NAIP[[chm_name]]
if (is.null(chm_test)) stop("CHM not found for: ", chm_name)

# Project CHM once
chm_test <- terra::project(chm_test, "EPSG:32612")
pix <- res(chm_test)[1]

# Match field plot boundary CRS
fieldPlots_extents <- st_transform(fieldPlots_extents, 32612)

# Change plot_id below if your field plot object uses a different column name
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

if (nrow(plot_extent) == 0) {
  stop("No matching plot boundary found for: ", chm_name)
}

## ---------------- TREE DETECTION ----------------

ttops <- tryCatch(
  locate_trees(chm_test, lmf(ws = h, hmin = 0.5)),
  error = function(e) {
    message("Treetop detection failed: ", e$message)
    NULL
  }
)

if (is.null(ttops) ||
    (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
    (inherits(ttops, "sf") && nrow(ttops) == 0) ||
    (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
  stop("No treetops detected for plot: ", chm_name)
}

ttops_sf <- st_as_sf(ttops) |>
  st_transform(32612)

if (nrow(ttops_sf) == 0) {
  stop("No valid treetop sf object for plot: ", chm_name)
}

## ---------------- CROWN SEGMENTATION ----------------

crowns_id <- tryCatch(
  dalponte2016(
    chm      = chm_test,
    treetops = ttops_sf,
    th_tree  = th_tree_NAIP,
    th_seed  = th_seed_NAIP,
    th_cr    = th_cr_NAIP,
    max_cr   = max_cr_NAIP
  )(),
  error = function(e) {
    message("Dalponte segmentation failed: ", e$message)
    NULL
  }
)

if (is.null(crowns_id)) {
  stop("No crown raster returned for plot: ", chm_name)
}

crowns_poly <- tryCatch(
  terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
  error = function(e) {
    message("as.polygons failed: ", e$message)
    NULL
  }
)

if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
  stop("No crown polygons returned for plot: ", chm_name)
}

names(crowns_poly) <- "treeID"
crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

if (terra::nrow(crowns_poly) == 0) {
  stop("All crown geometries were empty for plot: ", chm_name)
}

crowns_sf <- st_as_sf(crowns_poly) |>
  st_transform(32612) |>
  st_make_valid()

## ---------------- SMOOTH ONLY LARGE CROWNS ----------------

crowns_sf$area_m2 <- as.numeric(st_area(crowns_sf))

small_thresh <- small_thresh_mult * pix^2
smooth_dist  <- smooth_dist_mult * pix
smooth_tol   <- pix / 4

small_sf <- crowns_sf[crowns_sf$area_m2 <= small_thresh, ]
big_sf   <- crowns_sf[crowns_sf$area_m2 >  small_thresh, ]

if (nrow(big_sf) > 0) {
  new_geoms <- lapply(
    st_geometry(big_sf),
    safe_smooth_geom,
    dist_m = smooth_dist,
    tol_m  = smooth_tol
  )

  st_geometry(big_sf) <- st_sfc(new_geoms, crs = st_crs(big_sf))
  big_sf <- big_sf[!st_is_empty(big_sf), ]
}

crowns_sf <- do.call(
  rbind,
  Filter(function(x) !is.null(x) && nrow(x) > 0, list(big_sf, small_sf))
)

if (is.null(crowns_sf) || nrow(crowns_sf) == 0) {
  stop("No crowns remained after smoothing.")
}

## ---------------- OPTIONAL METRICS ----------------

crowns_vect <- terra::vect(crowns_sf)

crowns_sf$area_m2 <- terra::expanse(crowns_vect, unit = "m")
zmax <- terra::extract(chm_test, crowns_vect, fun = max, na.rm = TRUE)
crowns_sf$tree_height_m <- zmax[, 2]

## ---------------- FILTER TO PLOT USING TREETOPS ----------------

ttops_sf$plotID  <- chm_name
crowns_sf$plotID <- chm_name

ttops_in_plot <- ttops_sf[
  lengths(st_within(ttops_sf, plot_extent)) > 0,
]

if (nrow(ttops_in_plot) == 0) {
  stop("No treetops fall inside the selected plot boundary: ", chm_name)
}

keys <- ttops_in_plot |>
  st_drop_geometry() |>
  distinct(plotID, treeID)

ttops_sf <- ttops_sf |>
  semi_join(keys, by = c("plotID", "treeID"))

crowns_sf <- crowns_sf |>
  semi_join(keys, by = c("plotID", "treeID"))

if (nrow(crowns_sf) == 0) {
  stop("No crowns remained after filtering to plot boundary.")
}

## ---------------- MAPVIEW PREP ----------------

mapviewOptions(fgb = FALSE)

ttops_sf    <- drop_zm_safe(ttops_sf)
crowns_sf   <- drop_zm_safe(crowns_sf)
plot_extent <- drop_zm_safe(plot_extent)

## ---------------- MAPVIEW ----------------

m <-
  mapview(
    chm_test,
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))(100),
    alpha = 1,
    legend = FALSE
  ) +
  mapview(
    plot_extent,
    color = "blue",
    alpha.regions = 0,
    lwd = 3,
    legend = FALSE
  ) +
  mapview(
    crowns_sf,
    color = "black",
    alpha.regions = 0,
    lwd = 1,
    legend = FALSE
  ) +
  mapview(
    ttops_sf,
    col.regions = "white",
    cex = 4,
    legend = FALSE
  )

m


```

Now, tune the parameters on the lidar-derived tree list

```{r}

## PARAM 1: Variable window size 
h <- function(x) {
  cw <- 1.0827 * x - 0.3993
  ws <- pmin(10, pmax(3, cw/2))
  return(ws)
}

## PARAM 2: Max pixel height for crown segmentation
# The default is 2
# Threshold below which a pixel cannot be a tree
th_tree_lidar  = 0.5

## PARAM 3: Growing Threshold 1: 
# A pixel is added to a region if its height is greater than the tree height multiplied by this value
# The default is = 0.45
# At 0.45, pixels must be at least 45% as tall as the tree top to be included in the crown
th_seed_lidar  = 0.35

# PARAM 4: Growing Threshold 2: 
# A pixel is added to a region if its height is greater than the current mean height of the region multiplied by this value
# The default is = 0.55
# Low values = looser crowns, possible merging
# High values = stricter boundaries, possible over-segmentation
th_cr_lidar    = 0.6

# PARAM 5: Maximum value of the crown diameter.
# DEFAULT = 10
max_cr_lidar = 15

### END USER INPUT -------------------------------------------------------------
# Select the CHM name you want to test tree detection on
chm_name <- "UOFU_AZCV_P01"

# Run Tree Detection on a single NAIP CHM
chm_test <- allPlotCHMs_Lidar[[chm_name]]

# Reproject
chm_test <- project(chm_test, "EPSG:32612")
    
# Ensure proper projection
chm_test <- terra::project(chm_test, "EPSG:32612")
    
# Identify Tree Tops
ttops <- tryCatch(
      locate_trees(chm_test, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm_test,
        treetops = ttops_sf,
        th_tree  = th_tree_lidar,
        th_seed  = th_seed_lidar,
        th_cr    = th_cr_lidar,
        max_cr   = max_cr_lidar
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
# Store names as treeID
names(crowns_poly) <- "treeID"

# Save an original crown polygon segmentation
crowns_poly_original <- crowns_poly

# Return the area of each crown polygon
crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

# Set small crown threshold
small_thresh <- 3 * res(chm_test)[1]^2   # Returns 1.07 m2
# small_thresh <- 4 * res(chm_test)[1]^2

# Coerce the crowns into small/large classes
small_crowns <- crowns_poly[crowns_poly$area_m2 <= small_thresh, ]
big_crowns   <- crowns_poly[crowns_poly$area_m2 > small_thresh, ]

# Only perform segmentation on large crowns
big_crowns <- buffer(big_crowns, -1 * res(chm_test)[1]) |>
  buffer(1 * res(chm_test)[1]) |>
  simplifyGeom(res(chm_test)[1] / 4)


crowns_poly <- rbind(big_crowns, small_crowns)
crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

# Drop empty / invalid geometries
crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

if (nrow(crowns_poly) == 0) return(NULL)

# Calculate Crown Area in m2
crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

# Extract the maximum height for each crown polygon
zmax <- terra::extract(chm_test, crowns_poly, fun = max, na.rm = TRUE)
  
# Assign the maximum height to the crown polygons
crowns_poly$tree_height_m <- zmax[,2]

# Convert to sf 
crowns_sf <- sf::st_as_sf(crowns_poly)
plot(crowns_sf)

# Transform
ttops_sf <- st_transform(ttops_sf, st_crs("EPSG:32612"))
crowns_sf <- st_transform(crowns_sf, st_crs("EPSG:32612"))
fieldPlots_extents <- st_transform(fieldPlots_extents, st_crs("EPSG:32612"))

# Add plot ID
crowns_sf$plotID <- chm_name
ttops_sf$plotID <- chm_name

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- ttops_sf[
  lengths(st_within(ttops_sf, fieldPlots_extents)) > 0,
]

# Filter only the tree tops which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents


# Subset to only the crowns we want to keep
ttops_sf <- ttops_sf %>%
  inner_join(keys, by = c("plotID", "treeID"))

crowns_sf <- crowns_sf %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Get the field plot extent
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

# Plot the results
plot(chm_test)
plot(ttops_sf,
     col = "white",
     lwd = 4, 
     add = TRUE)
plot(st_geometry(crowns_sf), 
     add = TRUE, 
     fill = NA, 
     border = "red",
     lwd = 2)
plot(st_geometry(plot_extent),
    add = TRUE,
    border = "red",
    lwd = 2)

plot(crowns_poly)
  plot(crowns_poly_original)
  plot(big_crowns)
  plot(crowns_sf$geometry)

# Mapview it
  # Convert
  crowns_sf <- st_as_sf(crowns_sf)
  crowns_sf <- st_zm(crowns_sf, drop = TRUE, what = "ZM")
  
  ttops_sf <- st_as_sf(ttops_sf)
  ttops_sf <- st_zm(ttops_sf, drop = TRUE, what = "ZM")
  
m <- 
#   plot_extent,
#   alpha.regions = 0.1,
#   color = "white",
#   lwd = 2,
#   map.types = "Esri.WorldImagery"
# ) +
  mapview(chm_test, 
          col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))(100),
          alpha = 1, 
          legend = FALSE) +
  # mapview(
  #   ttops_sf,
  #   col.regions = colorRampPalette(c("lightgreen", "yellow", "red"))(20),
  #   alpha.regions = 1,
  #   legend = FALSE
  # ) +
  mapview(
    crowns_sf,
    color = "black",
    alpha.regions = 0,
    lwd = 2,
    legend = FALSE
)
  
m

```

```{r}
## ---------------- USER INPUTS ----------------

# Variable window size function
h <- function(x) {
  cw <- 1.0827 * x - 0.3993
  ws <- pmin(10, pmax(3, cw / 2))
  return(ws)
}

# Dalponte parameters
th_tree_lidar <- 0.5
th_seed_lidar <- 0.35
th_cr_lidar   <- 0.6
max_cr_lidar  <- 15

# Plot to run
chm_name <- "UOFU_AZCV_P04"

# Safer smoothing settings
smooth_dist_mult <- 0.5   # inward/outward smoothing distance in pixels
small_thresh_mult <- 6    # crowns <= this many pixels of area are not smoothed

## ---------------- HELPERS ----------------

drop_zm_safe <- function(x) {
  tryCatch(
    st_zm(x, drop = TRUE, what = "ZM"),
    error = function(e) x
  )
}

safe_smooth_geom <- function(g, dist_m, tol_m) {
  g2 <- tryCatch(
    g |>
      st_buffer(-dist_m) |>
      st_buffer(dist_m) |>
      st_simplify(dTolerance = tol_m, preserveTopology = TRUE) |>
      st_make_valid(),
    error = function(e) g
  )

  if (length(g2) == 0 || isTRUE(st_is_empty(g2))) {
    return(g)
  }

  g2
}

## ---------------- LOAD DATA ----------------

chm_test <- allPlotCHMs_Lidar[[chm_name]]
if (is.null(chm_test)) stop("CHM not found for: ", chm_name)

# Project CHM once
chm_test <- terra::project(chm_test, "EPSG:32612")
pix <- res(chm_test)[1]

# Match field plot boundary CRS
fieldPlots_extents <- st_transform(fieldPlots_extents, 32612)

# If your column is plotID instead of plot_id, change this line
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

if (nrow(plot_extent) == 0) {
  stop("No matching plot boundary found for: ", chm_name)
}

## ---------------- TREE DETECTION ----------------

ttops <- tryCatch(
  locate_trees(chm_test, lmf(ws = h, hmin = 0.5)),
  error = function(e) {
    message("Treetop detection failed: ", e$message)
    NULL
  }
)

if (is.null(ttops) ||
    (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
    (inherits(ttops, "sf") && nrow(ttops) == 0) ||
    (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
  stop("No treetops detected for plot: ", chm_name)
}

ttops_sf <- st_as_sf(ttops) |>
  st_transform(32612)

if (nrow(ttops_sf) == 0) {
  stop("No valid treetop sf object for plot: ", chm_name)
}

## ---------------- CROWN SEGMENTATION ----------------

crowns_id <- tryCatch(
  dalponte2016(
    chm      = chm_test,
    treetops = ttops_sf,
    th_tree  = th_tree_lidar,
    th_seed  = th_seed_lidar,
    th_cr    = th_cr_lidar,
    max_cr   = max_cr_lidar
  )(),
  error = function(e) {
    message("Dalponte segmentation failed: ", e$message)
    NULL
  }
)

if (is.null(crowns_id)) {
  stop("No crown raster returned for plot: ", chm_name)
}

crowns_poly <- tryCatch(
  terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
  error = function(e) {
    message("as.polygons failed: ", e$message)
    NULL
  }
)

if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
  stop("No crown polygons returned for plot: ", chm_name)
}

names(crowns_poly) <- "treeID"
crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

if (terra::nrow(crowns_poly) == 0) {
  stop("All crown geometries were empty for plot: ", chm_name)
}

crowns_sf <- st_as_sf(crowns_poly) |>
  st_transform(32612) |>
  st_make_valid()

## ---------------- SMOOTH ONLY LARGE CROWNS ----------------

# Initial area for smoothing split
crowns_sf$area_m2 <- as.numeric(st_area(crowns_sf))

small_thresh <- small_thresh_mult * pix^2
smooth_dist  <- smooth_dist_mult * pix
smooth_tol   <- pix / 4

small_sf <- crowns_sf[crowns_sf$area_m2 <= small_thresh, ]
big_sf   <- crowns_sf[crowns_sf$area_m2 >  small_thresh, ]

if (nrow(big_sf) > 0) {
  new_geoms <- lapply(
    st_geometry(big_sf),
    safe_smooth_geom,
    dist_m = smooth_dist,
    tol_m  = smooth_tol
  )

  st_geometry(big_sf) <- st_sfc(new_geoms, crs = st_crs(big_sf))
  big_sf <- big_sf[!st_is_empty(big_sf), ]
}

crowns_sf <- do.call(
  rbind,
  Filter(function(x) !is.null(x) && nrow(x) > 0, list(big_sf, small_sf))
)

if (is.null(crowns_sf) || nrow(crowns_sf) == 0) {
  stop("No crowns remained after smoothing.")
}

## ---------------- OPTIONAL METRICS ----------------

crowns_vect <- terra::vect(crowns_sf)

crowns_sf$area_m2 <- terra::expanse(crowns_vect, unit = "m")
zmax <- terra::extract(chm_test, crowns_vect, fun = max, na.rm = TRUE)
crowns_sf$tree_height_m <- zmax[, 2]

## ---------------- FILTER TO PLOT USING TREETOPS ----------------

ttops_sf$plotID  <- chm_name
crowns_sf$plotID <- chm_name

# Keep only treetops inside the chosen plot
ttops_in_plot <- ttops_sf[
  lengths(st_within(ttops_sf, plot_extent)) > 0,
]

if (nrow(ttops_in_plot) == 0) {
  stop("No treetops fall inside the selected plot boundary: ", chm_name)
}

keys <- ttops_in_plot |>
  st_drop_geometry() |>
  distinct(plotID, treeID)

ttops_sf <- ttops_sf |>
  semi_join(keys, by = c("plotID", "treeID"))

crowns_sf <- crowns_sf |>
  semi_join(keys, by = c("plotID", "treeID"))

if (nrow(crowns_sf) == 0) {
  stop("No crowns remained after filtering to plot boundary.")
}

## ---------------- MAPVIEW PREP ----------------

# Avoid FlatGeobuf errors
mapviewOptions(fgb = FALSE)

ttops_sf    <- drop_zm_safe(ttops_sf)
crowns_sf   <- drop_zm_safe(crowns_sf)
plot_extent <- drop_zm_safe(plot_extent)

## ---------------- MAPVIEW ----------------

m <-
  mapview(
    chm_test,
    col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))(100),
    alpha = 1,
    legend = FALSE
  ) +
  mapview(
    plot_extent,
    color = "blue",
    alpha.regions = 0,
    lwd = 3,
    legend = FALSE
  ) +
  mapview(
    crowns_sf,
    color = "black",
    alpha.regions = 0,
    lwd = 1,
    legend = FALSE
  ) +
  mapview(
    ttops_sf,
    col.regions = "white",
    cex = 4,
    legend = FALSE
  )

m
```

Once satisfied for parameters, run individual tree detection across the 180 plots for the NAIP-derived Canopy Height Models:

```{r}
# Store tree tops and crowns as a list
NAIP_treePoints_list <- list()
NAIP_treeCrowns_list <- list()

# Loop through each plot CHM and run tree detection
for (plot_id in NAIP_PlotNames) {

  
  message("Processing plot: ", plot_id)

   # Select a single plot CHM
    chm <- allPlotCHMs_Lidar[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    # Ensure proper projection
    chm <- terra::project(chm, "EPSG:32612")
    
    # Identify Tree Tops
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree_NAIP,
        th_seed  = th_seed_NAIP,
        th_cr    = th_cr_NAIP,
        max_cr   = max_cr_NAIP
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    # Store names as treeID
    names(crowns_poly) <- "treeID"
  
  # Clean up the oblong crown shapes by buffering in and out, then simplifying
  crowns_poly <- buffer(crowns_poly, -2 * res(chm_test)[1]) |>
    buffer(2 * res(chm_test)[1]) |>
    simplifyGeom(res(chm_test)[1] / 4)

  # Drop empty / invalid geometries
  crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

if (nrow(crowns_poly) == 0) return(NULL)

  # Calculate Crown Area in m2
  crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

  # Extract the maximum height for each crown polygon
  zmax <- tryCatch(
    terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
    error = function(e) {
      message("extract failed: ", e$message)
      NULL
    }
  )

  if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
    message("Height extraction row mismatch for plot: ", plot_id,
            " | crowns=", terra::nrow(crowns_poly),
            " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax)))
    next
  }

  # Assign the maximum height to the crown polygons
  crowns_poly$tree_height_m <- zmax[,2]

  # Convert to sf
  crowns_sf <- sf::st_as_sf(crowns_poly)

  # Add plot ID
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID <- plot_id

  # Store
  NAIP_treePoints_list[[plot_id]] <- ttops_sf
  NAIP_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Bind all rows
allTreePoints_NAIP <- dplyr::bind_rows(NAIP_treePoints_list)
allTreeCrowns_NAIP <- dplyr::bind_rows(NAIP_treeCrowns_list)

# Check projection
allTreePoints_NAIP <- st_transform(allTreePoints_NAIP, st_crs(fieldPlots_extents))
allTreeCrowns_NAIP <- st_transform(allTreeCrowns_NAIP, st_crs(fieldPlots_extents))

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- allTreePoints_NAIP[
  lengths(st_within(allTreePoints_NAIP, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
allTreeCrowns_NAIP_final <- allTreeCrowns_NAIP %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Subset to only the tree tops we want to keep
allTreePoints_NAIP_final <- allTreePoints_NAIP %>%
  inner_join(keys, by = c("plotID", "treeID"))

```


Then, run tree detection on the lidar-derived CHMs

```{r}
# Store tree tops and crowns as a list
Lidar_treePoints_list <- list()
Lidar_treeCrowns_list <- list()

# Loop through each plot CHM and run tree detection
for (plot_id in Lidar_PlotNames) {

  message("Processing plot: ", plot_id)
  
   # Select a single plot CHM
    chm <- allPlotCHMs_Lidar[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    # Ensure proper projection
    chm <- terra::project(chm, "EPSG:32612")
    
    # Identify Tree Tops
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    # Skip if no treetops
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    # Convert to sf object
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    # Run Dalponte2016 tree segmentation algorithm
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree_lidar,
        th_seed  = th_seed_lidar,
        th_cr    = th_cr_lidar,
        max_cr   = max_cr_lidar
      )(),
      error = function(e) {
        message("Dalponte CHM-only failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    # Convert crown segmentation raster to polygons
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    # Store names as treeID
    names(crowns_poly) <- "treeID"  
    
    
  # # Select a single plot CHM
  # chm <- allPlotCHMs_Lidar[[plot_id]]
  # 
  # if (is.null(chm)) {
  #   message("CHM is NULL for plot: ", plot_id)
  #   next
  # }

  # # Ensure proper projection
  # chm <- terra::project(chm, "EPSG:32612")
  # 
  # # Identify Tree Tops
  # ttops <- tryCatch(
  #   locate_trees(chm, lmf(ws = h, hmin = 0.5)),
  #   error = function(e) {
  #     message("Treetops failed: ", e$message)
  #     NULL
  #   }
  # )
  # 
  # # Skip if no treetops
  # if (is.null(ttops) ||
  #     (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
  #     (inherits(ttops, "sf") && nrow(ttops) == 0) ||
  #     (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
  #   message("No treetops detected for plot: ", plot_id)
  #   next
  # }
  # 
  # # Convert to sf object
  # ttops_sf <- tryCatch(
  #   st_as_sf(ttops),
  #   error = function(e) {
  #     message("st_as_sf failed: ", e$message)
  #     NULL
  #   }
  # )
  # 
  # if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
  #   message("No valid treetop sf object for plot: ", plot_id)
  #   next
  # }
  # 
  # # Run Dalponte2016 tree segmentation algorithm
  # crowns_id <- tryCatch(
  #   dalponte2016(
  #     chm      = chm,
  #     treetops = ttops_sf,
  #     th_tree  = th_tree_lidar,
  #     th_seed  = th_seed_lidar,
  #     th_cr    = th_cr_lidar,
  #     max_cr   = max_cr_lidar
  #   )(),
  #   error = function(e) {
  #     message("Dalponte CHM-only failed: ", e$message)
  #     NULL
  #   }
  # )
  # 
  # if (is.null(crowns_id)) {
  #   message("No crown raster returned for plot: ", plot_id)
  #   next
  # }
  # 
  # # Convert crown segmentation raster to a SpatRaster
  # crowns_id <- tryCatch(
  #   terra::rast(crowns_id),
  #   error = function(e) { message("Conversion to SpatRaster failed: ", e$message); NULL }
  # )
  # 
  # 
  # #  Now convert SpatRaster to polygon
  # crowns_poly <- tryCatch(
  #   terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
  #   error = function(e) { message("as.polygons failed: ", e$message); NULL }
  # )
  # 
  # if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
  #   message("No crown polygons for plot: ", plot_id)
  #   next
  # }
  # 
  # # Store names as treeID
  # names(crowns_poly) <- "treeID"

  # Clean up the oblong crown shapes by buffering in and out, then simplifying
  crowns_poly <- buffer(crowns_poly, -2 * res(chm_test)[1]) |>
    buffer(2 * res(chm_test)[1]) |>
    simplifyGeom(res(chm_test)[1] / 4)

  # Drop empty / invalid geometries
  crowns_poly <- crowns_poly[!terra::is.empty(crowns_poly), ]

if (nrow(crowns_poly) == 0) return(NULL)

  # Calculate Crown Area in m2
  crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")

  # Extract the maximum height for each crown polygon
  zmax <- tryCatch(
    terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
    error = function(e) {
      message("extract failed: ", e$message)
      NULL
    }
  )

  if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
    message("Height extraction row mismatch for plot: ", plot_id,
            " | crowns=", terra::nrow(crowns_poly),
            " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax)))
    next
  }

  # Assign the maximum height to the crown polygons
  crowns_poly$tree_height_m <- zmax[,2]

  # Convert to sf 
  crowns_sf <- sf::st_as_sf(crowns_poly)

  # Add plot ID
  crowns_sf$plotID <- plot_id
  ttops_sf$plotID <- plot_id

  # Store
  Lidar_treePoints_list[[plot_id]] <- ttops_sf
  Lidar_treeCrowns_list[[plot_id]] <- crowns_sf
}

# Bind all rows
allTreePoints_Lidar <- dplyr::bind_rows(Lidar_treePoints_list)
allTreeCrowns_Lidar <- dplyr::bind_rows(Lidar_treeCrowns_list)

# Check projection
allTreePoints_Lidar <- st_transform(allTreePoints_Lidar, st_crs(fieldPlots_extents))
allTreeCrowns_Lidar <- st_transform(allTreeCrowns_Lidar, st_crs(fieldPlots_extents))

# Keep only the tree points that fall within the field plot extents
ttops_in_plots <- allTreePoints_Lidar[
  lengths(st_within(allTreePoints_Lidar, fieldPlots_extents)) > 0,
]

# Filter only the tree crowns which have an associated tree top that falls within the field plot extents
keys <- ttops_in_plots %>%
  st_drop_geometry() %>%
  distinct(plotID, treeID)

# Subset to only the crowns we want to keep
allTreeCrowns_Lidar_final <- allTreeCrowns_Lidar %>%
  inner_join(keys, by = c("plotID", "treeID"))

# Subset to only the tree tops we want to keep
allTreePoints_Lidar_final <- allTreePoints_Lidar %>%
  inner_join(keys, by = c("plotID", "treeID"))

```

We've now detected tree tops and delineated tree crowns for both the NAIP-derived and Lidar-Derived CHM.

Let's check the results for a single NAIP-derived CHM and its corresponding Lidar-derived CHM, along with the field data for the same plot.

```{r}
# What's the name of the CHM you want to check?
chm_name <- "UOFU_AZCV_P04"

# Get the single NAIP CHM
chm_NAIP <- allPlotsCHMs_NAIP[[chm_name]]

# Get the single Lidar CHM
chm_Lidar <- allPlotCHMs_Lidar[[chm_name]]

# Subset the NAIP tree points to selected plot
treePoints_NAIP <- allTreePoints_NAIP_final |> 
  filter(plotID == chm_name)
treePoints_NAIP <- st_as_sf(treePoints_NAIP)
treePoints_NAIP <- st_zm(treePoints_NAIP, drop = TRUE, what = "ZM")

# Subset the NAIP tree crowns to the selected plot
treeCrowns_NAIP <- allTreeCrowns_NAIP_final |>
  filter(plotID == chm_name)
treeCrowns_NAIP <- st_as_sf(treeCrowns_NAIP)
treeCrowns_NAIP <- st_zm(treeCrowns_NAIP, drop = TRUE, what = "ZM")

# Subset the Lidar tree points to the selected plot
treePoints_Lidar <- allTreePoints_Lidar_final |> 
  filter(plotID == chm_name)
treePoints_Lidar <- st_as_sf(treePoints_Lidar)
treePoints_Lidar <- st_zm(treePoints_Lidar, drop = TRUE, what = "ZM")



# Subset the lidar tree crowns to the selected plot
treeCrowns_Lidar <- allTreeCrowns_Lidar_final |>
  filter(plotID == chm_name)

# Get the GPS tree points for the plot
fieldTrees <- treeDataInPlots %>%
  filter(plot_id == chm_name)

# Get the field plot extent 
plot_extent <- fieldPlots_extents |>
  filter(plot_id == chm_name)

```

Map the NAIP layers: 

```{r}
mapview(
  chm_NAIP,
  layer.name = "NAIP CHM",
  col.regions = colorRampPalette(c("red", "yellow", "darkgreen"))
) +
  mapview(treePoints_NAIP, color = "blue", layer.name = "NAIP Tree Tops") +
  mapview(treeCrowns_NAIP, color = "red", lwd = 3,layer.name = "NAIP Crowns") 


+
  mapview(fieldTrees, color = "black", col.regions = "white", layer.name = "Field Trees") +
  mapview(st_geometry(plot_extent), color = "blue", fill = NA, layer.name)

plot(treePoints_NAIP$geometry)

# mapview(chm_Lidar,
#         layer.name = "Lidar CHM",
#          col.regions = colorRampPalette(c("red", "yellow", "darkgreen")))
#         ) +
  mapview(treePoints_NAIP, color = "blue", layer.name = "NAIP Tree Tops") +
  mapview(
    treeCrowns_NAIP,
    color = "red",
    lwd = 3,
    fill = NA,
    layer.name = "NAIP Crowns"
  ) +
  mapview(
    fieldTrees,
    color = "black",
    col.regions = "white",
    cex = 2,
    alpha = 1,
    layer.name = "Field Trees"
  ) +
  mapview(st_geometry(plot_extent), color = "blue", lwd = 3, fill = NA, layer.name = "Plot Extent")

```

Now plot the lidar data
```{r}
mapview(
  chm_Lidar,
  layer.name = "Lidar CHM",
  col.regions = colorRampPalette(c("red", "yellow", "green"))
) +
  mapview(treePoints_Lidar, color = "blue", layer.name = "Lidar Tree Tops") +
  mapview(
    treeCrowns_Lidar,
    color = "red",
    lwd = 3,
    fill = NA,
    layer.name = "Lidar Crowns"
  ) +
  mapview(
    fieldTrees,
    color = "black",
    col.regions = "white",
    cex = 2,
    alpha = 1,
    layer.name = "Field Trees"
  ) +
  mapview(st_geometry(plot_extent), color = "blue", fill = NA, layer.name = "Plot Extent")



```

## Model Biomass of Individual Trees

We'll now use the metrics associated with the detected trees (tree height & crown area) to run numerical optimization. In essence, the individual tree-level predictor coefficients are tuned to minimize the sum of squared errors between the predicted plot-level biomass (sum of tree-level biomass) and the observed plot-level biomass from the field data.

Let's run it on the NAIP-derived tree list: 

```{r}
# Create tree and plot df
df_tree <- allTreeCrowns_NAIP_final
df_plot <- plotLevelBiomass

# Select tree variables
df_tree <- df_tree[,c("plotID", "tree_height_m", "area_m2")]
df_tree <- st_drop_geometry(df_tree)

# Select plot variables
df_plot <- df_plot[,c("plot_id", "total_biomass")]
df_plot <- st_drop_geometry(df_plot)

# Subset data to these variables
colnames(df_tree) <- c("plot_id", "tree_height", "tree_area")
colnames(df_plot) <- c("plot_id", "plot_biomass")

# Merge the data
df <- merge(df_plot, df_tree)

# Split into list by plot
plot_ids <- unique(df$plot_id)
tree_list <- split(df, df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(df$plot_biomass, df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  # ensure ordering matches
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1, 
                b = 1, 
                c = 1
)

# Fit the parameters
fit <- optim(
  par = 
    start_vals,
  fn = objective,
  
  # Broyden, Fletcher, Goldfarb, and Shanno
  # Uses function values and gradients to build up a picture of the surface to be optimized
  method = "L-BFGS-B",
  lower = c(1e-6, 
            -5,
            -5),
  upper = c(100,  
            5,
            5))

# Whats here?
# The best set of parameters found
# The coefficients for individual tree level biomass! 
fit$par 

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_df <- left_join(predicted_df, observed_df, by = "PlotID")

# scatterplot
plot(plot_biomass, predicted,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass")
abline(0, 1, col = "red")

# Observed and predicted vectors
obs <- plot_biomass
pred <- predicted[names(plot_biomass)]   

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2 <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
NAIP_TreeDetection_ModResults <- data.frame(
  RMSE = RMSE,
  MAE = MAE,
  Bias = Bias,
  MAPE_percent = MAPE,
  R2 = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result 
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)


```


And the lidar-derived tree list

```{r}
# Create tree and plot df
df_tree <- allTreeCrowns_Lidar_final
df_plot <- plotLevelBiomass

# Select tree variables
df_tree <- df_tree[,c("plotID", "tree_height_m", "area_m2")]
df_tree <- st_drop_geometry(df_tree)

# Select plot variables
df_plot <- df_plot[,c("plot_id", "total_biomass")]
df_plot <- st_drop_geometry(df_plot)

# Subset data to these variables
colnames(df_tree) <- c("plot_id", "tree_height", "tree_area")
colnames(df_plot) <- c("plot_id", "plot_biomass")

# Merge the data
df <- merge(df_plot, df_tree)

# Split into list by plot
plot_ids <- unique(df$plot_id)
tree_list <- split(df, df$plot_id)

# Extract the observed plot biomass
plot_biomass <- tapply(df$plot_biomass, df$plot_id, FUN = function(x) x[1])

# Define a predictive allometric model
predict_plot_biomass <- function(params) {
  a <- params[1]
  b <- params[2]
  c <- params[3]
  # d <- params[4]
  
  pred <- numeric(length(tree_list))
  names(pred) <- names(tree_list)
  
  for(i in seq_along(tree_list)) {
    trees <- tree_list[[i]]
    
    # Equation 1: Power Law
    tree_vals <- a * (trees$tree_height^b) * (trees$tree_area^c)
    
    # Equation 2: Additive 
    # tree_vals <- a + b * trees$tree_height + c * trees$tree_area
    
    # Equation 3: Additive non-linear
    # tree_vals <- a * (trees$tree_height^b) + c * (trees$tree_area^d)
    
    # Equation 4: Just tree height
    # tree_vals <- a * (trees$tree_height^b)
    
    # Equation 5: Just crown area
    # tree_vals <- a * (trees$tree_area^b)
    
    pred[i] <- sum(tree_vals, na.rm = TRUE)
  }
  
  pred
}

# Define the objective function (sum of squared errors)
objective <- function(params) {
  pred <- predict_plot_biomass(params)
  # ensure ordering matches
  resid <- plot_biomass - pred[names(plot_biomass)]
  sum(resid^2)
}

# Fit the model through numerical optimization
start_vals <- c(a = 0.1, 
                b = 1, 
                c = 1
)

# Fit the parameters
fit <- optim(
  par = 
    start_vals,
  fn = objective,
  
  # Broyden, Fletcher, Goldfarb, and Shanno
  # Uses function values and gradients to build up a picture of the surface to be optimized
  method = "L-BFGS-B",
  lower = c(1e-6, 
            -5,
            -5),
  upper = c(100,  
            5,
            5))

# Whats here?
# The best set of parameters found
# The coefficients for individual tree level biomass! 
fit$par 

# Return the predicted biomass
predicted <- predict_plot_biomass(fit$par)
print(predicted)

# Create a dataframe of the predicted biomass
predicted_df <- as.data.frame(predicted)
predicted_df$PlotID <- rownames(predicted_df)
rownames(predicted_df) <- NULL

# Create a dataframe of the measured plot biomass
observed_df <- as.data.frame(plot_biomass)
observed_df$PlotID <- rownames(observed_df)
rownames(observed_df) <- NULL

# Join the two biomass estimates
final_df <- left_join(predicted_df, observed_df, by = "PlotID")

# scatterplot
plot(plot_biomass, predicted,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass")
abline(0, 1, col = "red")

# Observed and predicted vectors
obs <- plot_biomass
pred <- predicted[names(plot_biomass)]   

# 1. Root Mean Squared Error
RMSE <- sqrt(mean((obs - pred)^2))
print(RMSE)

# 2. Mean Absolute Error
MAE <- mean(abs(obs - pred))
print(MAE)

# 3. Bias (signed error)
Bias <- mean(pred - obs)

# 4. Mean Absolute Percent Error
MAPE <- mean(abs((obs - pred) / obs)) * 100

# 5. R-squared (coefficient of determination)
SSE <- sum((obs - pred)^2)
SST <- sum((obs - mean(obs))^2)
R2 <- 1 - SSE/SST
print(R2)

# 6. Relative RMSE (RMSE / mean observed)
rRMSE <- RMSE / mean(obs) * 100

# Save metrics
Lidar_TreeDetection_ModResults <- data.frame(
  RMSE = RMSE,
  MAE = MAE,
  Bias = Bias,
  MAPE_percent = MAPE,
  R2 = R2,
  rRMSE_percent = rRMSE
)

# Now plot the results
plot(obs, pred,
     xlab = "Observed plot biomass",
     ylab = "Predicted plot biomass",
     main = "Observed vs Predicted Plot Biomass")

abline(0, 1, col = "red", lwd = 2)

# Print the R2 result 
text(x = min(obs),
     y = max(pred),
     labels = paste("R2 =", round(R2, 3)),
     pos = 4)


```

## Optimize the parameters

This script iteratively tests model performance on a multitude of different tree segmentation parameters. This can be performed on either the lidar or NAIP-derived tree lists. The parameters that yield the best performance can then be used to run tree detection across all plots.

The user inputs can be tuned below. 
```{r}
# Choose source and corresponding CHM list
source_name <- "NAIP"
# source_name <- "Lidar"

if (source_name == "Lidar") {
  chm_list <- allPlotCHMs_Lidar
  plots_to_run <- intersect(Lidar_PlotNames, plotLevelBiomass$plot_id)
}

if (source_name == "NAIP") {
  chm_list <- allPlotsCHMs_NAIP
  plots_to_run <- intersect(names(allPlotsCHMs_NAIP), plotLevelBiomass$plot_id)
}
  
# plots_to_run <- c("UOFU_AZCV_P01", "UOFU_AZCV_P02")

# Parameter grid
param_grid <- expand.grid(
  th_tree = c(0.50),
  th_seed = c(0.3, 0.35, 0.40, 0.45, 0.5, 0.55, 0.6),
  th_cr   = c(0.3, 0.35, 0.4, 0.45, 0.55, 0.55, 0.60),
  max_cr  = c(15, 20, 25),
  stringsAsFactors = FALSE
)

# Starting values for biomass model
# tree_biomass = a * tree_height^b * tree_area^c
start_vals   <- c(a = 0.1, b = 1, c = 1)
lower_bounds <- c(1e-6, -5, -5)
upper_bounds <- c(100, 5, 5)

# Observed plot biomass
df_plot_obs <- plotLevelBiomass %>%
  st_drop_geometry() %>%
  dplyr::select(plot_id, total_biomass) %>%
  rename(plot_biomass = total_biomass)

# ============================================================
# 2. MAIN ITERATION FUNCTION
# ============================================================

run_segmentation_iteration <- function(th_tree,
                                       th_seed,
                                       th_cr,
                                       max_cr,
                                       plots_to_run,
                                       chm_list,
                                       h,
                                       fieldPlots_extents,
                                       df_plot_obs,
                                       start_vals,
                                       lower_bounds,
                                       upper_bounds,
                                       source_name = "Lidar") {
  
  tree_points_list <- list()
  tree_crowns_list <- list()
  
  for (plot_id in plots_to_run) {
    
    message(
      "Processing plot: ", plot_id,
      " | th_tree=", th_tree,
      " th_seed=", th_seed,
      " th_cr=", th_cr,
      " max_cr=", max_cr
    )
    
    chm <- chm_list[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    chm <- terra::project(chm, "EPSG:32612")
    
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree,
        th_seed  = th_seed,
        th_cr    = th_cr,
        max_cr   = max_cr
      )(),
      error = function(e) {
        message("Dalponte failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    names(crowns_poly) <- "treeID"
    crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")
    
    zmax <- tryCatch(
      terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
      error = function(e) {
        message("Height extraction failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
      message(
        "Height extraction row mismatch for plot: ", plot_id,
        " | crowns=", terra::nrow(crowns_poly),
        " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax))
      )
      next
    }
    
    crowns_poly$tree_height_m <- zmax[, 2]
    
    crowns_sf <- st_as_sf(crowns_poly)
    crowns_sf$plotID <- plot_id
    ttops_sf$plotID <- plot_id
    
    tree_points_list[[plot_id]] <- ttops_sf
    tree_crowns_list[[plot_id]] <- crowns_sf
  }
  
  all_tree_points <- bind_rows(tree_points_list)
  all_tree_crowns <- bind_rows(tree_crowns_list)
  
  if (nrow(all_tree_points) == 0 || nrow(all_tree_crowns) == 0) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = 0,
        n_detected_trees_total = 0,
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = data.frame()
    ))
  }
  
  all_tree_points <- st_transform(all_tree_points, st_crs(fieldPlots_extents))
  all_tree_crowns <- st_transform(all_tree_crowns, st_crs(fieldPlots_extents))
  
  ttops_in_plots <- all_tree_points[
    lengths(st_within(all_tree_points, fieldPlots_extents)) > 0,
  ]
  
  keys <- ttops_in_plots %>%
    st_drop_geometry() %>%
    distinct(plotID, treeID)
  
  all_tree_crowns_final <- all_tree_crowns %>%
    inner_join(keys, by = c("plotID", "treeID"))
  
  if (nrow(all_tree_crowns_final) == 0) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = 0,
        n_detected_trees_total = 0,
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = data.frame()
    ))
  }
  
  # Prepare tree-level dataframe
  df_tree <- all_tree_crowns_final %>%
    st_drop_geometry() %>%
    dplyr::select(plotID, tree_height_m, area_m2) %>%
    rename(
      plot_id = plotID,
      tree_height = tree_height_m,
      tree_area = area_m2
    )
  
  df <- merge(df_plot_obs, df_tree, by = "plot_id")
  
  if (nrow(df) == 0) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = 0,
        n_detected_trees_total = nrow(df_tree),
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = df_tree
    ))
  }
  
  tree_list <- split(df, df$plot_id)
  plot_biomass <- tapply(df$plot_biomass, df$plot_id, function(x) x[1])
  
  predict_plot_biomass <- function(params) {
    a <- params[1]
    b <- params[2]
    c <- params[3]
    
    pred <- numeric(length(tree_list))
    names(pred) <- names(tree_list)
    
    for (i in seq_along(tree_list)) {
      trees <- tree_list[[i]]
      tree_vals <- a * (trees$tree_height ^ b) * (trees$tree_area ^ c)
      pred[i] <- sum(tree_vals, na.rm = TRUE)
    }
    
    pred
  }
  
  objective_function <- function(params) {
    pred <- predict_plot_biomass(params)
    obs <- plot_biomass[names(pred)]
    sum((obs - pred)^2, na.rm = TRUE)
  }
  
  fit <- tryCatch(
    optim(
      par = start_vals,
      fn = objective_function,
      method = "L-BFGS-B",
      lower = lower_bounds,
      upper = upper_bounds
    ),
    error = function(e) {
      message("Optimization failed: ", e$message)
      NULL
    }
  )
  
  if (is.null(fit)) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = length(plot_biomass),
        n_detected_trees_total = nrow(df_tree),
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = df_tree
    ))
  }
  
  pred <- predict_plot_biomass(fit$par)
  obs <- plot_biomass[names(pred)]
  
  RMSE <- sqrt(mean((obs - pred)^2, na.rm = TRUE))
  MAE <- mean(abs(obs - pred), na.rm = TRUE)
  Bias <- mean(pred - obs, na.rm = TRUE)
  MAPE <- mean(abs((obs - pred) / ifelse(obs == 0, NA, obs)), na.rm = TRUE) * 100
  SSE <- sum((obs - pred)^2, na.rm = TRUE)
  SST <- sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
  R2 <- ifelse(SST == 0, NA_real_, 1 - SSE / SST)
  rRMSE <- RMSE / mean(obs, na.rm = TRUE) * 100
  
  plot_metrics <- data.frame(
    plot_id = names(pred),
    observed_plot_biomass = as.numeric(obs),
    predicted_plot_biomass = as.numeric(pred),
    residual = as.numeric(pred - obs),
    stringsAsFactors = FALSE
  )
  
  modeled_trees <- df_tree %>%
    mutate(
      source = source_name,
      th_tree = th_tree,
      th_seed = th_seed,
      th_cr = th_cr,
      max_cr = max_cr,
      a = fit$par[1],
      b = fit$par[2],
      c = fit$par[3]
    )
  
  iteration_summary <- data.frame(
    source = source_name,
    th_tree = th_tree,
    th_seed = th_seed,
    th_cr = th_cr,
    max_cr = max_cr,
    a = fit$par[1],
    b = fit$par[2],
    c = fit$par[3],
    converged = fit$convergence == 0,
    objective_value = fit$value,
    n_plots_modeled = length(obs),
    n_detected_trees_total = nrow(df_tree),
    RMSE = RMSE,
    MAE = MAE,
    Bias = Bias,
    MAPE_percent = MAPE,
    R2 = R2,
    rRMSE_percent = rRMSE,
    stringsAsFactors = FALSE
  )
  
  list(
    iteration_summary = iteration_summary,
    plot_metrics = plot_metrics,
    modeled_trees = modeled_trees
  )
}

# ============================================================
# 3. RUN ALL ITERATIONS WITH TIMING + PROGRESS
# ============================================================

n_iters <- nrow(param_grid)
iteration_results <- vector("list", n_iters)

total_start_time <- Sys.time()
pb <- txtProgressBar(min = 0, max = n_iters, style = 3)

for (i in seq_len(n_iters)) {
  
  iter_start_time <- Sys.time()
  pars <- param_grid[i, ]
  
  message(
    "\nStarting iteration ", i, " of ", n_iters,
    " | th_tree=", pars$th_tree,
    " th_seed=", pars$th_seed,
    " th_cr=", pars$th_cr,
    " max_cr=", pars$max_cr
  )
  
  iteration_results[[i]] <- run_segmentation_iteration(
    th_tree = pars$th_tree,
    th_seed = pars$th_seed,
    th_cr = pars$th_cr,
    max_cr = pars$max_cr,
    plots_to_run = plots_to_run,
    chm_list = chm_list,
    h = h,
    fieldPlots_extents = fieldPlots_extents,
    df_plot_obs = df_plot_obs,
    start_vals = start_vals,
    lower_bounds = lower_bounds,
    upper_bounds = upper_bounds,
    source_name = source_name
  )
  
  iter_end_time <- Sys.time()
  iter_time_sec <- as.numeric(difftime(iter_end_time, iter_start_time, units = "secs"))
  iteration_results[[i]]$iteration_summary$runtime_sec <- iter_time_sec
  
  setTxtProgressBar(pb, i)
  
  elapsed_total_sec <- as.numeric(difftime(iter_end_time, total_start_time, units = "secs"))
  avg_time_per_iter <- elapsed_total_sec / i
  remaining_iters <- n_iters - i
  est_remaining_sec <- avg_time_per_iter * remaining_iters
  
  message(
    "Finished iteration ", i, "/", n_iters,
    " in ", round(iter_time_sec, 2), " sec",
    " | elapsed: ", round(elapsed_total_sec / 60, 2), " min",
    " | est. remaining: ", round(est_remaining_sec / 60, 2), " min"
  )
}

close(pb)

total_end_time <- Sys.time()
total_time_min <- as.numeric(difftime(total_end_time, total_start_time, units = "mins"))

message("\nAll iterations completed in ", round(total_time_min, 2), " minutes.")

# ============================================================
# 4. COLLECT RESULTS
# ============================================================

optimization_results <- tibble(
  iteration = seq_along(iteration_results),
  summary = lapply(iteration_results, `[[`, "iteration_summary"),
  plot_metrics = lapply(iteration_results, `[[`, "plot_metrics"),
  modeled_trees = lapply(iteration_results, `[[`, "modeled_trees")
) %>%
  mutate(summary = purrr::map(summary, as.data.frame)) %>%
  tidyr::unnest(summary) %>%
  arrange(RMSE, desc(R2))

optimization_plot_metrics <- bind_rows(
  lapply(seq_along(iteration_results), function(i) {
    x <- iteration_results[[i]]$plot_metrics
    if (nrow(x) == 0) return(NULL)
    
    pars <- param_grid[i, ]
    summ <- iteration_results[[i]]$iteration_summary
    
    x %>%
      mutate(
        iteration = i,
        source = source_name,
        th_tree = pars$th_tree,
        th_seed = pars$th_seed,
        th_cr = pars$th_cr,
        max_cr = pars$max_cr,
        a = summ$a,
        b = summ$b,
        c = summ$c,
        runtime_sec = summ$runtime_sec
      )
  })
)

optimization_tree_table <- bind_rows(
  lapply(seq_along(iteration_results), function(i) {
    x <- iteration_results[[i]]$modeled_trees
    if (nrow(x) == 0) return(NULL)
    x %>% mutate(iteration = i)
  })
)

optimization_results
best_iteration <- optimization_results %>% slice(1)
best_iteration


```

# Export the results
```{r}

outNAIP_model <- r'(S:\ursa\howell\Research\PJ_forests\08_ModelResults\01_NAIP)'
write_csv(optimization_results, paste0(outNAIP_model, "/NAIP_TreeDetection_Optimization_Results.csv"))


```