---
title: "Modeling Aboveground Biomass of Individual Trees in Pinyon-Juniper Woodlands"
author: "Jake Howell"
date: "`r Sys.Date()`"
purpose: 
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
---

# Clear Environment
```{r}


```

```{r}
# Set output formats
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```
## Script Purpose:
 Here we present our methods for estimating the aboveground biomass of individual trees within Pinyon-Juniper Woodlands using Using field data collected from 180 field plots, paired with two Canopy Height Model (CHM) sources.

Let's begin by getting things set up.

Be sure the required packages are installed and libraries are loaded.

```{r setup, include = FALSE}
# Check required packages are installed
pckgs <- c("sf",
           "dplyr",
           "sp",
           # "raster",
           "terra",
           "lidR",
           "mapview",
           "units",
           "ggplot2",
           "broom",
           "purrr",
           "stringr",
           "tidyr")

# Point to packages
newPckgs <- pckgs[!(pckgs %in% installed.packages()[,"Package"])]

# If length is greater than 0, install the packages
if(length(newPckgs) > 0) install.packages(newPckgs)

# Load libraries
library(sf)
library(dplyr)
library(sp)
# library(raster)
library(terra)
library(lidR)
library(mapview)
library(units)
library(ggplot2)
library(broom)
library(purrr)
library(stringr)
library(tidyr)

```

## Set up

We'll begin by setting the paths to our data. The required data are as follows 

- Lidar-Derived Canopy Height Models (.tif)
- NAIP-derived Canopy Height Models (.tif)
- Spreadsheet of field plot data (.csv)
- Spreadsheet of tree level data (.csv)
- Spreadsheet of AGB equations (.csv)

```{r}
# Set root data location
loc <- r'(S:\ursa\howell\Research\PJ_forests)'

# Set the name of the site data folder 
foldSite <- "01_Sites"

# Set the name of the plot level folder
foldPlot <- "02_Plots"

# Set the folder of NAIP-derived site-level CHMs
foldNAIP_CHM <- "01_NAIPDerivedCHMs"

# Set the folder of the lidar-derived plot-level CHMs
foldLidar_CHM <- "03_LidarDerivedCHMs/Original"

# Set the name of the plot level .csv
namePlotData <- "plot_level_data_uofu.csv"

## TREE-LEVEL DATA: 
foldTree <- "03_TreeData"

# Set the name of the tree data
nameTreeData <- "tree_level_data.csv"

# Set name to AGB equations
nameAGB <- "all_tree_species_in_pj_fia_plots_new.csv"

# Initialize paths
pathNAIPsites <- file.path(loc, foldPlot, foldNAIP_CHM)
pathLidarPlots <- file.path(loc, foldPlot, foldLidar_CHM)
pathPlotData <- file.path(loc, foldPlot, namePlotData)
pathTreeData <- file.path(loc, foldTree, nameTreeData)
pathAGBData <- file.path(loc, foldTree, nameAGB)

```

Now let's set our output paths
```{r}
# Path to export our clipped NAIP-derived CHMs
outClippedPlots_NAIP <- r'(S:\ursa\howell\Research\PJ_forests\02_Plots\01_NAIPDerivedCHMs)'

# Path to export our clipped lidar-derived CHMs
outClippedPlots_Lidar <- r'(S:\ursa\howell\Research\PJ_forests\02_Plots\03_LidarDerivedCHMs)'

# Path to export our lidar-derived tree crowns
# outTreeCrowns_Lidar <- 
# 
# # Path to export our lidar-derived tree points
# outTreePoints_Lidar <- 
#   
# # Path to export our NAIP-derived tree crowns
# outTreeCrowns_NAIP <-   
#   
# # Path to export our NAIP-derived tree points
# outTreePoints_NAIP <- 

```


```{r}
# List the NAIP-derived CHM sites
full_NAIPPlotNames <- list.files(pathNAIPsites, pattern = "\\.tif$", full.names = TRUE)

# Now read the rasters as a list
NAIPPlotRasters <- lapply(full_NAIPPlotNames, rast)


# Get the names of the NAIP plots with just the UOFU_XXXX_PX portion
NAIPPlotNames <- gsub("^buffered_|\\.tif$", "", NAIPPlotRasters)

# List the lidar-derived plots within the folder
lidarPlotCHM_names <- list.files(pathLidarPlots, pattern = "\\.tif$", full.names = TRUE)

# Now read the rasters as a list
lidarPlotRasters <- lapply(lidarPlotCHM_names, rast)

```


```{r}
# Read plot field data
plotData <- read.csv(pathPlotData)

# Read tree data
treeData <- read.csv(pathTreeData)

# Read biomass equations
biomassEquations <- read.csv(pathAGBData)
```


```{r}
# In the tree data table, alter the species column
treeData <- treeData |>
  rename(Species = tree_sp)

# In the biomass equations table, alter the species column so it matches with the tree table
biomassEquations <- biomassEquations |>
  rename(Species = plants.code)

# Join the tree data to the biomass equations by Species column
treeData <- left_join(treeData, biomassEquations, by = "Species")

# Now estimate biomass for the field data trees by applying the biomass equations
treeData <- treeData |>
  mutate(tree_biomass_kg = exp(beta.0 + beta.1 * log(tree_diam)))


```


```{r}
# Create function to make one polygon the four sides of each field plot extent
createFieldPlotExtents <- function(row) {
  coords <- matrix(c(
    row$plot_ne_x, row$plot_ne_y,
    row$plot_se_x, row$plot_se_y,
    row$plot_sw_x, row$plot_sw_y,
    row$plot_nw_x, row$plot_nw_y,
    row$plot_ne_x, row$plot_ne_y
  ), ncol = 2, byrow = TRUE)
  st_polygon(list(coords))
}

# Go through each row of the plot data spreadsheet and create the field plot extents
fieldPlots_extents <- plotData |>
  rowwise() |>
  mutate(geometry = st_sfc(createFieldPlotExtents(pick(everything())))) |>
  ungroup()

# Set the current CRS and reproject
fieldPlots_extents <- st_as_sf(fieldPlots_extents, crs = 4326) |>
  st_transform(32612)

```


```{r}
# Return just the plot ID's from the field plot data
plot_ids <- unique(fieldPlots_extents$plot_id)

# Subset the tree data to just the unique plot names
treeDataInPlots_raw <- treeData[treeData$plot_id %in% plot_ids, ]

# FIX THE ERROR IN LAT/LONG ASSIGNMENT:
treeDataInPlots_fixed <- treeDataInPlots_raw %>%
  mutate(
    # start by assuming they are correct
    lon = tree_x,
    lat = tree_y,

    # identify likely swapped rows:
    # lon should be negative ~ -180..0 and lat should be positive ~ 0..90
    swapped = (tree_x > 0 & tree_x < 90) & (tree_y < 0 & tree_y > -180),

    # for swapped rows, flip them
    lon = if_else(swapped, tree_y, lon),
    lat = if_else(swapped, tree_x, lat)
  )

# Convert the treeDataInPlots to a spatial object
treeDataInPlots_fixed <- st_as_sf(treeDataInPlots_fixed,
                            coords = c("lon", "lat"), 
                            crs = 4326,
                            remove = FALSE)

# Now transform to UTM Zone 12N
treeDataInPlots <- st_transform(treeDataInPlots_fixed, 32612)

# Estimate Biomass Summary for each plot
plotLevelBiomass <- treeDataInPlots |>
  group_by(plot_id) |>
  summarise(
    total_biomass = sum(tree_biomass_kg, na.rm = TRUE),
    mean_biomass = mean(tree_biomass_kg, na.rm = TRUE),
    n_trees_GPS = n(),
    .groups = 'drop'
  )

```


```{r}
# Choose source and corresponding CHM list
source_name <- "NAIP"
# source_name <- "Lidar"

if (source_name == "Lidar") {
  chm_list <- lidarPlotRasters
  plots_to_run <- intersect(lidar, plotLevelBiomass$plot_id)
}

if (source_name == "NAIP") {
  chm_list <- allPlotsCHMs_NAIP
  plots_to_run <- intersect(names(allPlotsCHMs_NAIP), plotLevelBiomass$plot_id)
}

# plots_to_run <- c("UOFU_AZCV_P01", "UOFU_AZCV_P02")

# Parameter grid
param_grid <- expand.grid(
  th_tree = c(0.50),
  th_seed = c(0.3, 0.35, 0.40, 0.45, 0.5, 0.55, 0.6),
  th_cr   = c(0.3, 0.35, 0.4, 0.45, 0.55, 0.55, 0.60),
  max_cr  = c(15, 20, 25),
  stringsAsFactors = FALSE
)

# Starting values for biomass model
# tree_biomass = a * tree_height^b * tree_area^c
start_vals   <- c(a = 0.1, b = 1, c = 1)
lower_bounds <- c(1e-6, -5, -5)
upper_bounds <- c(100, 5, 5)

# Observed plot biomass
df_plot_obs <- plotLevelBiomass %>%
  st_drop_geometry() %>%
  dplyr::select(plot_id, total_biomass) %>%
  rename(plot_biomass = total_biomass)

# ============================================================
# 2. MAIN ITERATION FUNCTION
# ============================================================

run_segmentation_iteration <- function(th_tree,
                                       th_seed,
                                       th_cr,
                                       max_cr,
                                       plots_to_run,
                                       chm_list,
                                       h,
                                       fieldPlots_extents,
                                       df_plot_obs,
                                       start_vals,
                                       lower_bounds,
                                       upper_bounds,
                                       source_name = "Lidar") {
  
  tree_points_list <- list()
  tree_crowns_list <- list()
  
  for (plot_id in plots_to_run) {
    
    message(
      "Processing plot: ", plot_id,
      " | th_tree=", th_tree,
      " th_seed=", th_seed,
      " th_cr=", th_cr,
      " max_cr=", max_cr
    )
    
    chm <- chm_list[[plot_id]]
    
    if (is.null(chm)) {
      message("CHM is NULL for plot: ", plot_id)
      next
    }
    
    chm <- terra::project(chm, "EPSG:32612")
    
    ttops <- tryCatch(
      locate_trees(chm, lmf(ws = h, hmin = 0.5)),
      error = function(e) {
        message("Treetops failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops) ||
        (inherits(ttops, "data.frame") && nrow(ttops) == 0) ||
        (inherits(ttops, "sf") && nrow(ttops) == 0) ||
        (inherits(ttops, "SpatVector") && terra::nrow(ttops) == 0)) {
      message("No treetops detected for plot: ", plot_id)
      next
    }
    
    ttops_sf <- tryCatch(
      st_as_sf(ttops),
      error = function(e) {
        message("st_as_sf failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(ttops_sf) || nrow(ttops_sf) == 0) {
      message("No valid treetop sf object for plot: ", plot_id)
      next
    }
    
    crowns_id <- tryCatch(
      dalponte2016(
        chm      = chm,
        treetops = ttops_sf,
        th_tree  = th_tree,
        th_seed  = th_seed,
        th_cr    = th_cr,
        max_cr   = max_cr
      )(),
      error = function(e) {
        message("Dalponte failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_id)) {
      message("No crown raster returned for plot: ", plot_id)
      next
    }
    
    crowns_poly <- tryCatch(
      terra::as.polygons(crowns_id, dissolve = TRUE, na.rm = TRUE),
      error = function(e) {
        message("as.polygons failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(crowns_poly) || terra::nrow(crowns_poly) == 0) {
      message("No crown polygons for plot: ", plot_id)
      next
    }
    
    names(crowns_poly) <- "treeID"
    crowns_poly$area_m2 <- terra::expanse(crowns_poly, unit = "m")
    
    zmax <- tryCatch(
      terra::extract(chm, crowns_poly, fun = max, na.rm = TRUE),
      error = function(e) {
        message("Height extraction failed: ", e$message)
        NULL
      }
    )
    
    if (is.null(zmax) || nrow(zmax) != terra::nrow(crowns_poly)) {
      message(
        "Height extraction row mismatch for plot: ", plot_id,
        " | crowns=", terra::nrow(crowns_poly),
        " | zmax=", ifelse(is.null(zmax), 0, nrow(zmax))
      )
      next
    }
    
    crowns_poly$tree_height_m <- zmax[, 2]
    
    crowns_sf <- st_as_sf(crowns_poly)
    crowns_sf$plotID <- plot_id
    ttops_sf$plotID <- plot_id
    
    tree_points_list[[plot_id]] <- ttops_sf
    tree_crowns_list[[plot_id]] <- crowns_sf
  }
  
  all_tree_points <- bind_rows(tree_points_list)
  all_tree_crowns <- bind_rows(tree_crowns_list)
  
  if (nrow(all_tree_points) == 0 || nrow(all_tree_crowns) == 0) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = 0,
        n_detected_trees_total = 0,
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = data.frame()
    ))
  }
  
  all_tree_points <- st_transform(all_tree_points, st_crs(fieldPlots_extents))
  all_tree_crowns <- st_transform(all_tree_crowns, st_crs(fieldPlots_extents))
  
  ttops_in_plots <- all_tree_points[
    lengths(st_within(all_tree_points, fieldPlots_extents)) > 0,
  ]
  
  keys <- ttops_in_plots %>%
    st_drop_geometry() %>%
    distinct(plotID, treeID)
  
  all_tree_crowns_final <- all_tree_crowns %>%
    inner_join(keys, by = c("plotID", "treeID"))
  
  if (nrow(all_tree_crowns_final) == 0) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = 0,
        n_detected_trees_total = 0,
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = data.frame()
    ))
  }
  
  # Prepare tree-level dataframe
  df_tree <- all_tree_crowns_final %>%
    st_drop_geometry() %>%
    dplyr::select(plotID, tree_height_m, area_m2) %>%
    rename(
      plot_id = plotID,
      tree_height = tree_height_m,
      tree_area = area_m2
    )
  
  df <- merge(df_plot_obs, df_tree, by = "plot_id")
  
  if (nrow(df) == 0) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = 0,
        n_detected_trees_total = nrow(df_tree),
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = df_tree
    ))
  }
  
  tree_list <- split(df, df$plot_id)
  plot_biomass <- tapply(df$plot_biomass, df$plot_id, function(x) x[1])
  
  predict_plot_biomass <- function(params) {
    a <- params[1]
    b <- params[2]
    c <- params[3]
    
    pred <- numeric(length(tree_list))
    names(pred) <- names(tree_list)
    
    for (i in seq_along(tree_list)) {
      trees <- tree_list[[i]]
      tree_vals <- a * (trees$tree_height ^ b) * (trees$tree_area ^ c)
      pred[i] <- sum(tree_vals, na.rm = TRUE)
    }
    
    pred
  }
  
  objective_function <- function(params) {
    pred <- predict_plot_biomass(params)
    obs <- plot_biomass[names(pred)]
    sum((obs - pred)^2, na.rm = TRUE)
  }
  
  fit <- tryCatch(
    optim(
      par = start_vals,
      fn = objective_function,
      method = "L-BFGS-B",
      lower = lower_bounds,
      upper = upper_bounds
    ),
    error = function(e) {
      message("Optimization failed: ", e$message)
      NULL
    }
  )
  
  if (is.null(fit)) {
    return(list(
      iteration_summary = data.frame(
        source = source_name,
        th_tree = th_tree,
        th_seed = th_seed,
        th_cr = th_cr,
        max_cr = max_cr,
        a = NA_real_,
        b = NA_real_,
        c = NA_real_,
        converged = FALSE,
        objective_value = NA_real_,
        n_plots_modeled = length(plot_biomass),
        n_detected_trees_total = nrow(df_tree),
        RMSE = NA_real_,
        MAE = NA_real_,
        Bias = NA_real_,
        MAPE_percent = NA_real_,
        R2 = NA_real_,
        rRMSE_percent = NA_real_,
        stringsAsFactors = FALSE
      ),
      plot_metrics = data.frame(),
      modeled_trees = df_tree
    ))
  }
  
  pred <- predict_plot_biomass(fit$par)
  obs <- plot_biomass[names(pred)]
  
  RMSE <- sqrt(mean((obs - pred)^2, na.rm = TRUE))
  MAE <- mean(abs(obs - pred), na.rm = TRUE)
  Bias <- mean(pred - obs, na.rm = TRUE)
  MAPE <- mean(abs((obs - pred) / ifelse(obs == 0, NA, obs)), na.rm = TRUE) * 100
  SSE <- sum((obs - pred)^2, na.rm = TRUE)
  SST <- sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
  R2 <- ifelse(SST == 0, NA_real_, 1 - SSE / SST)
  rRMSE <- RMSE / mean(obs, na.rm = TRUE) * 100
  
  plot_metrics <- data.frame(
    plot_id = names(pred),
    observed_plot_biomass = as.numeric(obs),
    predicted_plot_biomass = as.numeric(pred),
    residual = as.numeric(pred - obs),
    stringsAsFactors = FALSE
  )
  
  modeled_trees <- df_tree %>%
    mutate(
      source = source_name,
      th_tree = th_tree,
      th_seed = th_seed,
      th_cr = th_cr,
      max_cr = max_cr,
      a = fit$par[1],
      b = fit$par[2],
      c = fit$par[3]
    )
  
  iteration_summary <- data.frame(
    source = source_name,
    th_tree = th_tree,
    th_seed = th_seed,
    th_cr = th_cr,
    max_cr = max_cr,
    a = fit$par[1],
    b = fit$par[2],
    c = fit$par[3],
    converged = fit$convergence == 0,
    objective_value = fit$value,
    n_plots_modeled = length(obs),
    n_detected_trees_total = nrow(df_tree),
    RMSE = RMSE,
    MAE = MAE,
    Bias = Bias,
    MAPE_percent = MAPE,
    R2 = R2,
    rRMSE_percent = rRMSE,
    stringsAsFactors = FALSE
  )
  
  list(
    iteration_summary = iteration_summary,
    plot_metrics = plot_metrics,
    modeled_trees = modeled_trees
  )
}

# ============================================================
# 3. RUN ALL ITERATIONS WITH TIMING + PROGRESS
# ============================================================

n_iters <- nrow(param_grid)
iteration_results <- vector("list", n_iters)

total_start_time <- Sys.time()
pb <- txtProgressBar(min = 0, max = n_iters, style = 3)

for (i in seq_len(n_iters)) {
  
  iter_start_time <- Sys.time()
  pars <- param_grid[i, ]
  
  message(
    "\nStarting iteration ", i, " of ", n_iters,
    " | th_tree=", pars$th_tree,
    " th_seed=", pars$th_seed,
    " th_cr=", pars$th_cr,
    " max_cr=", pars$max_cr
  )
  
  iteration_results[[i]] <- run_segmentation_iteration(
    th_tree = pars$th_tree,
    th_seed = pars$th_seed,
    th_cr = pars$th_cr,
    max_cr = pars$max_cr,
    plots_to_run = plots_to_run,
    chm_list = chm_list,
    h = h,
    fieldPlots_extents = fieldPlots_extents,
    df_plot_obs = df_plot_obs,
    start_vals = start_vals,
    lower_bounds = lower_bounds,
    upper_bounds = upper_bounds,
    source_name = source_name
  )
  
  iter_end_time <- Sys.time()
  iter_time_sec <- as.numeric(difftime(iter_end_time, iter_start_time, units = "secs"))
  iteration_results[[i]]$iteration_summary$runtime_sec <- iter_time_sec
  
  setTxtProgressBar(pb, i)
  
  elapsed_total_sec <- as.numeric(difftime(iter_end_time, total_start_time, units = "secs"))
  avg_time_per_iter <- elapsed_total_sec / i
  remaining_iters <- n_iters - i
  est_remaining_sec <- avg_time_per_iter * remaining_iters
  
  message(
    "Finished iteration ", i, "/", n_iters,
    " in ", round(iter_time_sec, 2), " sec",
    " | elapsed: ", round(elapsed_total_sec / 60, 2), " min",
    " | est. remaining: ", round(est_remaining_sec / 60, 2), " min"
  )
}

close(pb)

total_end_time <- Sys.time()
total_time_min <- as.numeric(difftime(total_end_time, total_start_time, units = "mins"))

message("\nAll iterations completed in ", round(total_time_min, 2), " minutes.")

# ============================================================
# 4. COLLECT RESULTS
# ============================================================

optimization_results <- tibble(
  iteration = seq_along(iteration_results),
  summary = lapply(iteration_results, `[[`, "iteration_summary"),
  plot_metrics = lapply(iteration_results, `[[`, "plot_metrics"),
  modeled_trees = lapply(iteration_results, `[[`, "modeled_trees")
) %>%
  mutate(summary = purrr::map(summary, as.data.frame)) %>%
  tidyr::unnest(summary) %>%
  arrange(RMSE, desc(R2))

optimization_plot_metrics <- bind_rows(
  lapply(seq_along(iteration_results), function(i) {
    x <- iteration_results[[i]]$plot_metrics
    if (nrow(x) == 0) return(NULL)
    
    pars <- param_grid[i, ]
    summ <- iteration_results[[i]]$iteration_summary
    
    x %>%
      mutate(
        iteration = i,
        source = source_name,
        th_tree = pars$th_tree,
        th_seed = pars$th_seed,
        th_cr = pars$th_cr,
        max_cr = pars$max_cr,
        a = summ$a,
        b = summ$b,
        c = summ$c,
        runtime_sec = summ$runtime_sec
      )
  })
)

optimization_tree_table <- bind_rows(
  lapply(seq_along(iteration_results), function(i) {
    x <- iteration_results[[i]]$modeled_trees
    if (nrow(x) == 0) return(NULL)
    x %>% mutate(iteration = i)
  })
)

optimization_results
best_iteration <- optimization_results %>% slice(1)
best_iteration


```